---
title: "California Cooperative Snow Survey Summary"
author: "D.E. Beaudette"
date: "`r Sys.Date()`"
output:
  html_document:
  mathjax: null
jquery: null
smart: no
keep_md: no
---
  
```{r setup, echo=FALSE, results='hide', warning=FALSE}
# setup
library(knitr, quietly=TRUE)
library(kableExtra, quietly=TRUE)
opts_chunk$set(message=FALSE, warning=FALSE, background='#F7F7F7', fig.align='center', fig.retina=2, dev='png', tidy=FALSE, verbose=FALSE)
options(width=100, stringsAsFactors=FALSE, cache=TRUE)
```

These are my notes on getting, processing, and plotting the CA cooperative snow survey data each winter.

Key points:

   * data are fetched from CDEC via `sharpshootR::CDECsnowQuery()` by course number
   * data are compared using a combination of SWE percentiles, [SPI](http://iridl.ldeo.columbia.edu/maproom/Global/Precipitation/SPI.html), and z-score transformation of SWE
   * downloading the full record for all courses take a little while
   * some fairly bizzare Lattice graphics witchcraft is employed to get the first figure
   * the data object `CDEC.snow.courses` from the `sharpshootR` package contains a snapshot of snow course metadata
   

There is plenty of room for improvement.

## Custom Functions
```{r}
# http://joewheatley.net/20th-century-droughts/
getSPI <- function(y){
  #empirical, symmetrized Standard Precipitation Index
  fit.cdf <- ecdf(y)
  fit.rcdf <- ecdf(-y)
  cdfs <- fit.cdf(y)
  rcdfs <- 1 -fit.rcdf(-y)
  #invert normal
  spi.t <- qnorm(cdfs)
  spi.tp <- na.omit(spi.t[ spi.t != Inf ]) #drop Inf
  #reversed
  rspi.t <- qnorm(rcdfs)
  rspi.tp <- na.omit(rspi.t[ rspi.t != -Inf ]) #drop Inf
  #symmetrise
  spi.sym <- (spi.t+rspi.t)/2
  spi.sym[which(spi.t == Inf)] <- rspi.t[which(spi.t==Inf)]
  spi.sym[which(rspi.t == -Inf)] <- spi.t[which(rspi.t==-Inf)]
  return(spi.sym)
}

# custom panel function, run once within each lattice panel
panel.SWE <- function(x, y, ...) {
  panel.grid(h=FALSE, v=-1, lty=2)
  # make box-whisker plot
  panel.bwplot(x, y, ...)
  # tabulate number of obs (years) of data per station
  course.levels <- levels(y)
  n.yrs <- tapply(x, y, function(i) length(na.omit(i)))
  # determine station position on y-axis
  y.pos <- match(names(n.yrs), course.levels)
  # add 'yrs'
  n.yrs.txt <- paste('(', n.yrs, ' yrs)', sep='')
  # annotate panel with number of years / box-whisker
  grid.text(label=n.yrs.txt, x=unit(0.99, 'npc'), y=unit(y.pos, 'native'), gp=gpar(cex=0.65, font=2), just='right')
}

# custom stats for box-whisker plot: 5th-25th-50th-75th-95th percentiles
custom.bwplot <- function(x, coef=NA, do.out=FALSE) {
  stats <- quantile(x, p=c(0.05, 0.25, 0.5, 0.75, 0.95), na.rm=TRUE)
  n <- length(na.omit(x))
  out.low <- x[which(x < stats[1])]
  out.high <- x[which(x > stats[5])]
  return(list(stats=stats, n=n, conf=NA, out=c(out.low, out.high)))
}

# custom strip function, allowing more than 1 strip color
strip.color.fun <- function(..., bg) {
  strip.default(..., bg=trellis.par.get("strip.background")$col[which.packet()])
}
```


## Data and Summary for the Stanislaus
```{r results='hide'}
library(sharpshootR)
library(latticeExtra)
library(grid)
library(plyr)

# current data are:
yr <- 2019
mo <- 'February'

# Stanislaus River watershed
course.data <- data.frame(course_number=c(129, 323, 131, 132, 134, 345, 344, 138, 139, 384, 140, 142, 143, 145, 152))
```

```{r results='hide'}
# get historic data
d.long_term <- ddply(course.data, 'course_number', .progress='text', .fun=function(i) {CDECsnowQuery(course=i$course_number, start_yr=1900, end_yr=2019)})

# join-in metadata
data(CDEC.snow.courses)
d.long_term <- join(d.long_term, CDEC.snow.courses, by='course_number', type='left')

# reset levels of site names, sorting by elevation
names.elevations <- unique(d.long_term[, c('name', 'elev_feet')])
new.levels <- names.elevations$name[order(names.elevations$elev_feet)]
d.long_term$name <- factor(d.long_term$name, levels=new.levels)

# c('February', 'March', 'April', 'May')


## compute long-term average, by course / month, using the Adjusted 
d.long_term.avg <- ddply(d.long_term, c('course_number', 'month'), summarise, 
                         avg.SWE=mean(SWE, na.rm=TRUE), 
                         avg.density=mean(density, na.rm=TRUE),
                         avg.Depth=mean(Depth, na.rm=TRUE),
                         no.yrs=length(na.omit(SWE))
)

# make a copy of the current yr / mo data
d.current <- subset(d.long_term, subset=month == mo & year == yr)

# join current data with long term averages
# keep only data that exists in both tables
d.merged <- join(d.current, d.long_term.avg, by=c('course_number', 'month'), type='left')

# compute pct of normal of depth, SWE, density
d.merged$pct_of_normal_Depth <- with(d.merged, (Depth / avg.Depth) * 100.0)
d.merged$pct_of_normal_SWE <- with(d.merged, (SWE / avg.SWE) * 100.0)
d.merged$pct_of_normal_density <- with(d.merged, (density / avg.density) * 100.0)


## using customized boxplot !!
## compare with Feb, March, April, May only

# determine which panel contains the current month's data
this.month.panel <- match(mo, levels(d.current$month))
# strip colors
strip.colors <- rep(grey(0.95), times=length(levels(d.current$month)))
# modifiy this month's strip color
strip.colors[this.month.panel] <- grey(0.75)

## compare this year's data with long-term variability
p1 <- xyplot(name ~ SWE | month, data=d.current, fill='RoyalBlue', pch=21, cex=0.75, ylab='')
p2 <- bwplot(name ~ SWE | month, data=d.long_term, panel=panel.SWE, stats=custom.bwplot, xlab='Snow Water Eqivalent (in)', layout=c(5, 1), strip=strip.color.fun, as.table=TRUE, scales=list(x=list(alternating=3)), subset=month %in% c('January','February','March','April','May'))


# combine figures
p3 <- p2 + p1

# add title, and legend
main.title <- paste(mo, yr, 'Snow Survey')
p3 <- update(p3, main=main.title, key=list(corner=c(0.97,-0.15), points=list(pch=21, fill='RoyalBlue', cex=1), text=list(paste('Current (', mo, ' ', yr, ') SWE', sep='')), between=1))


# make legend sub-plot
set.seed(101010)
x <- rnorm(100)
x.stats <- custom.bwplot(x)
p4 <- bwplot(x, stats=custom.bwplot, scales=list(draw=FALSE), xlab=list('Percentiles', cex=0.75), ylab='', ylim=c(0.5, 1.75))
p4 <- p4 + layer(panel.text(x=x.stats$stats, y=1.5, label=c('5%','25%','50%','75%','95%'), cex=0.75))
```


```{r fig.width=12, fig.height=6}
# setup plot style
tps <- list(box.dot=list(col='black', pch='|'), box.rectangle=list(col='black'), box.umbrella=list(col='black', lty=1), plot.symbol=list(col='black', cex=0.33), strip.background=list(col=strip.colors))

trellis.par.set(tps)
print(p3, more=TRUE, position=c(0,0.08,1,1))
print(p4, more=FALSE, position=c(0.125,0,0.425,0.175))
```



## Analysis of the Entire Record, all Courses
```{r results='hide'}
# all snow courses
data(CDEC.snow.courses)

# get historic data
d.long_term <- ddply(CDEC.snow.courses, 'course_number', .progress='text', .fun=function(i) {CDECsnowQuery(course=i$course_number, start_yr=1900, end_yr=2019)})

# save local cache
# save(d.long_term, file='long_term_data-cache.rda')

# join-in the snow metadata
d.long_term <- join(d.long_term, CDEC.snow.courses, by='course_number', type='left')

# reset levels of site names, sorting by elevation
names.elevations <- unique(d.long_term[, c('name', 'elev_feet')])
new.levels <- names.elevations$name[order(names.elevations$elev_feet)]
d.long_term$name <- factor(d.long_term$name, levels=new.levels)

## scale by course / month, across all years
scaled.data <- ddply(d.long_term, c('name', 'month'), function(i) {
  scaled.SWE <- scale(i$SWE)
  emp.pctile <- ecdf(i$SWE)(i$SWE)
  spi <- getSPI(i$SWE)
  d <- data.frame(course_number=i$name, Date=i$Date, year=i$year, month=i$month, scaled.SWE=scaled.SWE, emp.pctile=emp.pctile, SWE=i$SWE, spi=spi)
  return(d)
})

# re-order by course, Date
scaled.data <- scaled.data[order(scaled.data$name, scaled.data$Date), ]

# subset out the typical months
scaled.data <- subset(scaled.data, month %in% c('February', 'March', 'April', 'May') & year > 1930)
scaled.data$month <- factor(scaled.data$month)
```


```{r fig.width=10, fig.height=7}
# plot all data, each line is a course
xyplot(SWE ~ year | month, data=scaled.data, groups=name, col=rgb(0.1, 0.1, 0.1, alpha=0.25), type=c('l','g'), span=0.125, as.table=TRUE, scales=list(alternating=3, x=list(tick.number=10, rot=90)))

# plot all data: each dot is a survey
xyplot(SWE ~ year | month, data=scaled.data, type=c('p','g', 'smooth'), span=0.125, as.table=TRUE, scales=list(alternating=3, x=list(tick.number=10, rot=90)), cex=0.25, pch=16, lwd=2)

xyplot(scaled.SWE ~ year | month, data=scaled.data, type=c('p','g', 'smooth'), span=0.125, as.table=TRUE, scales=list(alternating=3, x=list(tick.number=10, rot=90)), cex=0.25, pch=16, lwd=2)

xyplot(emp.pctile ~ year | month, data=scaled.data, type=c('p','g', 'smooth'), span=0.125, as.table=TRUE, scales=list(alternating=3, x=list(tick.number=10, rot=90)), cex=0.25, pch=16, lwd=2)

xyplot(emp.pctile ~ year | month, data=scaled.data, type=c('p','g', 'smooth'), span=0.25, as.table=TRUE, scales=list(alternating=3, x=list(tick.number=10, rot=90)), cex=0.25, pch=16, lwd=2, subset=year > 1970)
```


```{r fig.width=10, fig.height=6}
tps <- list(box.dot=list(col='black', pch='|'), box.rectangle=list(col='black'), box.umbrella=list(col='black', lty=1), plot.symbol=list(col='black', cex=0.33), strip.background=list(col=grey(0.85)))

bwplot(emp.pctile ~ factor(year) | month, data=scaled.data, as.table=TRUE, par.settings=tps, scales=list(alternating=3, x=list(tick.number=10, rot=90)), subset=year > 1970 & month == 'February', stats=custom.bwplot, panel=function(...) {panel.grid(-1, -1); panel.bwplot(...)})

bwplot(spi ~ factor(year) | month, data=scaled.data, as.table=TRUE, par.settings=tps, scales=list(alternating=3, x=list(tick.number=10, rot=90)), subset=year > 1970 & month == 'February', stats=custom.bwplot, panel=function(...) {panel.grid(-1, -1); panel.bwplot(...)})

bwplot(emp.pctile ~ factor(year), data=scaled.data, main='February Snow Survey', sub='Period of Record: 1939-2019', ylim=c(-0.05, 1.1), ylab='Percentile', par.settings=tps, scales=list(alternating=3, y=list(tick.number=10)), subset=year > 1997 & month == 'February', panel=function(x=x, y=y, ...) {
  panel.grid(-1, -1)
  panel.abline(h=0.5, lty=2, col='orange')
  panel.bwplot(x=x, y=y, ...)
  n.stations <- tapply(y, x, length)
  panel.text(x=1:length(n.stations), y=1.03, labels=n.stations, cex=0.65, font=2)
  panel.text(x=10, y=1.075, labels='Snow Courses Reporting', cex=0.85, font=3)
  })
```



----------------------------
This document is based on `sharpshootR` version `r utils::packageDescription("sharpshootR", field="Version")`.


