---
title: "CDEC: CA Climate Data"
author: "D.E. Beaudette"
date: "`r Sys.Date()`"
output:
  html_document:
  mathjax: null
jquery: null
smart: no
keep_md: no
---
  
```{r setup, echo=FALSE, results='hide', warning=FALSE}
# setup
library(knitr, quietly=TRUE)
library(kableExtra, quietly=TRUE)
opts_chunk$set(message=FALSE, warning=FALSE, background='#F7F7F7', fig.align='center', fig.retina=2, dev='png', tidy=FALSE, verbose=FALSE)
options(width=100, stringsAsFactors=FALSE, cache=TRUE)
```


# Background

The [California Data Exchange Center (CDEC)](https://cdec.water.ca.gov/) hosts all kinds of useful climate data. You can get these easily by station and sensor code with the `CDECquery()` function. Lookup sensor codes by station with the `CDEC_StationInfo()` function.

Find CDEC stations via:

   * [map](http://cdec.water.ca.gov/cdecstation2/)
   * [search](http://cdec.water.ca.gov/cgi-progs/staSearch)
   * [station ID](http://cdec.water.ca.gov/dynamicapp/staMeta)
   * [other](http://cdec.water.ca.gov/staInfo.html)


List of sensor codes [by ID](http://cdec.water.ca.gov/misc/senslist.html) and associated units of measure.



# Examples


## Setup R Environment
With a recent version of R (>= 2.15), it is possible to get all of the packages that this tutorial depends on via:
```{r install-deps, eval=FALSE}
# run these commands in the R console
install.packages('sharpshootR', dep=TRUE)
install.packages('soilDB', dep=TRUE)
install.packages('plyr', dep=TRUE)
install.packages('seas', dep=TRUE)
install.packages('viridis', dep=TRUE)
install.packages('latticeExtra', dep=TRUE)
install.packages('RColorBrewer', dep=TRUE)
install.packages('zoo', dep=TRUE)

# get latest version from GitHub
install.packages('remotes', dep=TRUE)
remotes::install_github("ncss-tech/sharpshootR", dependencies=FALSE, upgrade_dependencies=FALSE)
```


Load required packages.
```{r}
library(latticeExtra)
library(plyr)
library(sharpshootR)
library(viridis)
library(seas)
library(RColorBrewer)
library(zoo)
```

## Basic Usage

Get metadata for the Stanislaus Powerhouse, [SPW](http://cdec.water.ca.gov/dynamicapp/staMeta?station_id=SPW), station. The results are stored in a list containing site and sensor metadata, along with any commentary related to the history of the station.
```{r }
meta <- CDEC_StationInfo('SPW')
str(meta, 1)
```

Site metadata.
```{r}
kable_styling(kable(meta$site.meta, format = 'html'), font_size = 10)
```

Sensor metadata.
```{r}
kable_styling(kable(meta$sensor.meta, format = 'html'), font_size = 10)
```


Get daily precipitation (sensor = **45**, interval = **'D'**) for a specified time interval.
```{r}
res <- CDECquery(id='SPW', sensor=45, interval='D', start='2018-01-01', end='2018-01-31')

# first 10 rows
kable_styling(kable(head(res, 10), format = 'html'), font_size = 10)
```


## Soil Temperature and Moisture
Some of the CDEC stations have below-ground sensors. The [Highland Meadow](http://cdec.water.ca.gov/dynamicapp/staMeta?station_id=HHM) station was the first one to have soil sensors installed at three depths. These data are collected every 20 minutes.

Get the 3 soil temperature and volumetric water content (VWC) sensor IDs for HHM. Queries are by sensor ID.
```{r results='hide'}
# CDEC station ID
stn <- 'HHM'

# note that station metadata is a list
stn.info <- CDEC_StationInfo(stn)
sensors <- stn.info$sensor.meta

# lookup sensor codes for soil temp and moisture
soil.temp.sensors <- sensors$sensor[grep('soil temp', sensors$sensor_details, ignore.case = TRUE)]
soil.moist.sensors <- sensors$sensor[grep('soil moistr', sensors$sensor_details, ignore.case = TRUE)]
```


Query the data. It might take a couple of seconds. Note that we are making two distinct requests, each for three sensors.
```{r}
d.soil_temp <- CDECquery(id = stn, sensor = soil.temp.sensors, interval = 'E', start = '2017-09-15', end = '2019-09-15')
d.soil_moist <- CDECquery(id = stn, sensor = soil.moist.sensors, interval = 'E', start = '2017-09-15', end = '2019-09-15')

# quick check: nice!
kable_styling(kable(head(d.soil_temp, 10), format = 'html'), font_size = 10)

# this is a lot of data
nrow(d.soil_temp)
```


Perform some basic filtering, note units.
```{r results='hide'}
# remove soil temperature less than 20 deg F
d.soil_temp <- subset(d.soil_temp, subset=value > 20)

# remove soil moisture (VWC) < 1%
d.soil_moist <- subset(d.soil_moist, subset=value > 1)
```


Plot soil moisture and temperature in a single figure. There are many ways to do this, here we are using concepts from the `lattice` package. Note that we stack the two data sets into "long format" for plotting.
```{r results='hide', fig.width=12, fig.height=8}
# combine into long format for plotting
d <- make.groups('Soil Temperature (F)'=d.soil_temp, 'Soil Volumetric Water Content (%)'=d.soil_moist)

# define some reasonable plot colors and line widths
tps <- list(superpose.line=list(col=brewer.pal(3, 'Set1'), lwd=1.5))

# plot
xyplot(value ~ datetime | which, groups=sensor_type, data=d, type=c('g', 'l'), auto.key=list(cex=0.5, columns=2, points=FALSE, lines=TRUE), strip=strip.custom(bg=grey(0.90)), par.settings=tps, layout=c(1,2), scales=list(alternating=3, x=list(format="%b %d\n%Y", tick.number=10), y=list(relation='free', rot=0)), xlab='', ylab='', sub=stn)
```


Sometimes the daily median is sufficient.
```{r results='hide', fig.width=12, fig.height=8}
# truncate datetime to date
d$date <- as.Date(d$datetime)

# aggregating over date / sensor group / specific sensor
a <- ddply(d, c('date', 'which', 'sensor_type'), .progress = 'text', .fun = plyr::summarize, q50=median(value, na.rm=TRUE))

# plot daily median
xyplot(q50 ~ date | which, groups=sensor_type, data=a, type=c('g', 'l'), auto.key=list(cex=0.5, columns=2, points=FALSE, lines=TRUE), strip=strip.custom(bg=grey(0.90)), par.settings=tps, layout=c(1,2), scales=list(alternating=3, x=list(format="%b\n%Y", tick.number=10), y=list(relation='free', rot=0)), xlab='', ylab='', sub=stn)
```

## Precipitation

### Daily Precipitation

Query the daily precipitation measurements spanning the 2018-2019 water year from Sonora Ranger Station ([SOR](http://cdec.water.ca.gov/cgi-progs/staMeta?station_id=SOR)).

```{r, fig.height=5, fig.width=7}
# get data, note sensor number
x <- CDECquery(id='sor', sensor=45, interval='D', start='2018-09-01', end='2019-09-01')

# compute cumulative PPT
# replace NA with 0 in cumulative calc
x$cumulative <- cumsum(ifelse(is.na(x$value), 0, x$value))

# plot
xyplot(value ~ datetime, data=x, type=c('h', 'g'), scales=list(x=list(alternating=3, format="%b %d\n%Y", tick.number=10)), xlab='', ylab='Daily PPT (inches)', main='Sonora Ranger Station\nDaily Total Precipitation (inches)')

xyplot(cumulative ~ datetime, data=x, type=c('s', 'g'), scales=list(x=list(alternating=3, format="%b %d\n%Y", tick.number=10)), xlab='', ylab='Cumulative PPT (inches)', main='Sonora Ranger Station\nCumulative Precipitation (inches)')

# total
sum(x$value, na.rm = TRUE)
```

### Monthly Precipitation
Query the entire record of monthly precipitation measurements from Sonora Ranger Station ([SOR](http://cdec.water.ca.gov/cgi-progs/staMeta?station_id=SOR)).
```{r, fig.height=5, fig.width=7}
x <- CDECquery(id='SOR', sensor=2, interval='M', start='1900-01-01', end='2030-01-01')

# plot
bwplot(month ~ value, data=x, ylab='', main='Sonora Ranger Station\nVariation in Monthly Total PPT', xlab='Monthly Total PPT (inches)')
```


### Water Year Summaries

Cumulative PPT over the course of an idealized water year (October 1 through September 30). Measurements are selected from water years with annual PPT close to the 5th, 25th, 50th, 75th, and 95th percentiles of annual PPT (over all water years). The current water year is symbolized with a red line. The red box and whisker plot (modified, whiskers are 5th and 95th percentiles) summarizes cumulative PPT on the current "water day".
```{r fig.height=5.5, fig.width=10}
# define station of interest
s <- 'SPW'
# get metadata
s.info <- CDEC_StationInfo(s)
# format title for cumulative PPT
title.text <- sprintf("%s [%s]", s.info$site.meta$Name, s)

# get data
x <- CDECquery(id=s, sensor=45, interval='D', start='1999-01-01', end='2030-01-01')

## NOTE: requires sharpshootR >= 1.4.02
# plot
par(mar=c(4.5, 4.5, 2.5, 1.5))
PCP_plot(x, ylab='Cumulative PPT (inches)', main=title.text, this.year = 2019)
```


### Visualization Ideas
Query the entire record of daily precipitation measurements from New Hogan Lake ([NHG](http://cdec.water.ca.gov/cgi-progs/staMeta?station_id=NHG)).
```{r, fig.width=7.5, fig.height=5}
x <- CDECquery(id='NHG', sensor=45, interval='D', start='1900-01-01', end='2020-01-01')
summary(x$value)
head(x)
nrow(x)
```

Plot raw data.
```{r, fig.width=8, fig.height=4}
xyplot(value ~ datetime, data=x, type='h', main='NEW HOGAN LAKE (NHG)', ylab='Daily Precipitation (inches)', xlab='')
```

Sum daily precipitation by year/month and display as box-whisker plot.
```{r, fig.width=8, fig.height=4.5}
a <- ddply(x, c('year','month'), plyr::summarise, ppt=sum(value, na.rm=TRUE))

# plot
bwplot(ppt ~ month, data=a, xlab='', ylab='Monthly Total Precipitation (inches)', main='NEW HOGAN LAKE (NHG)', scales=list(y=list(tick.number=10), x=list(cex=0.85, rot=45)))
```

Visualize monthly precipitation as grouped by year/month.
```{r, fig.width=12, fig.height=5}
# color ramp for monthly PPT
# denote 0 with grey 
cols <- colorRampPalette(rev(c(viridis(10), grey(0.90))), space='Lab', interpolate='spline')

levelplot(ppt ~ factor(year) * month, data=a, col.regions=cols, xlab='', 
          ylab='', scales=list(x=list(tick.number=10, rot=45)), 
          main='NEW HOGAN LAKE (NHG)\nMonthly Total Precipitation (inches)')
```


Alternate representation via `seas` package.
```{r, fig.width=11, fig.height=5}
# alter column names and convert units from inches to mm
x$date <- as.Date(x$datetime)
x$value <- x$value * 25.4
x$precip <- x$value

# init seas object
ss <- seas.sum(x, 'precip', width='month', start.day = as.Date('2015-09-01'), prime='precip')

# monthly summaries, arranged according to CA water year
plot(ss, main='NEW HOGAN LAKE (NHG)')
```

Alternate representation
```{r, fig.width=10, fig.height=10}
image(ss)
```

Monthly normals.
```{r, fig.width=11, fig.height=5}
sn <- seas.norm(ss)
plot(sn)
```




## Reservoir Levels
```{r, fig.width=11, fig.height=5}
# get daily resevoir storage (ac. ft) from Pinecrest, New Melones, Lyons, and Don Pedro Reservoirs
pinecrest <- CDECquery(id='swb', sensor=15, interval='D', start='2015-01-01', end='2020-01-01')
new.melones <- CDECquery(id='nml', sensor=15, interval='D', start='2015-01-01', end='2020-01-01')
lyons <- CDECquery(id='lys', sensor=15, interval='D', start='2015-01-01', end='2020-01-01')
dp <- CDECquery(id='dnp', sensor=15, interval='D', start='2015-01-01', end='2020-01-01')

# compute storage capacity, based on published capacities
pinecrest$capacity <- pinecrest$value / 18312 * 100
new.melones$capacity <- new.melones$value / 2400000 * 100
lyons$capacity <- lyons$value / 6228 * 100
dp$capacity <- dp$value / 	2030000 * 100

# combine
g <- make.groups('New Melones'=new.melones, 'Lyons'=lyons, 'Pinecrest'=pinecrest, 'Don Pedro'=dp)

# resonable date scale
r <- range(g$datetime)
s.r <- seq(from=r[1], to=r[2] + 6000000, by='6 months')

# better colors
tps <- list(superpose.line=list(lwd=2, col=brewer.pal(n=4, name='Set1')))

# plot
xyplot(capacity ~ datetime, groups=which, data=g, type='l', 
       xlab='', ylab='Capacity (%)',
       scales=list(alternating=3, y=list(tick.number=10),x=list(at=s.r, labels=format(s.r, "%B\n%Y"))),
       auto.key=list(columns=4, lines=TRUE, points=FALSE),
       par.settings=tps,
       panel=function(...) {
         panel.abline(h=seq(0, 100, by=10), col='grey')
         panel.abline(v=s.r, col='grey')
         panel.xyplot(...)
       })
```


Investigate monthly levels and capacity of New Melones Reservoir.
```{r, fig.width=11, fig.height=5}
# New Melones monthly data, retrieve as far back in time as possible 
new.melones.monthly <- CDECquery(id='nml', sensor=15, interval='M', start='1900-01-01', end='2020-01-01')

# convert to pct. capacity
new.melones.monthly$capacity <- new.melones.monthly$value / 2400000 * 100

# color ramp
cols <- colorRampPalette(rev(viridis(10)), space='Lab', interpolate='spline')

# plot, each pixel is colored by the total precip by year/month
levelplot(capacity ~ year * month, data=new.melones.monthly, col.regions=cols, xlab='', ylab='', scales=list(x=list(tick.number=20)), main='New Melones Capacity (%)')
```






----------------------------
This document is based on `sharpshootR` version `r utils::packageDescription("sharpshootR", field="Version")`.


