---
title: "A Novel Display of Categorical Pedon Data"
author: "D.E. Beaudette"
date: "`r Sys.Date()`"
output:
  html_document:
    mathjax: null
    jquery: null
    smart: no
---


```{r setup, echo=FALSE, results='hide', warning=FALSE}
library(knitr, quietly=TRUE)

opts_chunk$set(message=FALSE, warning=FALSE, background='#F7F7F7', fig.align='center', fig.retina=2, dev='png', tidy=FALSE, verbose=FALSE)
options(width=100, stringsAsFactors=FALSE, cache=TRUE)
```


# Introduction
This tutorial describes a simple process for visualizing patterns in categorical data associated with pedon data.

## Adapting Content to Your Data
Copy and paste blocks of code in this tutorial into a new R Studio script file (*ctrl + shift + n* makes a new file), edit, and then run. Running lines or blocks of code in an RStudio script file is as simple as moving the cursor to the line (or selecting a block) of code and press *ctrl + enter*.

# Setup R Environment
This step is only required the **first time** you open R. These packages will be available via `library()` in later sessions.

With a recent version of R (>= 2.15), it is possible to get all of the packages that this tutorial depends on via:
```{r install-deps, eval=FALSE}
# run these commands in the R console, only once
install.packages('aqp', dep=TRUE) # stable version from CRAN + dependencies
install.packages('soilDB', dep=TRUE) # stable version from CRAN + dependencies
install.packages('sharpshootR', dep=TRUE) # stable version from CRAN + dependencies
```


# Examples
The following examples are based on the `loafercreek` example dataset from the `soilDB` package. This is a good example of the type of `SoilProfileCollection` object returned by `fetchNASIS(from='pedons')`. The 

```{r, fig.width=8, fig.height=7}
library(aqp)
library(sharpshootR)

# load some example NASIS data
data(loafercreek, package='soilDB')

# cut-down to a subset
loafercreek <- loafercreek[1:20, ]

# get depth class
sdc <- getSoilDepthClass(loafercreek)

# check it out
head(sdc)

# join back into site-level attributes
site(loafercreek) <- sdc

# diagnostic properties to consider, no need to convert to factors
# these are stored in the site-level attributes of the SPC
v <- c('lithic.contact', 'paralithic.contact', 'argillic.horizon', 
       'cambic.horizon', 'ochric.epipedon', 'mollic.epipedon', 'very.shallow',
       'shallow', 'mod.deep', 'deep', 'very.deep')


# presence / absense matrix of several diagnostic features
x <- diagnosticPropertyPlot(loafercreek, v, k=5, grid.label='bedrock_kind', dend.label = 'taxonname')
x <- diagnosticPropertyPlot(loafercreek, v, k=5, grid.label='pedon_id', dend.label = 'taxonname')

x <- diagnosticPropertyPlot2(loafercreek, v, k=5, grid.label='pedon_id')
x <- diagnosticPropertyPlot2(loafercreek, v, k=5, grid.label='taxonname')
```

It is possible to include other categorical variables such as hillslope position. The `multinominal2logical()` is used to expand a set of categories into a matrix of TRUE/FALSE.

```{r, fig.width=8, fig.height=7}
# work-around: new function in sharpshootR
hp <- multinominal2logical(loafercreek, 'hillslope_pos')

# check it out
head(hp)

# join back into site-level attributes
site(loafercreek) <- hp

# init variable names
v <- c('lithic.contact', 'paralithic.contact', 'argillic.horizon', 
       'ochric.epipedon', 'mollic.epipedon', 'mod.deep', 'deep', 'very.deep', levels(loafercreek$hillslope_pos))

# should work as before
# use the 'sort.vars = FALSE' option to set the order of the variable names to the order they are listed in 'v'
x <- diagnosticPropertyPlot(loafercreek, v, k=5, grid.label='bedrock_kind', dend.label = 'taxonname', sort.vars = FALSE)
x <- diagnosticPropertyPlot2(loafercreek, v, k=5, grid.label='bedrock_kind', sort.vars = FALSE)
```


----------------------------
This document is based on `sharpshootR` version `r utils::packageDescription("sharpshootR", field="Version")`, `aqp` version `r utils::packageDescription("aqp", field="Version")`, and `soilDB` version `r utils::packageDescription("soilDB", field="Version")`.

