---
title: "Frost-Free Day Estimation"
author: "D.E. Beaudette"
date: "`r Sys.Date()`"
output:
  html_document:
  mathjax: null
jquery: null
smart: no
keep_md: no
---
  
```{r setup, echo=FALSE, results='hide', warning=FALSE}
# setup
library(knitr, quietly=TRUE)
opts_chunk$set(message=FALSE, warning=FALSE, background='#F7F7F7', fig.align='center', fig.retina=2, dev='png', tidy=FALSE, verbose=FALSE)
options(width=100, stringsAsFactors=FALSE, cache=TRUE)
```


# Background

... finish this

NASIS component table (ffd_r):
  <i>The expected number of days between the last freezing temperature (0 degrees Celsius) in spring (Jan-Jul) and the first freezing temperature (0 degrees Celsius) in the fall (Aug-Dec). The number of days is based on the probability that the values for the standard "normal" period of 1961 to 1990 will be exceeded in 5 years out of 10.</i>

Links:
 
 * https://www.ncdc.noaa.gov/news/when-expect-your-last-spring-freeze
 * ...


## Graphical Demonstration

```{r eval=FALSE, echo=FALSE}
ffp <- ddply(x, "year", sharpshootR:::.frostFreePeriod, minDays = 165, frostTemp=32)

plot(0,0, xlim=c(0,366), ylim=c(0, 0.1), type='n')
lines(density(ffp$lastSpring))
lines(density(ffp$firstFall))
```



# Examples


## Setup R Environment
With a recent version of R (>= 2.15), it is possible to get all of the packages that this tutorial depends on via:
```{r install-deps, eval=FALSE}
# run these commands in the R console
install.packages('sharpshootR', dep=TRUE)
install.packages('soilDB', dep=TRUE)
install.packages('plyr', dep=TRUE)

# get latest version from GitHub
install.packages('devtools', dep=TRUE)
devtools::install_github("ncss-tech/soilDB", dependencies=FALSE, upgrade_dependencies=FALSE)
devtools::install_github("ncss-tech/sharpshootR", dependencies=FALSE, upgrade_dependencies=FALSE)
```


Load required packages.
```{r}
library(sharpshootR)
library(soilDB)
library(plyr)
```


## CDEC
The [California Data Exchange Center (CDEC)](https://cdec.water.ca.gov/) hosts all kinds of useful climate data. You can get these easily by station and sensor code with the `CDECquery()` function.
```{r, fig.width=10, fig.height=5}
# HHM: Highland Meadows
# 32: daily min air temperature
x <- CDECquery(id='HHM', sensor=32, interval='D', start='1900-01-01', end='2016-12-31')

# estimate FFD over all years
# air temperature are in deg F
x.ffd <- FFD(x, returnDailyPr = TRUE, frostTemp=32)

# basic summary
kable(x.ffd$summary)

# graphical summary of FFD estimates
par(mar=c(4,3.5,3,1))
FFDplot(x.ffd, 'CDEC - Highland Meadows')
```


## SNOTEL
Data from [SNOTEL Station 365](https://wcc.sc.egov.usda.gov/nwcc/site?sitenum=365).
```{r, fig.width=10, fig.height=5}
# site code and year range are required
x <- fetchSCAN(site.code=365, year=c(1995:2016))

# extract the daily min air temperature records
x <- x$TMIN

# re-name and add "year" column
names(x)[2] <- 'datetime'
x$year <- as.integer(format(x$datetime, "%Y"))

# estimate FFD over all years
# air temperature are in deg C
x.ffd <- FFD(x, returnDailyPr = TRUE, frostTemp=0)

# basic summary
kable(x.ffd$summary)

# graphical summary of FFD estimates
par(mar=c(4,3.5,3,1))
FFDplot(x.ffd, 'SNOTEL Station 365')
```


## SCAN
Data from [SCAN Station 808](https://wcc.sc.egov.usda.gov/nwcc/site?sitenum=808).
```{r, fig.width=10, fig.height=5}
# same interface as SNOTEL
# site code and year range are required
x <- fetchSCAN(site.code=808, year=c(1995:2016))

# extract the daily min air temperature records
x <- x$TMIN

# re-name and add "year" column
names(x)[2] <- 'datetime'
x$year <- as.integer(format(x$datetime, "%Y"))

# estimate FFD over all years
# air temperature are in deg C
x.ffd <- FFD(x, returnDailyPr = TRUE, frostTemp=0)

# basic summary
kable(x.ffd$summary)

# graphical summary of FFD estimates
par(mar=c(4,3.5,3,1))
FFDplot(x.ffd, 'SCAN Station 808')
```


## Henry Mount Soil Climate DB
Air temperature data associated with NASIS user site ID "1997NV005022".
```{r, fig.width=10, fig.height=5}
# must specify gran='hour' for hourly data
# otherwise daily mean values are returned
x <- fetchHenry(usersiteid = '1997NV005022', what = 'airtemp', gran='hour')

# extract air temperature records
x <- x$airtemp
# convert date-time stamp to date (truncate time component)
x$datetime <- as.Date(x$date_time)

# compute daily min air temperature
x <- ddply(x, 'datetime', summarize, value=min(sensor_value, na.rm=TRUE))
# add "year" column
x$year <- as.integer(format(x$datetime, "%Y"))

# estimate FFD over all years
# air temperature are in deg C
x.ffd <- FFD(x, returnDailyPr = TRUE, frostTemp=0)

# basic summary
kable(x.ffd$summary)

# graphical summary of FFD estimates
par(mar=c(4,3.5,3,1))
FFDplot(x.ffd, 'Henry: 1997NV005022')
```



----------------------------
This document is based on `sharpshootR` version `r utils::packageDescription("sharpshootR", field="Version")` and `soilDB` version `r utils::packageDescription("soilDB", field="Version")`.


