---
title: "A Simple Monthly Water Balance Simulation"
author: "D.E. Beaudette and J.M. Skovlin"
date: "`r Sys.Date()`"
output:
  html_document:
    mathjax: null
    jquery: null
    smart: no
    keep_md: no
---
  
```{r setup, echo=FALSE, results='hide', warning=FALSE}
# setup
library(knitr, quietly=TRUE)
opts_chunk$set(message=FALSE, warning=FALSE, background='#F7F7F7', fig.align='center', dev='svglite', tidy=FALSE, verbose=FALSE)
options(width=100, stringsAsFactors=FALSE, cache=TRUE)
```

## Introduction



## Setup
You will need these packages. Must use the development version of `hydromad`, requires a working compiler (RTools on Windows).
```{r}
library(aqp)
library(soilDB)
library(sharpshootR)

# special installation instructions:
# https://github.com/josephguillaume/hydromad#installation
library(hydromad)
```

## Simple Example
Simulate monthly water balances for 50cm of sandy loam vs. 50cm of silt loam soil material in a xeric/thermic climate.
```{r fig.width=16, fig.height=6.5}
data("ROSETTA.centroids")

idx <- which(ROSETTA.centroids$texture %in% c('sandy loam', 'silt loam'))
vars <- c('texture', 'pwp', 'fc', 'sat', 'awc')

knitr::kable(
  ROSETTA.centroids[idx, vars],
  row.names = FALSE
)

# near Sonora, CA
# thermic / xeric climate
# units are mm
PET <- c(0, 0, 5,80, 90, 120, 130, 140, 110, 90, 20, 5)
PPT <- c(0, 150, 200, 120, 20, 0, 0, 0, 10, 20, 30, 60)

# 50cm of soil material -> convert to mm
thick <- 50 * 10

# total water storage for sandy loam texture soil material
AWC.sl <- round(0.1049114 * thick)

# total water storage for silt loam texture soil material
AWC.sil <- round(0.2080796 * thick)

# simulate water balances:
# * run for 5 cycles
# * start in January
# * start with "completely dry" soil conditions
# * keep only the last cycle
wb.sl <- monthlyWB(AWC.sl, PPT, PET, S_init = 0, starting_month = 1, rep = 5, keep_last = TRUE)
wb.sil <- monthlyWB(AWC.sil, PPT, PET, S_init = 0, starting_month = 1, rep = 5, keep_last = TRUE)
```

All values are mm. U: surplus, S: soil water storage, ET: actual ET, D: deficit.
```{r}
kable(wb.sil, digits = 2, row.names = FALSE)
```


```{r fig.width=16, fig.height=6.5}
# compare
par(mar=c(4,4,4,1), bg = 'white', mfcol = c(1, 2))
plotWB(WB = wb.sl, legend.cex = 0.75)
title('50cm of sandy loam texture soil material', line = 3)

plotWB(WB = wb.sil, legend.cex = 0.75)
title('50cm of silt loam texture soil material', line = 3)
```

```{r fig.width=16, fig.height=6}
# compare
par(mar=c(4,4,4,1), bg = 'white', mfcol = c(1, 2))
plotWB_lines(WB = wb.sl)
title('50cm of sandy loam texture soil material', line = 3)

plotWB_lines(WB = wb.sil)
title('50cm of silt loam texture soil material', line = 3)
```



## Real Data

Series-level summaries of climate provided by [`fetchOSD()`](http://ncss-tech.github.io/AQP/soilDB/soil-series-query-functions.html).

AWC estimated from SSURGO data, via SDA.

```{r}
s <- 'amador'
x <- fetchOSD(s, extended = TRUE)

# get representative, profile-total AWC from SSURGO
sql <- sprintf("
SELECT chorizon.cokey AS cokey, 
SUM(awc_r * (hzdepb_r - hzdept_r)) AS ws 
FROM 
legend JOIN mapunit ON legend.lkey = mapunit.lkey
JOIN component ON mapunit.mukey = component.mukey
JOIN chorizon ON component.cokey = chorizon.cokey 
WHERE compname = '%s'
AND areasymbol != 'US'
GROUP BY chorizon.cokey;", s
)

# get via SDA
res <- SDA_query(sql)

# median AWC in mm
# over all components correlated to named series 
AWC <- round(median(res$ws, na.rm = TRUE) * 10)


# monthly climate data from series summary
PPT <- x$climate.monthly$q50[x$climate.monthly$variable == 'Precipitation (mm)']
PET <- x$climate.monthly$q50[x$climate.monthly$variable == 'Potential ET (mm)']
```


Water year.
```{r fig.width=8, fig.height=6}
# tighter margins
par(mar=c(4,4,2,1), bg = 'white')

# water year
# last iteration
x.wb <- monthlyWB(AWC, PPT, PET, S_init = 0, starting_month = 9, rep = 3, keep_last = TRUE)
plotWB(WB = x.wb)

# convert total ETa into inches
sum(x.wb$ET) * 0.03937
```

Calendar year.
```{r fig.width=8, fig.height=5.5}
# tighter margins
par(mar=c(4,4,2,1), bg = 'white')

# calendar year
# last iteration
x.wb <- monthlyWB(AWC, PPT, PET, S_init = 0, starting_month = 1, rep = 3, keep_last = TRUE)
plotWB(WB = x.wb)
```



### Bar vs. Line Graph
```{r fig.width=16, fig.height=6.5}
s <- 'auburn'
x <- fetchOSD(s, extended = TRUE)

# get representative, profile-total AWC from SSURGO
sql <- sprintf("SELECT chorizon.cokey AS cokey, 
SUM(awc_r * (hzdepb_r - hzdept_r)) AS ws 
FROM 
legend JOIN mapunit ON legend.lkey = mapunit.lkey
JOIN component ON mapunit.mukey = component.mukey
JOIN chorizon ON component.cokey = chorizon.cokey 
WHERE compname = '%s'
AND areasymbol != 'US'
GROUP BY chorizon.cokey;", s)

# get via SDA
res <- SDA_query(sql)

# median AWC in mm
# over all components correlated to named series 
AWC <- round(median(res$ws, na.rm = TRUE) * 10)

# monthly climate data from series summary
PPT <- x$climate.monthly$q50[x$climate.monthly$variable == 'Precipitation (mm)']
PET <- x$climate.monthly$q50[x$climate.monthly$variable == 'Potential ET (mm)']

# 3 warm-up cycles
# keep last iteration
# calendar year
x.wb <- monthlyWB(AWC, PPT, PET, S_init = 0, starting_month = 1, rep = 3, keep_last = TRUE)

# tighter margins
par(mar=c(4,4,3,1), bg = 'white', mfcol = c(1, 2))

plotWB(x.wb)
title(sprintf('Monthly Water Balance: %s Series', toupper(s)), line = 2)

plotWB_lines(x.wb)
title(sprintf('Monthly Water Balance: %s Series', toupper(s)), line = 2)

```



## Comparisons

Abstract into a function, and use to compare multiple soils / climatic parameters. A more efficient approach would load all of the data then iterate over chunks.
```{r}
# x: soil series name
compareMonthlyWB <- function(x, ylim) {
  
  # get climate data
  osd <- fetchOSD(x, extended = TRUE)
  
  # get representative, profile-total AWC from SSURGO
  sql <- sprintf("
SELECT chorizon.cokey AS cokey, 
SUM(awc_r * (hzdepb_r - hzdept_r)) AS ws 
FROM 
legend JOIN mapunit ON legend.lkey = mapunit.lkey
JOIN component ON mapunit.mukey = component.mukey
JOIN chorizon ON component.cokey = chorizon.cokey 
WHERE compname = '%s'
AND areasymbol != 'US'
GROUP BY chorizon.cokey;", x
  )
  
  # get via SDA
  res <- suppressMessages(SDA_query(sql))
  
  # median AWC in mm
  # over all components correlated to named series 
  AWC <- round(median(res$ws, na.rm = TRUE) * 10)
  
  
  # monthly climate data from series summary
  PPT <- osd$climate.monthly$q50[osd$climate.monthly$variable == 'Precipitation (mm)']
  PET <- osd$climate.monthly$q50[osd$climate.monthly$variable == 'Potential ET (mm)']
  
  # do water balance simulation
  x.wb <- monthlyWB(AWC, PPT, PET, S_init = 0, starting_month = 1, rep = 3, keep_last = TRUE)
  
  # compose WB figure
  plotWB(WB = x.wb, legend.cex = 0.9, ylim = ylim)
  
  # annotate with series name / family level classification
  s <- site(osd$SPC)
  txt <- sprintf("%s\n%s", s$id, s$family)
  title(txt, line = 3, cex.main = 1)
}
```

```{r fig.width=14, fig.height=12}
# custom margins / multi-figure
par(mar=c(4,4,5,1), bg = 'white', mfrow = c(2, 2))

# iterate over these soil series names
soils <- c('amador', 'lucy', 'drummer', 'pierre')
for(i in soils) {
  # adjust ylim manually
  compareMonthlyWB(i, ylim = c(-250, 180))
}
```

```{r fig.width=14, fig.height=12}
# custom margins / multi-figure
par(mar=c(4,4,5,1), bg = 'white', mfrow = c(2, 2))

# iterate over these soil series names
soils <- c('leon', 'tristan', 'miami', 'freznik')
for(i in soils) {
  # adjust ylim manually
  compareMonthlyWB(i, ylim = c(-180, 190))
}
```


----------------------------
This document is based on `sharpshootR` version `r utils::packageDescription("sharpshootR", field="Version")` and `soilDB` version `r utils::packageDescription("soilDB", field="Version")`.
