---
title: "OSD Dendrogram via Soil Taxonomy"
author: "D.E. Beaudette"
date: "`r Sys.Date()`"
output:
  html_document:
   mathjax: null
   jquery: null
   smart: no
   keep_md: no
---
  
```{r setup, echo=FALSE, results='hide', warning=FALSE}
# setup
library(knitr, quietly=TRUE)
opts_chunk$set(message=FALSE, warning=FALSE, background='#F7F7F7', fig.align='center', dev='svglite', tidy=FALSE, verbose=FALSE)
options(width=100, stringsAsFactors=FALSE, cache=TRUE)
```


```{r get-data}
library(aqp)
library(soilDB)
library(sharpshootR)
library(dendextend)

# soils of interest
s.list <- c('hornitos', 'perkins', 'argonaut', 'inks', 'mokelumne', 'dunstone', 'auburn', 'pentz', 'pardee', 'peters', 'amador', 'laniger')

# fetch and convert data into an SPC
s <- fetchOSD(s.list)
```


```{r check-via-plotting, fig.width=16, fig.height=6}
# estimate soil depth based on horizon designation
s$soil.depth <- profileApply(s, estimateSoilDepth, name='hzname')

# plot profiles, order by soil depth
par(mar=c(0, 0, 0, 0))
plotSPC(s, name='hzname', plot.order=order(s$soil.depth), cex.names=0.85, axis.line.offset=-5, width=0.25, id.style='top', name.style = 'center-center', plot.depth.axis = FALSE, hz.depths = TRUE)
```

```{r plot-dendrogram, fig.width=14, fig.height=6}
# plot dendrogram + profiles
# result is the distance matrix used to generate dendrogram
d <- SoilTaxonomyDendrogram(s, cex.taxon.labels=0.8, width=0.25, y.offset = 0.4)
```


Approximate sorting of dendrogram according to catenary position ([hillslope position proportions](https://www.nrcs.usda.gov/wps/portal/nrcs/detail/?cid=nrcs142p2_054252#describe)).
```{r fig.width=8, fig.height=4}
# TODO: try norforlk and siblings
# http://soilmap2-1.lawr.ucdavis.edu/dylan/soilweb/sde/?series=norfolk

# set of soil series to query
soils <- c('cecil', 'altavista', 'lloyd', 'wickham', 'wilkes',  'chewacla', 'congaree')

# get morphology + extended summaries
s <- fetchOSD(soils, extended = TRUE)

# extract catenary relationships from hillslope proportion summary
# and display
res <- vizHillslopePosition(s$hillpos)
print(res$fig)
```

Organize ST dendrogram according to catenary position.
```{r fig.width=12, fig.height=12}
par(mar=c(0,0,1,1), mfrow=c(2,1))

SoilTaxonomyDendrogram(s$SPC, width=0.25)
mtext('default sorting', side = 2, line=-1, font=3, cex=1.25)

# labels used to re-order (if possible) terminal branches of dendrogram
labs <- profile_id(s$SPC)[res$order]

SoilTaxonomyDendrogram(s$SPC, rotationOrder = labs, width=0.25)
mtext('approx. catenary sorting', side = 2, line=-1, font=3, cex=1.25)
```


A classic *chronosequence* from the San Joaquin Valley, CA (Jenny, 1981).
```{r fig.width=8, fig.height=6}
# series of interest, in order from youngest -> oldest
s <- c('tujunga', 'hanford', 'greenfield', 'snelling', 'san joaquin')
osds <- fetchOSD(s)

# ordering vector for sketches
idx <- match(toupper(s), profile_id(osds))

# encode horizon boundary distinctness via vertical offset
osds$hd <- hzDistinctnessCodeToOffset(
  osds$distinctness,
  codes=c('very abrupt', 'abrupt', 'clear', 'gradual', 'diffuse')
)

# encode horizon boundary topography via vertical offset
osds$hzto <- hzTopographyCodeToOffset(
  osds$topography,
  codes = c('smooth', 'wavy', 'irregular', 'broken')
)

# also encode horizon boundary topography las line type
osds$hzto.lty <- hzTopographyCodeToLineType(
  osds$topography,
  codes = c('smooth', 'wavy', 'irregular', 'broken')
)


# concise representation of hz bnd distinctness and topography
# similar to field notes
osds$bnd.code <- sprintf(
  "%s%s",
  substr(osds$distinctness, 1, 1),
  substr(osds$topography, 1, 1)
)

# remove NA
osds$bnd.code <- gsub('NANA', '', osds$bnd.code)

par(mar = c(0, 0, 0, 1), bg = 'black', fg = 'white')

plotSPC(
osds,
plot.order = idx,
width = 0.3,
name.style = 'center-center',
cex.names = 0.66,
plot.depth.axis = FALSE,
hz.depths = TRUE,
shrink = TRUE,
hz.distinctness.offset = 'hd',
hz.topography.offset = 'hzto',
hz.boundary.lty = 'hzto.lty'
)

legend(
'bottomright',
horiz = TRUE,
legend = c('Smooth', 'Wavy', 'Irregular', 'Broken'),
lty = 1:4,
inset = 0.05,
bty = 'n',
cex = 0.85
)

# note that `rotationOrder` uses the ordering of series names (uppercase to match profile IDs)
# to re-order the terminal branches of the dendrogram
SoilTaxonomyDendrogram(
osds,
rotationOrder = toupper(s),
cex.taxon.labels = 0.85,
width = 0.3,
name.style = 'center-center',
cex.names = 0.66,
plot.depth.axis = FALSE,
hz.depths = TRUE,
shrink = TRUE,
hz.distinctness.offset = 'hd',
hz.topography.offset = 'hzto',
hz.boundary.lty = 'hzto.lty'
)

legend('bottomright', horiz = TRUE, legend = c('Smooth', 'Wavy', 'Irregular', 'Broken'), lty = 1:4, inset = 0.01, bty = 'n', cex = 0.85)

```


----------------------------
This document is based on `aqp` version `r utils::packageDescription("aqp", field="Version")`, `soilDB` version `r utils::packageDescription("soilDB", field="Version")`, and `sharpshootR` version `r utils::packageDescription("sharpshootR", field="Version")`.
