---
title: "Evaluating Raster Sampling Density"
author: "D.E. Beaudette"
date: "`r Sys.Date()`"
output:
  html_document:
    mathjax: null
    jquery: null
    smart: no
    keep_md: no
---

```{r setup, echo=FALSE, results='hide', warning=FALSE}
# setup
library(knitr)
opts_chunk$set(message=FALSE, warning=FALSE, background='#F7F7F7', fig.retina=2, dev='png', tidy=FALSE, verbose=FALSE, cache = FALSE)
options(width=100, stringsAsFactors=FALSE, cache=FALSE)
```

The following can be used to evaluate the influence of sampling density on the stability of estimates derived from a raster data source. Adjust `mu` (map unit polygons), `mu.set` (select map unit symbols), and `r` (raster data source) accordingly.
```{r}
library(latticeExtra)
library(tactile)
library(plyr)
library(rgdal)
library(raster)
library(sharpshootR)
library(parallel)

# load map unit polygons
mu <-  readOGR(dsn='E:/gis_data/ca630/FG_CA630_OFFICIAL.gdb', layer='ca630_a', stringsAsFactors = FALSE)

# best possible scenario: rasters are in memory
r <- readAll(raster('E:/gis_data/ca630/ca630_slope/hdr.adf'))

# subset to just those map units we are interested in
mu.set <- c('5012', '5201', '7089', '3058', '6054', '7083')
mu <- mu[mu$MUSYM %in% mu.set, ]

# add a unique polygon ID
mu$pID <- seq(from=1, to=length(mu))

# split select map units into a list of SPDF objs
# will be processed in parallel
mu.list <- split(mu, mu$MUSYM)

# init cluster
cl <- makeCluster(4)
setDefaultCluster(cl)

clusterEvalQ(NULL, {
  library(raster)
  library(sharpshootR)
  })

# this function is run on each core
f.par <- function(i) {
  # eval stability
  s <- samplingStability(i, r, n.set = c(0.5, 1, 2, 5), n.reps = 10)
  # keep only stability indices
  s <- s$stability
  # convert fractions to percent
  s$stability <- s$stability * 100
  # http://stackoverflow.com/questions/12023403/using-parlapply-and-clusterexport-inside-a-function/12024448#12024448
  gc()
  return(s)
}

# export objects that must be shared across cores
clusterExport(cl=cl, varlist=c("r", "f.par"))

# parallel version 2-3x faster ~ 5.4 minutes
system.time(mu.stability <- parLapply(cl, mu.list, f.par))

# done with multiple cores
stopCluster(cl)

# convert to DF
mu.stability <- ldply(mu.stability)

# compute increase in stability from lowest to highest sampling density
delta.stability <- ddply(mu.stability, '.id', function(i) {
  return(-sum(diff(i$stability)))
})
names(delta.stability) <- c('MUSYM', 'Increase in Stability')

# area summary
a <- ldply(mu.list, function(i) {
  res <- sum(sapply(slot(i, 'polygons'), slot, 'area') * 2.47e-4)
  return(res)
})
names(a) <- c('MUSYM', 'Map Unit Acreage')
a <- a[order(a[, 2], decreasing = TRUE), ]

# join-in stability increase
a <- join(a, delta.stability, by='MUSYM')
```

The following table contains a summary of map unit acreage and total increase in stability (from lowest to highest sampling density). In general, smaller map unit polygons will require a higher sampling density.
```{r}
kable(a, digits = 0, row.names = FALSE)
```

The following figure demonstrates the relationship between stability and sampling density. For most map units a sampling density between 1-2 points per acre should suffice. A larger sampling density (e.g. 5 points per acre) should be used for very small delineations (e.g. less than 5-10 acres).
```{r fig.width=8, fig.height=6, fig.cap='Stability of sample median (10m slope raster) for slect map units.'}
# custom tick, label, and line positions
h.lines <- c(0:4, seq(5, 30, by=2))
# custom plot style
tps <- tactile.theme(superpose.symbol=list(pch=15, cex=1.5), superpose.line=list(lwd=2))

xyplot(stability ~ factor(sampling.density), groups=.id, data=mu.stability, type=c('b'), scales=list(alternating=3, y=list(at=h.lines)), ylab='Percent of Population Median', xlab='Sampling Density (pts/ac.)', strip=strip.custom(bg=grey(0.85)), main='90% Confidence in Stability of (Sample) Median', auto.key=list(columns=3, lines=TRUE, points=FALSE), par.settings=tps, panel=function(...) {
  panel.abline(h=h.lines, v=1:4, lty=3, col='grey')
  panel.xyplot(...)
})
```
