---
title: "Pedon Data with R Cheatsheet"
subtitle: 'v0.1.0 - last updated 2019/01/31'
author: "Andrew Brown; andrew.g.brown@ca.usda.gov"
version: "0.1.0"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

# Pedon Data with R Cheatsheet

This "cheatsheet" provides explanation of some of the most common workflows and techniques you will need when using the _aqp_ __SoilProfileCollection__ object to hold soil profile (pedon) data.

To get this document as an .R file [click here](spc-cheatsheet.R).

## NCSS-Tech (Soil-related) Packages

With `install.packages()` you can get the "stable" versions of _aqp_, _soilDB_ and _sharpshootR_ off of the official R package repository (CRAN) .

```{r, eval=F, warning=F, message=F}
install.packages(c("aqp","soilDB","sharpshootR"))
```

You can get the latest "development" versions of ncss-tech packages from GitHub. 

```{r, eval=F, warning=F, message=F}
# install.packages("devtools") # if needed
devtools::install_github(repo = 'ncss-tech/aqp', dependencies = FALSE, build = FALSE)
devtools::install_github(repo = 'ncss-tech/soilDB', dependencies = FALSE, build = FALSE)
devtools::install_github(repo = 'ncss-tech/sharpshootR', dependencies = FALSE, build = FALSE)
```

Note that _soilReports_ package (containing, for instance, the mapunit comparison report) is ONLY available __FROM GitHub__.

```{r, eval=F, warning=F, message=F}
devtools::install_github(repo = 'ncss-tech/soilReports', dependencies = FALSE, build = FALSE)
```

After installing, load the packages with `library()` to use them in your R environment.

```{r, eval=T, warning=F, message=F}
library(aqp)
library(soilDB)
library(sharpshootR)
library(soilReports)
```

# SoilProfileCollection

## SPC from soilDB sample dataset

Load one of the sample (NASIS) datasets from _soilDB_ package

```{r, eval=F, warning=F, message=F}
data("loafercreek")
```

## SPCs from NASIS

Get data from NASIS (requires ODBC connection to local NASIS db). You can create SPCs from pedon (point data) or aggregated data (map unit components).

```{r, eval=F, warning=F, message=F}
fetchNASIS() # default argument of fetchNASIS() gets pedons
fetchNASIS_pedons()
fetchNASIS_components()
```

## SPCs from Kellogg Soil Survey Lab database

KSSL Data by taxon name 
```{r, eval=F, warning=F, message=F}
fetchKSSL(series = "Holland")
```

KSSL Data by MLRA (overlap of site coordinates with MLRA shapefile)
```{r, eval=F, warning=F, message=F}
fetchKSSL(mlra = "15")
```

KSSL Data by rectangular bounding box (specified in longitude and latitude)
```{r, eval=F, warning=F, message=F}
fetchKSSL(bbox = c(-119, 37, -120, 38))
```

## SPCs from SSURGO/STATSGO (SDA) components 

Fetch component and component horizon information from Soil Data Access (interface to SSURGO/STATSGO) subject to a user-specified WHERE constraint.

```{r, eval=F, warning=F, message=F}
fetchSDA_component(WHERE = "compname = 'Mantree'")
```

See the "SDA Query Help" page for information on database schema, table/column definitions, etc.

[https://sdmdataaccess.nrcs.usda.gov/QueryHelp.aspx](https://sdmdataaccess.nrcs.usda.gov/QueryHelp.aspx)

You can also evaluate _arbitrary_ SDA queries (where you do all the joins to tables of interest _yourself_).
```{r, eval=F, warning=F, message=F}
SDA_query(q = "SELECT * FROM legend AS l 
                INNER JOIN mapunit mu ON l.lkey = mu.lkey
                INNER JOIN component co ON mu.mukey = co.mukey
               WHERE areasymbol = 'CA630' AND compname = 'Mantree';")
```

## SPCs from Soil Classification database

Create an SPC containing OSDs from Soil Classification (SC) database (an input vector of series names is provided).

```{r, eval=F, warning=F, message=F}
your.spc <- fetchOSD(soils = c("Holland", "Musick"))
```

Summarize climatic/physiographic/component attribute breakdown and get morphologic data using argument `extended=TRUE`.

```{r, eval=F, warning=F, message=F}
your.extended.list <- fetchOSD(soils = "Holland", extended = TRUE) 

#the SPC can be found in the SPC attribute when extended=TRUE
your.extended.list$SPC
```

# Accessing SPCs

Get site _data.frame_ from `loafercreek` SPC.

```{r, eval=F, warning=F, message=F}
site(loafercreek)
```

These are two equivalent expressions for finding the __number of sites/profiles__ in an SPC.
```{r,eval=F, warning=F, message=F}
nrow(site(loafercreek))
length(loafercreek) # length(spc) is an alias for nrow(site(spc))
```

Get horizon _data.frame_ from `loafercreek` SPC.
 
```{r, eval=F, warning=F, message=F}
horizons(loafercreek)
```

These are two equivalent expressions for finding the __number of horizons__ in an SPC.
```{r,eval=F, warning=F, message=F}
nrow(horizons(loafercreek))
nrow(loafercreek) # nrow(spc) is an alias for nrow(horizons(spc))
```

Create a subset SPC based on a site-level attribute using`subsetProfiles()` and argument `s` containing an expression to evaluate. Here we write an expression to select all profiles that have a field-measured slope gradient _less than 8 percent_.

```{r, eval=F, warning=F, message=F}
sub1 <- subsetProfiles(loafercreek, s='slope_field < 8')
```

Create a subset SPC based on a horizon-level attribute using`subsetProfiles()` and argument `h` containing an expression to evaluate. Here we write an expression select all profiles that have _one or more horizons_ with clay content greater than 35 percent.

```{r, eval=F, warning=F, message=F}
sub2 <- subsetProfiles(loafercreek, h = 'clay > 35')
```

## Create a subset SPC based on a list of userpedonIDs

First, make a character vector (using `c()`) to concatenate a set of character strings (the userpedonIDs that you want).
```{r, eval=F, warning=F, message=F}
your.list <- c("07SKC016","10MJE016","10CKS034","S2000CA007012","06SMM013" )
```

Create a `TRUE`/`FALSE` (logical) vector by checking each userpedonID in `loafercreek` to see if it matches one in `your.list`. 

```{r, eval=F, warning=F, message=F}
matches.your.list <- site(loafercreek)$pedon_id %in% your.list 
```

Note that the length of the vector `matches.your.list` equals `nrow(site(loafercreek))`. 

The number of `TRUE` values in `matches.your.list` (`sum(matches.your.list)`) is equal to `length(your.list)`. This is because all profile IDs in the list were found. If `sum(matches.your.list) < length(your.list)` some in the list were NOT found.

To get just the positions (index) where the logical vector `matches.your.list` is `TRUE` use `which()`

```{r, eval=F, warning=F, message=F}
your.matching.indices <- which(matches.your.list)
```

Create a new SPC `loafercreek.subset` that is the subset corresponding to your "matching" indices. 

Use the _data.frame_-like bracket notation `spc[site.index, horizon.index]` to access the sites corresponding to your matching userpedonIDs.

```{r, eval=F, warning=F, message=F}
loafercreek.subset <- loafercreek[your.matching.indices, ]
```

# Editing SPCs

Create a site or horizon-level variable.
```{r, eval=F, warning=F, message=F}
loafercreek$foo <- 1:nrow(site(loafercreek))

loafercreek$foohz <- 1:nrow(horizons(loafercreek))

# note that this one causes an error. Test your understanding: why?
horizons(loafercreek)$foohz2 <- 1:nrow(site(loafercreek))
```

Create a site or horizon-level variable _without having replacement equal to number of sites/horizons_ by explicitly calling `site()` or `horizons()`. Here since the replacement is length `1` it is set for all sites/horizons. This is the only way to set a site or horizon attribute that is not equal to the length of the respective _data.frame_.

```{r, eval=F, warning=F, message=F}
site(loafercreek)$foo <- 2
horizons(loafercreek)$foo <- 2
```

Add multiple attributes to the site data of a SPC based on common (unique) site id (for example: `id`). An SPC is smart enough to to _LEFT JOIN_ on the shared unique site ID (must be present in new data data.frame `your.site.data`). 

This type of site level assignment results in an SPC with the unique site level attributes resulting from the join. Note we are assiging the new datato a call to `site()` on the SPC we are editing.

```{r, eval=F, warning=F, message=F}
site(your.spc) <- your.site.data
```

Adding horizon data to the SPC based on common (unique) horizon id (for example: `hzid`) is a little more complicated. You need to be careful when using direct replacement (like below) to make sure that you do not accidentally remove horizon records. Force all records from the original data to be retained by setting `all.x = TRUE`

```{r, eval=F, warning=F, message=F}
horizons(your.spc) <- merge(horizons(your.spc), your.horizon.data, by = "hzid", all.x=TRUE)
```

In the future, SPCs will allowq 

# Combining SPCs

Load _another_ SPC to combine with `loafercreek`

```{r, eval=T, warning=F, message=F}
data("loafercreek")
data("gopheridge")
```

Make a `list()` of SPCs, and combine them with `union()`

```{r, eval=F, warning=F, message=F}
loafergopher <- union(list(loafercreek, gopheridge))
```

# Plotting SPCs

Make a plot of first 10 profiles in `loafercreek`. `plot(your.spc)` is an alias (shorthand) for `plotSPC()`

```{r, eval=F, warning=F, message=F}
plot(loafercreek[1:10, ], label = "pedon_id")
```

Make a plot of first 10 profiles in `loafercreek`, this time grouped by _taxonname_. 

```{r, eval=T, warning=F, message=F}
groupedProfilePlot(loafercreek[1:10, ], groups = "taxonname", 
                  label = "pedon_id", id.style="side",
                  axis.line.offset = -1.5, y.offset=25)
```

To make this plot "nice" we set the default label position style to "side" and label with the the `pedon_id` (_NASIS userpedonID_) instead of `peiid` (_NASIS pedon record id_). 

We adjust the depth axis offset (units of "profile index" - the _x_ dimension) to the right. This is done so the axis does not plot on top of the horizon designations of the farthest profile to the right.

Also, we set the vertical (_y_ dimension) offset of top of profiles (specified using depth units, _centimeters_ in this case) so the top of the profiles/y-axis do not overlap the group labels containing the _taxonname_.

There are _many, many more_ plotting options that can be found in the helpfiles. See `?plotSPC` and `?groupedProfilePlot`.

# Making Spatial SPCs

## Set coordinates and reference system for NASIS-derived pedon

Set coordinates to be NASIS standard decimal degrees longitude `x_std` and latitude `y_std`. You provide the right hand side of a formula that describes how to promote the SPC to spatial with `coordinates()`.

```{r, eval=F, warning=F, message=F}
coordinates(loafercreek) <- ~ x_std + y_std
```

You need to set the [_initially blank_] coordinate reference system (CRS) to the NASIS standard (WGS84 decimal degrees). We use a string representation of the coordinate reference system called the "proj4string" to specify what want.

```{r, eval=F, warning=F, message=F}
proj4string(loafercreek) <- '+proj=longlat +datum=WGS84'
```

If you specify _longitude_ first (`x_std`) in your formula to `coordinates()`, use `longlat` in your proj4string. Otherwise use `latlong`. 

If your coordinates are in a different system (e.g. projected in UTM) you will need to specify e.g. UTM zone in the `proj4string()` you supply. Check out `?proj4string` for more info on how to use other CRSs.

## Working with spatial data

Create a _SpatialPointsDataFrame_ from SPC (contains just __SITE__-level data)

```{r, eval=F, warning=F, message=F}
loafercreek.spdf <- as(loafercreek, 'SpatialPointsDataFrame')
```

Plot site locations in _SpatialPointsDataFrame_

```{r, eval=F, warning=F, message=F}
plot(loafercreek.spdf)
```

Write _SpatialPointsDataFrame_ derived from an SPC to a shapefile (_do not use '.shp' extension in `layer` argument that specifies the file name_)

```{r, eval=F, warning=F, message=F}
rgdal::writeOGR(obj = loafercreek.spdf, layer = 'output.shapefile.name', driver = 'ESRI Shapefile')
```