```{r setup, echo=FALSE, results='hide'}
opts_chunk$set(message=FALSE, warning=FALSE, background='#F7F7F7', dpi=100, fig.align='center', dev='png', dev.args=list(pointsize=10), tidy=FALSE)
options(width=100, stringsAsFactors=FALSE)
```

**Load data**
```{r load-data, results='hide'}
library(aqp)
library(soilDB)
library(plyr)
library(lattice)
library(maps)
library(digest)

# colors and line styles for later:
cols <- c('RoyalBlue', 'DarkGreen', 'Orange', 'Red')

# load sample data set
data(amarillo)

# extract pieces
l <- amarillo$lab
n <- amarillo$dmu
```


**pre-process NASIS/DMU/SSURGO data**
OK. It looks like there are a couple of things in the NASIS/DMU/SSURGO data:
* Duplication of horizon-level attributes: this is common, as components are frequently copied. Fixed this problem with MD5 digests to test for unique-ness
* The `cokey` column isn't really a `cokey`. Fixed it by pasting together with `mukey`.
* The data aren't in order. Fixed it by re-ordering.

```{r nasis-data-prep}
# it looks like there may be some duplicates in these data, same cokey among several mukey values
# make a proper cokey as `mukey:cokey`
n$cokey <- as.character(paste(n$mukey, n$cokey, sep=':'))

# re-order by id/depth
n <- n[order(n$cokey, n$top), ]

# normalize component / taxon name
n$taxon_name <- tolower(n$Component.Name)

# convert to SoilProfileCollection objects and extract some site-level data
depths(n) <- cokey ~ top + bott
site(n) <- ~ taxon_name

# check for duplicate records from nasis (using only horizon attributes of interest)
v <- c('top', 'bott', 'PHWR', 'CLAYR', 'CECR', 'PHWL', 'CLAYL', 'CECL', 'PHWH', 'CLAYH', 'CECH')
# this is just an index to unique profiles
u.profiles <- unique(n, vars=v)

# keep only the unique profiles [this takes a while]
n.pct.duplication <- round(length(u.profiles) / length(n) * 100)
n <- n[u.profiles, ]

# give some feedback degree of duplication
n.dupes <- paste(n.pct.duplication, '% duplication in DMU records', sep='')
print(n.dupes)
```

**pre-process lab data**
Data issues:
* The data aren't in order. Fixed it by re-ordering.
* Some minor horizonation problems, due to missing bottom horizon depths. Fixed by using the corresponding top depth (lab data only).

```{r lab-data-prep}
# re-order by id/depth
l <- l[order(l$user_pedon_id, l$hzn_top), ]

# some bottom horizon depths are 0 OR missing
# fix them by repeating the corresponding horizon top
l.bottom.zero <- which(l$hzn_bot == 0)
l.bottom.na <- which(is.na(l$hzn_bot))
l$hzn_bot[l.bottom.zero] <- l$hzn_top[l.bottom.zero]
l$hzn_bot[l.bottom.na] <- l$hzn_top[l.bottom.na]

# normalize taxonname
l$taxon_name <- tolower(l$taxon_name)

# convert to SoilProfileCollection objects and extract some site-level data
depths(l) <- user_pedon_id ~ hzn_top + hzn_bot
site(l) <- ~ user_site_id + taxon_name + class_type + longitude_std_decimal_degrees + latitude_std_decimal_degrees
```


**The lab data have coordinates, map them.**
```{r lab-data-map, fig.height=5, results='hide'}
# init spatial data (no missing coordinates allowed!)
coordinates(l) <- ~ longitude_std_decimal_degrees + latitude_std_decimal_degrees
proj4string(l) <- '+proj=longlat +datum=NAD83'

# make map
map('state')
points(as(l, 'SpatialPointsDataFrame'), col=NA, bg=rgb(1,0,0,0.5), cex=0.5, pch=21)
```




**aggregate pH, total clay, and CEC by 1-cm slices**
This can take a while, why not enjoy a short coffee break? Alternatively, this could be done as 2 steps instead of 4-- however it would require some extra fussing around at the expense of a little computer time. When you run these lines, you may see some warnings; these are related to missing horizons within a profile. Safe to ignore for now.
```{r aggregate-data, results='hide'}
# lab data
l.a <- slab(l, taxon_name ~ ph_h2o + Total.Clay + CEC, progress='text')
# nasis data: RVs
n.a <- slab(n, taxon_name ~ PHWR + CLAYR + CECR, progress='text')
# nasis data: LOWs
n.a.low <- slab(n, taxon_name ~ PHWL + CLAYL + CECL, progress='text')
# nasis data: HIGHs
n.a.high <- slab(n, taxon_name ~ PHWH + CLAYH + CECH, progress='text')
```

**combine lab data and nasis data**
This is important. It is possible to combine the two (i.e. stack them) because they have identical formats now. The output from `slab()` always give the same format, regardless of what the input data are. The only caveat is that we have to change the variable names in one dataset to match the other. In this case, we change the NASIS/DMU/SSURGO data: PHWR->ph_h2o, CLAYR->Total.Clay, CECR->CEC. Same process for the low and high values.

```{r stack-agg-data}
# make sure that the ordering of variables is the same between datasets: OK
cbind(lab=levels(l.a$variable), 
			nasis.rv=levels(n.a$variable), 
			nasis.low=levels(n.a.low$variable),
			nasis.high=levels(n.a.high$variable))

# fix nasis variable names: this shortcut works because the ordering is the same!
n.a$variable <- factor(n.a$variable, levels=levels(n.a$variable), labels=levels(l.a$variable))
n.a.low$variable <- factor(n.a.low$variable, levels=levels(n.a.low$variable), labels=levels(l.a$variable))
n.a.high$variable <- factor(n.a.high$variable, levels=levels(n.a.high$variable), labels=levels(l.a$variable))

# stack: the source lable is stored in the 'which' column
g <- make.groups(lab=l.a, nasis.rv=n.a, nasis.low=n.a.low, nasis.high=n.a.high)

# optional:
# since the NASIS/DMU/SSURGO low-rv-high values aren't likely independent, 
# throw-out the q5 and q95 values for a cleaner plot
g$p.q5[g$which %in% c('nasis.rv', 'nasis.low', 'nasis.high')] <- NA
g$p.q95[g$which %in% c('nasis.rv', 'nasis.low', 'nasis.high')] <- NA
```


**visual check on aggregate properties**
These figures depict the slice-wise range in values (q5-median-q95) for lab data, along with the median values from the nasis low-rv-high data.
```{r plot-agg-data, fig.width=10, fig.height=14, out.width='100%'}
xyplot(top ~ p.q50 | variable, groups=which, data=g, ylab='Depth',
			 xlab='median bounded by 5th and 95th percentiles',
			 lower=g$p.q5, upper=g$p.q95, ylim=c(250,-5), alpha=0.25,
			 strip=strip.custom(bg='yellow'), layout=c(3,1),
			 par.settings=list(superpose.line=list(col=cols, lwd=2)),
			 as.table=TRUE, scales=list(alternating=3, x=list(relation='free')),
			 panel=panel.depth_function, sync.colors=TRUE,
			 prepanel=prepanel.depth_function,
			 auto.key=list(columns=4, lines=TRUE, points=FALSE)
)
```
*****

