---
title: "Soil Color Tiles"
author: "D.E. Beaudette"
date: "`r Sys.Date()`"
output:
  html_document:
    mathjax: null
    jquery: null
    smart: no
---
  
```{r setup, echo=FALSE, results='hide', warning=FALSE}
library(knitr)
opts_chunk$set(message=FALSE, warning=FALSE, background='#F7F7F7', dev='svglite', tidy=FALSE, verbose=FALSE, cache = FALSE, fig.align = 'center')
options(width=100, stringsAsFactors=FALSE, cache=FALSE)
```

[Clarksville](https://casoilresource.lawr.ucdavis.edu/sde/?series=clarksville).
```{r}
library(aqp)
library(soilDB)

# get lab / morphologic data
x <- fetchKSSL(series = 'clarksville', returnMorphologicData = TRUE, simplifyColors = TRUE)

# extract SoilProfileCollection
s <- x$SPC

# remove horizons that are missing moist colors
s <- subsetHz(s, !is.na(m_hue) & !is.na(m_value) & !is.na(m_chroma))

# remove profiles with missing horizons due to above steps
s <- HzDepthLogicSubset(s)

# keep only profiles with > 2 horizons
idx <- which(profileApply(s, nrow) > 2)
s <- s[idx, ]

# re-assemble Munsell color notation for moist color
s$color <- sprintf("%s %s/%s", s$m_hue, s$m_value, s$m_chroma)
```

One approach: extract the first $n$ profiles having > 5 horizons, and first 6 horizons.
```{r fig.width=8, fig.height=6.5}
# filter to just those profiles with > 5 horizons
idx <- which(profileApply(s, nrow) > 5)

# take the first n profiles, all horizons
n <- 6
w <- s[idx[1:n], 1:6]

# check
par(mar=c(0,0,0,0))
plotSPC(w, color = 'moist_soil_color', cex.names = 0.75, width = 0.35, name.style = 'center-center', plot.depth.axis = FALSE, hz.depths = TRUE, fixLabelCollisions = TRUE, hz.depths.offset = 0.06)
```

Less restrictive: first $n$ profiles, all horizons.
```{r fig.width=8, fig.height=6.5}
n <- 6
w <- s[1:n, ]

# check
par(mar=c(0,0,0,0))
plotSPC(w, color = 'moist_soil_color', cex.names = 0.75, width = 0.35, name.style = 'center-center', plot.depth.axis = FALSE, hz.depths = TRUE, hz.depths.offset = 0.06)
```

Regularize horizons / colors to $k$ rows.
```{r}
regularizeColors <- function(i, k = 6) {
  .col <- colorRampPalette(na.omit(i$moist_soil_color), space = 'Lab', interpolate = 'spline')(k)
  .munsell <- rgb2munsell(t(col2rgb(.col)/255))
  .munsell <- sprintf("%s %s/%s", .munsell$hue, .munsell$value, .munsell$chroma)
  
  return(list(col = .col, munsell = .munsell))
}

mm <- profileApply(w, simplify = FALSE, FUN = regularizeColors, k = 6)
```


Make a $k$ row by $n$ column grid of soil colors from above.
```{r fig.width=3, fig.height=3}

m <- lapply(mm, '[[', 'col')
m <- do.call('cbind', m)

# m <- matrix(m$moist_soil_color, nrow = 6, ncol = 5, byrow = FALSE)

par(mar=c(0,0,0,0))
plot(1, 1, type='n', axes=FALSE, xlab='', ylab='', ylim=c(6.5, 0.5), xlim=c(0.5, n+0.5), asp = 1)

# vectorized functions are the best
rect(xleft = col(m) - 0.4, ybottom = row(m) + 0.4, xright = col(m) + 0.4, ytop = row(m) - 0.4, col = m, border = NA)
```

Annotate with Munsell colors.
```{r fig.width=5, fig.height=5}
.munsell <- lapply(mm, '[[', 'munsell')
.munsell <- do.call('c', .munsell)

par(mar=c(0,0,0,0))
plot(1, 1, type='n', axes=FALSE, xlab='', ylab='', ylim=c(6.5, 0.5), xlim=c(0.5, n+0.5), asp = 1)

# vectorized functions are the best
rect(xleft = col(m) - 0.4, ybottom = row(m) + 0.4, xright = col(m) + 0.4, ytop = row(m) - 0.4, col = m, border = NA)

text(x = col(m), y = row(m), labels = .munsell, col = invertLabelColor(m), cex = 0.66, font = 2)
```

This time with rounded corners
```{r fig.width=3, fig.height=3}
library(berryFunctions)

par(mar=c(0,0,0,0))
plot(1, 1, type='n', axes=FALSE, xlab='', ylab='', ylim=c(6.5, 0.5), xlim=c(0.5, n+0.5), asp = 1)

# ugh
col.m <- col(m)
row.m <- row(m)

# not vectorized
for(i in seq_along(m)) {
  roundedRect(xleft = col.m[i] - 0.4, ybottom = row.m[i] + 0.4, xright = col.m[i] + 0.4, ytop = row.m[i] - 0.4, col = m[[i]], border = NA, rounding = 0.25)
}

```




Arbitrary collection of profiles
```{r fig.width=10, fig.height=4.75}
library(sharpshootR)
data("OSDexamples")

s <- OSDexamples$SPC
s$color <- sprintf("%s %s/%s", s$hue, s$value, s$chroma)

s$moist_soil_color <- s$soil_color

mm <- profileApply(s, simplify = FALSE, FUN = regularizeColors, k = 8)

m <- lapply(mm, '[[', 'col')
m <- do.call('cbind', m)

n <- length(s)

# init plot region
par(mar=c(0,0,0,0))
plot(1, 1, type='n', axes=FALSE, xlab='', ylab='', ylim=c(6.5, 0.5), xlim=c(0.5, n+0.5), asp = 1)

# keep track of column and row indices for simpler plotting
col.m <- col(m)
row.m <- row(m)

# roundedRect() not vectorized
for(i in seq_along(m)) {
  roundedRect(
    xleft = col.m[i] - 0.4, 
    ybottom = row.m[i] + 0.4, 
    xright = col.m[i] + 0.4, 
    ytop = row.m[i] - 0.4, 
    col = m[[i]], 
    border = NA, 
    rounding = 0.25
  )
}

```
[US State Soils](https://www.nrcs.usda.gov/resources/education-and-teaching-materials/state-soils)
```{r fig.width=10, fig.height=4.75}
library(sharpshootR)
data("us.state.soils")

s <- fetchOSD(us.state.soils$series)

# splice-in state names
s$series <- site(s)$id
us.state.soils$series <- toupper(us.state.soils$series)
site(s) <- us.state.soils

s$color <- sprintf("%s %s/%s", s$hue, s$value, s$chroma)

s$moist_soil_color <- s$soil_color

mm <- profileApply(s, simplify = FALSE, FUN = regularizeColors, k = 8)

m <- lapply(mm, '[[', 'col')
m <- do.call('cbind', m)

n <- length(s)
```

```{r results='hide', fig.show='hide'}
# hide this plot
d <- SoilTaxonomyDendrogram(s, KST.order = TRUE)
```

```{r fig.width=20, fig.height=4.75}
# re-order based on subgroup classification
m <- m[, d$order]

# pdf(file = 'state-soils-chips.pdf', width = 18, height = 5)

# init plot region
par(mar=c(0,0,0,0.5))
plot(1, 1, type='n', axes=FALSE, xlab='', ylab='', ylim=c(6.5, 0.5), xlim=c(0.5, n+0.5), asp = 1)

# keep track of column and row indices for simpler plotting
col.m <- col(m)
row.m <- row(m)

# roundedRect() not vectorized
for(i in seq_along(m)) {
  roundedRect(
    xleft = col.m[i] - 0.4, 
    ybottom = row.m[i] + 0.4, 
    xright = col.m[i] + 0.4, 
    ytop = row.m[i] - 0.4, 
    col = m[[i]], 
    border = NA, 
    rounding = 0.25
  )
}

# annotate 
# text(x = seq_along(s), y = 0.33, labels = site(s)$id[d$order], cex = 0.66, srt = 90, adj = 0, font = 2)
# text(x = seq_along(s), y = 0.33, labels = s$subgroup[d$order], cex = 0.5, srt = 45, adj = 0, font = 2)

# text(x = seq_along(s), y = 0.33, labels = s$state[d$order], cex = 0.66, srt = 45, adj = 0, font = 2)

# dev.off()
```

----------------------------
This document is based on `aqp` version `r utils::packageDescription("aqp", field="Version")` and `soilDB` version `r utils::packageDescription("soilDB", field="Version")`.
