---
title: "Pair-Wise Distances by Generalized Horizon Labels"
author: "D.E. Beaudette"
date: "`r Sys.Date()`"
output:
  html_document:
    mathjax: null
    jquery: null
    smart: no
---
  
```{r setup, echo=FALSE, results='hide', warning=FALSE}
library(knitr)
opts_chunk$set(message=FALSE, warning=FALSE, background='#F7F7F7', fig.retina=2, dev='png', tidy=FALSE, verbose=FALSE, cache = FALSE, fig.align = 'center')
options(width=100, stringsAsFactors=FALSE, cache=FALSE)
```


Question: is it possible to generate a reasonable sorting of soil profiles (a `SoilProfileCollection`) using pair-wise distances computed only from generalized horizon labels (GHL)? I think that it might, as long as these labels are encoded as [ordered factors](http://ncss-tech.github.io/AQP/aqp/profile-compare-factors.html). We can use `profile_compare` to test this hypothesis. One might attempt such a comparison within a collection of similar soils where the only reliable data available are horizon depths and designations.

We will be using a subset of the the `loafercreek` sample data from the {soilDB} package.
```{r, fig.width=12, fig.height=6}
# latest version from github
library(aqp)
library(soilDB)
library(sharpshootR)

library(cluster)
library(dendextend)
library(MASS)

data("loafercreek")

# profiles 20-40
x <- loafercreek[20:40, ]

# modify default arguments for some flair
par(mar = c(0, 0, 0, 0))
plotSPC(x, width = 0.3, name.style = 'center-center', plot.depth.axis = FALSE, hz.depths = TRUE, cex.names = 0.65, print.id = FALSE)
```


Apply generalized horizon labels using overly-simplistic rules. This is only a demonstration.
```{r}
# new labels
n <- c('O', 'A', 'B', 'Bt', 'BC', 'Cr', 'R')
# GHL patterns
p <- c('O', 'A', 'B', 'Bt', 'BC', 'Cr', 'R')

# note additional argument
horizons(x)$genhz <- generalize.hz(x$hzname, new = n, pat = p)

# the most important step, genhz must be encoded as an ordered factor
x$genhz <- ordered(x$genhz)

# check frequency of GHL assignment
table(x$genhz)
```


Remove profiles containing horizons assigned to the "not-used" GHL.
```{r, fig.width=12, fig.height=6}
x$flag <- profileApply(x, function(i) {
  any(i$genhz == 'not-used')
})

# SPC-aware version of subset()
y <- subset(x, !flag)

# check: ok
table(y$genhz)


par(mar = c(0, 0, 3, 0))
plotSPC(y, color = 'genhz', width = 0.3, name.style = 'center-center', plot.depth.axis = FALSE, hz.depths = TRUE, cex.names = 0.65, col.label = 'Generalized Horizon Label', print.id = FALSE)
```


Add empty horizons to bottom of the profiles with bottom-most horizon depth < 150cm, label with 'R' GHL. Truncate deeper profiles to 150cm. See `?fillHzGaps` and `?trunc` for details.
```{r, fig.width=12, fig.height=6}
# extend with empty horizons
z <- fillHzGaps(y, to_top = NA, to_bottom = 150)

# label as 'R'
z$hzname[is.na(z$genhz)] <- 'R'
z$genhz[is.na(z$genhz)] <- 'R'

# truncate all profiles to 150cm
z <- trunc(z, z1 = 0, z2 = 150)

par(mar = c(0, 0, 3, 0))
plotSPC(z, color = 'genhz', width = 0.3, name.style = 'center-center', plot.depth.axis = FALSE, hz.depths = TRUE, cex.names = 0.65, col.label = 'Generalized Horizon Label', print.id = FALSE)
```


Arrange profiles by depth to contact using `plot.order` argument to `plotSPC`. See `?getSoilDepthClass` and `?estimateSoilDepth` for details on estimating soil depth.
```{r, fig.width=12, fig.height=6}
# add soil depth information to site
sdc <- getSoilDepthClass(z)
site(z) <- sdc

par(mar = c(0, 0, 3, 0))
plotSPC(z, color = 'genhz', width = 0.3, name.style = 'center-center', plot.depth.axis = FALSE, hz.depths = TRUE, cex.names = 0.65, plot.order = order(z$depth), col.label = 'Generalized Horizon Label', print.id = FALSE)
```


Compute pair-wise distances between profiles using `genhz` (an ordered factor) to a maximum depth of 150cm. See `?profile_compare` and `?daisy` for details. Divisive hierarchical clustering is applied to the distance matrix, with a final re-ordering step according to soil depth (`dendextend::rotate)`.
```{r}
# note funky syntax required to trick profile_compare... will be fixed soon
d <- profile_compare(z, vars = c('genhz', 'genhz'), max_d = 150, k = 0, rescale.result = TRUE)

# divisive hierarchical clustering
h <- as.hclust(diana(d))

# attempt sorting dendrogram by soil depth
h <- dendextend::rotate(h, order(z$depth))
```


`plotProfileDendrogram` from the sharpshootR package aligns profile sketches with the terminal leaves of a dendrogram.
```{r, fig.width=12, fig.height=8}
par(mar = c(0, 0, 3, 0))
plotProfileDendrogram(z, clust = h, scaling.factor = 0.0095, y.offset = 0.05, color = 'genhz', width = 0.3, name.style = 'center-center', col.label = 'Generalized Horizon Label', plot.depth.axis = FALSE, hz.depths = TRUE, cex.names = 0.65, print.id = FALSE)
```


Use distance from 1st profile as a virtual transect via `alignTransect`, `fixOverlap` attempts to spread-out profiles along the x-axis to reduce overlap. 
```{r, fig.width=12, fig.height=6}
m <- as.matrix(d)

set.seed(10101)
pos <- alignTransect(m[1, ], x.min = 1, x.max = length(z), thresh = 0.5)
```

Distance from 1st profile (left-most, marked with a red **asterisk**) used to order collection along an integer sequence.
```{r, fig.width=12, fig.height=6}
par(mar = c(4, 0, 3, 0))

plotSPC(z, plot.order = pos$order, color = 'genhz', width = 0.25, name.style = 'center-center', plot.depth.axis = FALSE, hz.depths = TRUE, cex.names = 0.65, col.label = 'Generalized Horizon Label', print.id = FALSE)

points(x = 1, y = -5, pch = '*', cex = 3, col = 'firebrick')

axis(1, at = 1:length(z), labels = round(pos$grad, 3), line = 0, cex.axis = 0.85)
mtext("Distance from 1st Profile (not to scale)", side = 1, line = 2.25)
```

Distance from 1st profile (left-most, marked with a red **asterisk**) used to place profiles at relative distances along the x-axis.
```{r, fig.width=12, fig.height=6}
par(mar = c(4, 0, 3, 1))

plotSPC(z, plot.order = pos$order, relative.pos = pos$relative.pos, color = 'genhz', width = 0.2, name.style = 'center-center', name = NA, col.label = 'Generalized Horizon Label', print.id = FALSE)

points(x = 1, y = -5, pch = '*', cex = 3, col = 'firebrick')

axis(1, at = pos$relative.pos, labels = round(pos$grad, 3), line = 0, cex.axis = 0.66)
mtext("Distance from 1st Profile", side = 1, line = 2.25)
```

Create an ordination from pair-wise distances using non-metric multidimensional scaling (`MASS::sammon`).
```{r fig.width=8, fig.height=4.5}
# add tiny amount for 0-distance pairs (duplicate profiles)
mds <- sammon(d + 0.001)$points

# ensure ordering is preserved:
# profile IDs -> distance matrix -> nMDS
all(row.names(mds) == profile_id(z))
```

Re-scale nMDS coordinates to typical coordinate system used by `plotSPC`.
```{r fig.width=6, fig.height=5}
# re-scale nMDS axis 1 to the typical horizontal scale used by plotSPC 
xoff <- aqp:::.rescaleRange(mds[, 1], x0 = 1, x1 = length(z))

# re-scale nMDS axis 2 to the typical vertical scale used by plotSPC  
yoff <- aqp:::.rescaleRange(mds[, 2], x0 = -10, x1 = max(z))
```

Place ("hang") soil profile sketches according to nMDS ordination (points). Overlap is a problem.
```{r, fig.width=8, fig.height=6}
par(mar = c(0.25, 0.25, 3.5, 0.25))
plotSPC(z, y.offset = yoff, relative.pos = xoff, color = 'genhz', width = 0.25, name.style = 'center-center', hz.depths = FALSE, name = NA, col.label = 'Generalized Horizon Label', print.id = FALSE, scaling.factor = 0.75)

points(xoff, yoff, pch = 16)

grid(nx = 10, ny = 10)
abline(v = mean(pos$relative.pos), h = mean(yoff), lty = 2)

box()
```

Demonstrate the effect of `alignTransect` and `fixOverlap` on relative postion along nMDS axis 1.
```{r fig.width=6, fig.height=5}
# use re-scaled nMDS coordinates as virtual transect
# adjust to reduce overlap
# with an expansion of the x-axis out to length(z) + 5
set.seed(10110)
pos <- alignTransect(xoff, x.min = 1, x.max = length(z) + 5, thresh = 0.65)

par(mar = c(3, 3, 1, 1))
plot(cbind(pos$relative.pos, yoff[pos$order]), pch = 1, las = 1, axes = FALSE, ylab = '', xlab = '')
grid(nx= 20, ny = 20)
mtext('nMDS axis 1', side = 1, line = 1)
mtext('nMDS axis 2', side = 2, line = 1)

points(cbind(pos$grad, yoff[pos$order]), pch = 16)

# arrows connect original, scaled nMDS axis 1 -> adjusted-for-overlap nMDS axis 1
arrows(
  x1 = pos$relative.pos, y1 = yoff[pos$order],
  x0 = pos$grad, y0 = yoff[pos$order],
  len = 0.05  
)

box()
```

Place soil profile sketches according to approximate positions within nMDS ordination (points). Note the use of `n` argument to `plotSPC`, adding extra "room" on the x-axis. A very large output device is required for legible text.
```{r, fig.width=8, fig.height=6}
par(mar = c(0.25, 0.25, 3.5, 0.25))
plotSPC(z, n = length(z) + 5, y.offset = yoff[pos$order], plot.order = pos$order, relative.pos = pos$relative.pos, color = 'genhz', width = 0.25, name.style = 'center-center', cex.names = 0.45, hz.depths = TRUE, col.label = 'Generalized Horizon Label', print.id = TRUE, scaling.factor = 0.75)

# add adjusted nMDS points
points(pos$relative.pos, yoff[pos$order], pch = 16, cex = 0.75)

box()
```

Another approach, leaving out all of the text annotation. Morphologic patterns and their similarity are clear.
```{r, fig.width=8, fig.height=6}
par(mar = c(0.25, 0.25, 3.5, 0.25))
plotSPC(z, n = length(z) + 5, y.offset = yoff[pos$order], plot.order = pos$order, relative.pos = pos$relative.pos, color = 'genhz', width = 0.25, name.style = 'center-center', hz.depths = FALSE, name = NA, col.label = 'Generalized Horizon Label', print.id = FALSE, scaling.factor = 0.75)

grid(nx = 10, ny = 10)
abline(v = mean(pos$relative.pos), h = mean(yoff), lty = 2)

box()
```


----------------------------
This document is based on `aqp` version `r utils::packageDescription("aqp", field="Version")`.


