---
output:
  html_vignette:
    mathjax: null
    jquery: null
    smart: no
---

```{r setup, echo=FALSE, results='hide', warning=FALSE}
library(knitr, quietly = TRUE)
library(printr, quietly = TRUE)
opts_chunk$set(message=FALSE, warning=FALSE, out.width="100%", fig.retina = 2, fig.align='center', dev='png', dev.args=list(pointsize=10, antialias='cleartype'), tidy=TRUE)
options(width=100, stringsAsFactors=FALSE)

knit_hooks$set(htmlcap = function(before, options, envir) {
  if(!before) {
    paste('<p class="caption" style="font-size:85%; font-style: italic; font-weight: bold;">',options$htmlcap,"</p><hr>",sep="")
    }
    })
```


Horizon Sequence Transition Probability
=======================================
D.E. Beaudette
<br>
`r format(Sys.time(), "%Y-%m-%d")`
<br>
This document is based on `aqp` version `r utils::packageDescription("aqp", field="Version")` and `sharpshootR` version `r utils::packageDescription("sharpshootR", field="Version")`.


# Introduction



# Setup R Envionment
If you have never used the [aqp](https://r-forge.r-project.org/scm/viewvc.php/*checkout*/docs/aqp/aqp-intro.html?root=aqp) or [soildb](https://r-forge.r-project.org/scm/viewvc.php/*checkout*/docs/soilDB/soilDB-Intro.html?root=aqp) packages before, you will likely need to install them. This only needs to be done once. 
```{r install-packages, eval=FALSE}
# stable version from CRAN + dependencies
install.packages('aqp', dep=TRUE) 
install.packages('soilDB', dep=TRUE)
install.packages('sharpshootR', dep=TRUE)
install.packages('markovchain', dep=TRUE)
# latest versions from R-Forge:
install.packages('aqp', repos="http://R-Forge.R-project.org", type='source')
install.packages('soilDB', repos="http://R-Forge.R-project.org", type='source')
install.packages('sharpshootR', repos="http://R-Forge.R-project.org", type='source')
```

Once you have all of the R packages on which  this document depends, it is a good idea to load them. R packages must be **installed** anytime you change versions of R (e.g., after an upgrade) and **loaded** anytime you want to access functions from within those packages.

```{r load-packages}
library(markovchain)
library(aqp)
library(sharpshootR)
library(soilDB)
library(igraph)
```

# Sample Data
While the methods outlined in this document can be applied to any collection of pedons, it is convenient to work with a standardized set of data. You can follow along with the analysis by copying code from the following blocks and running it in your **R** session. The sample data used in this document is based on soil profiles that have been correlated to the [Loafercreek](https://soilseries.sc.egov.usda.gov/OSD_Docs/L/LOAFERCREEK.html) soil series from the Sierra Nevada Foothill Region of California. Note that the internal structure of the `loafercreek` data is identical to the structure returned by [`fetchNASIS()` from the soilDB package](https://r-forge.r-project.org/scm/viewvc.php/*checkout*/docs/soilDB/soilDB-Intro.html?root=aqp). All horizon-level values are pulled from the pedon horizon table of the pedons being analyzed.

```{r load-data, fig.width=9, fig.height=4, htmlcap='Pedons correlated to the Loafercreek soil series.'}
# load sample data from the soilDB package
data(loafercreek, package = 'soilDB')
pedons <- loafercreek
# plot profile sketches
par(mar=c(0,0,0,0))
plot(pedons, name='', print.id=FALSE, cex.names=0.75, axis.line.offset=-4)
```


## Optional: Follow Along with Your Data
The following code block demonstrates how to pull data in using the [`fetchNASIS()` function from the soilDB package](https://r-forge.r-project.org/scm/viewvc.php/*checkout*/docs/soilDB/soilDB-Intro.html?root=aqp).

```{r use-nasis-data, eval=FALSE}
# first load the desired data set within NASIS into your NASIS selected set
# then load data from the NASIS selected set into R
# note that the `nullFragsAreZero` argument converts NULL rock fragment class data into 0s
pedons <- fetchNASIS(nullFragsAreZero=TRUE)
# optionally subset the data by taxon name - enter your taxon name
pedons <- pedons[grep(pattern='ENTER_YOUR_TAXON_NAME', pedons$taxonname, ignore.case=TRUE), ]
```

# Methods

```{r}
sort(table(pedons$hzname), decreasing=TRUE)
```

Constructing a graph of transitions from one horizon to the next ("transition probabilities") can highlight those horizon designations that are most commonly used together. A follow-up tutorial on transition probability matrix interpretation is planned.

```{r transition-probabilities, fig.width=10, fig.height=6, htmlcap='Graph constructed from transition probabilities. Thicker lines denote more likely transitions.'}
# sample data
data(sp4)
depths(sp4) <- id ~ top + bottom

# horizon transition probabilities: row -> col transitions
(tp <- hzTransitionProbabilities(sp4, 'name'))

# abuse sharpshootR code to display as graph
par(mar=c(0,0,0,0), mfcol=c(1,2))
plot(sp4)
plotSoilRelationGraph(tp, graph.mode = 'directed', edge.arrow.size=0.5, edge.scaling.factor=2, vertex.label.cex=0.75, vertex.label.family='sans')
```

More examples.
```{r, fig.width=10, fig.height=6}
# try some other examples, seems logical
tp <- hzTransitionProbabilities(pedons, 'hzname')
# tp <- hzTransitionProbabilities(pedons, 'genhz')

# abuse sharpshootR code
# sometimes default layout doesn't look so good
par(mar=c(0,0,0,0), mfcol=c(1,2))
plotSoilRelationGraph(tp, graph.mode = 'directed', edge.arrow.size=0.5)
plotSoilRelationGraph(tp, graph.mode = 'directed', edge.arrow.size=0.5, g.layout = layout_with_lgl)
```

Convert contingency table of generalized horizon labels to an adjacency (similar to transition probability) matrix.
```{r, fig.width=10, fig.height=6}
tab <- table(pedons$hzname, pedons$genhz)
m <- genhzTableToAdjMat(tab)

par(mar=c(0,0,0,0), mfcol=c(1,1))
plotSoilRelationGraph(m, graph.mode = 'directed', edge.arrow.size=0.5)
```


Tinker with the [markovchain](https://cran.rstudio.com/web/packages/markovchain/index.html) package. Details [here](https://cran.rstudio.com/web/packages/markovchain/vignettes/an_introduction_to_markovchain_package.pdf).
```{r, fig.width=10, fig.height=6}
## init a markovchain object from TP
# this trquires "looping" terminal states
# e.g. an absorbing MC: https://en.wikipedia.org/wiki/Absorbing_Markov_chain
tp.loops <- hzTransitionProbabilities(sp4, 'name', loopTerminalStates = TRUE)

mc <- new("markovchain", states=dimnames(tp.loops)[[1]], transitionMatrix = tp.loops)

# simple plot
plot(mc, edge.arrow.size=0.5)

# check absorbing states
absorbingStates(mc)

# steady-state:
steadyStates(mc)
```

```{r, fig.width=10, fig.height=6}
# try some more complex TP matrices
tp <- hzTransitionProbabilities(pedons, 'hzname', loopTerminalStates = TRUE)
mc <- new("markovchain", states=dimnames(tp)[[1]], transitionMatrix = tp)

# better plotting
par(mar=c(0,0,0,0), mfcol=c(1,2))
plot(mc, vertex.size=10, edge.arrow.size=0.5, edge.label.cex=0.75, layout=layout_with_lgl)
plot(mc, vertex.size=10, edge.arrow.size=0.5, edge.label.cex=0.75, layout=layout_as_tree)

# another 
tp <- hzTransitionProbabilities(pedons, 'genhz', loopTerminalStates = TRUE)
mc <- new("markovchain", states=dimnames(tp)[[1]], transitionMatrix = tp)

# better plotting
par(mar=c(0,0,0,0), mfcol=c(1,2))
plot(mc, vertex.size=10, edge.arrow.size=0.5, edge.label.cex=0.75, layout=layout_as_tree)
plot(mc, vertex.size=10, edge.arrow.size=0.5, edge.label.cex=0.75, layout=layout_with_lgl)
```


Simulation from a markovchain object.
```{r, fig.width=10, fig.height=6}
# simulate
rmarkovchain(10, mc, include.t0=TRUE, t0='A')

# multiple simulations, as column vectors
replicate(10, rmarkovchain(10, mc, include.t0=TRUE, t0='A'))

# what comes after a state?
conditionalDistribution(mc, 'A')

# this is now in AQP, but requires markovchain package
mostLikelyHzSequence(mc, 'A')
```


Try with some KSSL data.
```{r, fig.width=10, fig.height=10}
# some series to query
s <- c('amador', 'auburn', 'musick', 'holland')
s.data <- lapply(s, fetchKSSL)

par(mar=c(0,0,2,0), mfcol=c(2,2))
for(i in 1:length(s)) {
  tp <- hzTransitionProbabilities(s.data[[i]], 'hzn_desgn')
  plotSoilRelationGraph(tp, graph.mode = 'directed', edge.arrow.size=0.3, edge.scaling.factor=2, main=s[i], vertex.label.cex=0.75, vertex.label.family='sans')
}
```

