---
title: "Investigating Soil Color"
author: "D.E. Beaudette"
date: "`r Sys.Date()`"
output:
  html_document:
    mathjax: null
    jquery: null
    smart: no
    number_sections: no
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: false
---
  
```{r setup, echo=FALSE, results='hide', warning=FALSE}
library(knitr)
opts_chunk$set(message=FALSE, warning=FALSE, background='#F7F7F7', dev='svglite', tidy=FALSE, verbose=FALSE, cache = FALSE)
options(width=100, stringsAsFactors=FALSE, cache=FALSE)
```


```{r}
library(aqp)
library(soilDB)
# library(plyr)
library(sharpshootR)
library(latticeExtra)
library(colorspace)
library(ggplot2)
library(treemapify)
```


# Topics

  * photos
  * box photos
  * aqp sketches via `plotSPC()`
  * Munsell color spectra
  * [`colorscience` package](https://cran.r-project.org/web/packages/colorscience/colorscience.pdf)
  * `previewColors()`
  * `colorQuantiles()`
  * binning of soil profile photos via `colordist` package
  * `soilColorSignature()`
  * color mixing `mix_and_clean_colors()`
  * color RIC vis color book presentation
  * colors in context with common soil pigments
  * colors in context with full CIELAB space
  * examples of exotic colors: sulphides, manganese, etc.
  * poorly drained soils, reduced iron color ranges, gley pages
  * color by drainage class


[Clarksville](https://casoilresource.lawr.ucdavis.edu/sde/?series=clarksville).
```{r}
# get lab / morphologic data
# simplify colors
x <- fetchKSSL(series='clarksville', returnMorphologicData = TRUE, simplifyColors = TRUE)

# extract pedons into SoilProfileCollection
s <- x$SPC

# genhz
s$genhz <- generalize.hz(
  x = s$hzn_desgn, 
  new = c('A', 'E', 'Bt', '2Bt', '3Bt'), 
  pattern = c('A', 'E', '^Bt', '2B', '3B'),
  non.matching.code = NA
)

s$genhz <- factor(s$genhz, levels = guessGenHzLevels(s, "genhz")$levels)

table(s$genhz, useNA = 'always')
```


# Previewing Colors

MDS representation of unique colors, based on CIE2000 distances.
```{r fig.width=5, fig.height=5, fig.align='center'}
previewColors(s$moist_soil_color, method = 'MDS', pt.cex = 1.5)
title('Clarksville Soil Colors')
```

Arrange colors in a grid, based on CIE2000 distances.
```{r fig.width=5, fig.height=5, fig.align='center'}
previewColors(s$moist_soil_color)
title('Clarksville Soil Colors')
```

White borders.
```{r fig.width=5, fig.height=5, fig.align='center'}
previewColors(s$moist_soil_color, border.col = 'white')
title('Clarksville Soil Colors')
```

No borders.
```{r fig.width=5, fig.height=5, fig.align='center'}
previewColors(s$moist_soil_color, border.col = NA)
title('Clarksville Soil Colors')
```

Order colors by horizon top depth.
```{r fig.width=5, fig.height=5, fig.align='center'}
no.na.idx <- which(!is.na(s$moist_soil_color))
previewColors(s$moist_soil_color[no.na.idx], method = 'manual', col.order = order(s$hzn_top[no.na.idx]), border.col = NA)
title('Clarksville Soil Colors')
```

Order colors by generalized horizon.
```{r fig.width=5, fig.height=5, fig.align='center'}
no.na.idx <- which(!is.na(s$moist_soil_color))
previewColors(s$moist_soil_color[no.na.idx], method = 'manual', col.order = order(s$genhz[no.na.idx]), border.col = NA)
title('Clarksville Soil Colors')
```


# Color Book Style Presentation
Color book style range for colors, frequency is represented as chip size.
```{r fig.width=9, fig.height=8.5}
# prepare vectors of Munsell chips + groups (generalized horizon labels)
m <- paste0(s$m_hue, ' ', s$m_value, '/', s$m_chroma)
g <- s$genhz

colorChart(m, g = g, chip.cex = 2)
```

Constant chip size, annotate with chip frequency.
```{r fig.width=9, fig.height=8.5}
colorChart(m, g = g, chip.cex = 2, size = FALSE, annotate = TRUE)
```



```{r eval=FALSE, echo = FALSE}
# remove horizons that are missing moist colors
h <- horizons(s)
idx <- which(complete.cases(h[, c('m_hue', 'm_value', 'm_chroma', 'genhz')]))
h <- h[idx, ]

# 
h$m_munsell <- paste0(h$m_hue, ' ', h$m_value, '/', h$m_chroma)
h$m_hue <- factor(h$m_hue, levels = c('10Y', '2.5Y', '10YR', '7.5YR', '5YR', '2.5YR'))
h$m_hue <- factor(h$m_hue)

xyplot(m_value ~ m_chroma | m_hue, as.table=TRUE, data=h, subscripts = TRUE, xlim=c(0.5,8.5), ylim=c(0.5,8.5), scales=list(alternating=3, tick.number=8, y=list(rot=0)), xlab='Chroma', ylab='Value', layout=c(length(levels(h$m_hue)),1), strip=strip.custom(bg=grey(0.85)), panel=function(x, y, subscripts=subscripts, ...) {
  
  p.data <- data.frame(x=x, y=y, col=h$moist_soil_color[subscripts], m=h$m_munsell[subscripts], stringsAsFactors = FALSE)
  tab <- prop.table(table(p.data$m, useNA = 'always'))
  tab <- as.data.frame(tab)
  names(tab) <- c('m', 'freq')
  p.data <- merge(p.data, tab, by = 'm', all.x = TRUE, sort = FALSE)
  p.data <- na.omit(p.data)
  p.data <- subset(p.data, subset=freq > 0.05)
  panel.grid(-1, -1)
  panel.xyplot(p.data$x, p.data$y, pch = 22, col = 'black', fill = p.data$col, cex = 4 * sqrt(p.data$freq))
})


pp <- xyplot(m_value ~ m_chroma | m_hue + genhz, as.table=TRUE, data=h, subscripts = TRUE, xlim=c(0.5,8.5), ylim=c(0.5,8.5), scales=list(alternating=3, tick.number=8, y=list(rot=0)), xlab='Chroma', ylab='Value', subset=genhz != 'not-used', panel=function(x, y, subscripts=subscripts, ...) {
  
  p.data <- data.frame(x=x, y=y, col=h$moist_soil_color[subscripts], m=h$m_munsell[subscripts], stringsAsFactors = FALSE)
  tab <- prop.table(table(p.data$m, useNA = 'always'))
  tab <- as.data.frame(tab)
  names(tab) <- c('m', 'freq')
  p.data <- merge(p.data, tab, by = 'm', all.x = TRUE, sort = FALSE)
  p.data <- na.omit(p.data)
  p.data <- subset(p.data, subset=freq > 0.05)
  panel.grid(-1, -1)
  panel.xyplot(p.data$x, p.data$y, pch=22, col='black', fill=p.data$col, cex=4 * sqrt(p.data$freq))
})

useOuterStrips(pp, strip=strip.custom(bg=grey(0.85)), strip.left = strip.custom(bg=grey(0.85)))
```



# Soil Color Aggregation 
```{r}
a <- aggregateColor(s, "genhz", col = 'moist_soil_color')
a.reduced <- aggregateColor(s, "genhz", col = 'moist_soil_color', k = 8)

par(mar = c(4.5, 2.5, 4.5, 0))
aggregateColorPlot(a, label.cex = 0.65, main = "Clarksville Moist Colors\nGeneralized Horizons", print.n.hz = FALSE, print.label = FALSE, rect.border = NA, horizontal.borders = TRUE)

par(mar = c(4.5, 2.5, 4.5, 0))
aggregateColorPlot(a.reduced, label.cex = 0.65, main = "Clarksville Moist Colors\nGeneralized Horizons\n8 Colors per Group", print.n.hz = TRUE)
```

# Soil Color RIC via treemap
```{r fig.width=8, fig.height=6.5}
a <- do.call('rbind', a.reduced$scaled.data)
a.unique <- unique(a[, c('munsell', 'moist_soil_color')])
a.cols <- a.unique$moist_soil_color
names(a.cols) <- a.unique$munsell

ggplot(data = a) + 
  geom_treemap(aes(area = weight, fill = munsell)) +
  geom_treemap_text(aes(area = weight, label = munsell, colour = I(invertLabelColor(moist_soil_color)))) +
  facet_wrap(~ .id, strip.position = "top") + 
  scale_fill_manual(
    guide = 'none',
    values = a.cols
  ) +
  coord_equal() +
  theme_bw() + 
  labs(title = 'Clarksville Moist Color RIC, KSSL Pedons') + 
  theme(
    strip.background = element_blank(),
    strip.text = element_text(face = "bold"),
    panel.border = element_rect(fill = NA, colour = NA),
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.title = element_blank()
  )
```



# Soil Color Simulation
```{r fig.width = 8.5, fig.height = 4}
# simulation parameters
p <- list(
  list(m = '7.5YR 4/6', thresh = 12, hues = c('5YR', '7.5YR', '10YR'))
)

# simulation
sim <- simulateColor(method = 'dE00', n = 100, parameters = p)

# present via color chart / tabulation
pp <- colorChart(sim[[1]], annotate = FALSE, chip.cex = 3)
update(pp, asp = 1, main = 'RV Color: 7.5YR 4/6\ndE00 threshold < 12')
```


```{r fig.width = 8.5, fig.height = 4}
# extract horizons, to generate a data.frame of moist colors
h <- horizons(s)

# remove horizons that are missing moist colors
idx <- which(complete.cases(h[, c('m_hue', 'm_value', 'm_chroma', 'genhz')]))
h <- h[idx, ]

# simulation parameters
p <- list(
  list(hvc = data.frame(hue = h$m_hue, value = h$m_value, chroma = h$m_chroma))
)

# simulation
sim <- simulateColor(method = 'mvnorm', n = 100, parameters = p)

# present via color chart / tabulation
pp <- colorChart(sim[[1]], annotate = FALSE, chip.cex = 3)
update(pp, main = 'Multivariate Simulation (All Colors)')
```



# Color Quantiles
```{r fig.width=7, fig.height=7}
x <- colorQuantiles(na.omit(s$moist_soil_color))

# result from plotColorQuantiles() is a lattice graphics object
update(
  plotColorQuantiles(x),
  main = 'All Colors'
)
```


Compute color quantiles by generalized horizon label.
```{r fig.width=7, fig.height=7}
h.by.genhz <- split(s$moist_soil_color, f = s$genhz)

l <- lapply(h.by.genhz, function(i) {
  colorQuantiles(na.omit(i))
})

# print figures to current graphics device (multiple pages)
for(i in names(l)) {
  # make figure
  fig <- plotColorQuantiles(l[[i]])
  # format title
  main.txt <- sprintf("Generalized Horizon: %s", i)
  # add figure title
  fig <- update(fig, main = main.txt)
  print(fig)
}
```


Reduce lists to `data.frame` objects for further use or reporting.
```{r}
L1.data <- do.call('rbind', lapply(l, '[[', 'L1'))
m.data <- do.call('rbind', lapply(l, '[[', 'marginal'))
```


# Soil Color Palettes
Examples for creating a soil-inspired palette of colors.
```{r fig.width=7, fig.height=1.5}
# aggregate colors over all depths into 6 medoid colors
site(s)$grp <- 'a'
a <- aggregateColor(s, group = "grp", col = 'moist_soil_color', k = 6)

# setup margins
par(mar=c(0.25,0.25,0.25,0.25), bg='white')

# using L1 medians, in depth-order
soilPalette(L1.data$L1_color, L1.data$L1_chip)

# using 6 medoid colors
soilPalette(a$scaled.data$a$moist_soil_color, a$scaled.data$a$munsell)
```

```{r fig.width=7, fig.height=4}
L1.cols <- L1.data$L1_color
medoid.cols <- a$scaled.data$a$moist_soil_color

swatchplot(
  list(
    "L1 Median Colors" = L1.cols,
    "Medoid Colors" = medoid.cols)
)
```

```{r fig.width=7, fig.height=6}
swatchplot(
  list(
    "L1 Median Colors" = L1.cols,
    "Medoid Colors" = medoid.cols),
  cvd = TRUE
)
```


----------------------------
This document is based on `aqp` version `r utils::packageDescription("aqp", field="Version")`, `soilDB` version `r utils::packageDescription("soilDB", field="Version")`, and `sharpshootR` version `r utils::packageDescription("sharpshootR", field="Version")`.


