---
title: "L1 Profiles"
author: "D.E. Beaudette"
date: "`r Sys.Date()`"
output:
  html_document:
    mathjax: null
    jquery: null
    smart: no
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: false
---
  
```{r setup, echo=FALSE, results='hide', warning=FALSE}
# setup
library(knitr, quietly=TRUE)
opts_chunk$set(message=FALSE, warning=FALSE, background='#F7F7F7', fig.align='center', fig.retina=1, dev='svglite', tidy=FALSE, verbose=FALSE)
options(width=100, stringsAsFactors=FALSE)
```

# Introduction
The concept of an "L1 profile" is based on the [multivariate median](https://en.wikipedia.org/wiki/Median#Multivariate_median), also referred to as the "geometric median" or "L1 median". The L1 median represents an n-dimensional point that minimizes some distance metric to all other points in a data set. Consider the following 2-dimensional data (black points), estimated marginal medians (blue), and L1 median (red). The marginal medians are computed independently along each dimension which the L1 median is computed across all dimensions.

```{r fig.width=6, fig.height=6, echo=FALSE}
library(Gmedian)

set.seed(4)
x <- rnorm(100, mean = 1, sd = 0.25)
y <- x * rnorm(100, mean = 1, sd = 2) + x

# marginal medians
med.x <- median(x)
med.y <- median(y)

# L1 median
G <- Gmedian(cbind(x, y))

par(mar = c(1, 1, 1, 1))
plot(x, y, axes = FALSE, xlab = '', ylab = '', type = 'n')
segments(x0 = G[, 1], y0 = G[, 2], x1 = x, y1 = y, col = rgb(0.8, 0.8, 0.8, alpha = 0.5))
abline(v = med.x, h = med.y, lty = 3)
points(med.x, med.y, pch = 15, col = 'royalblue', cex = 2)
points(G, pch = 15, col = 'firebrick', cex = 2)
points(x, y, pch = 16, cex = 0.85)

legend('topleft', legend = c('Marginal Median', 'L1 Median', 'Data'), pch = c(15, 15, 16), pt.cex = c(2, 2, 0.85), col = c('royalblue', 'firebrick', 'black'), bty = 'n', cex = 0.8)
box()
```


The "L1 profile" is a synthetic soil profile built from L1 median values along 1cm depth intervals. Consider 10 simulated profiles, each with 3 soil properties, all sharing a relatively common distribution with depth.
```{r fig.width=8, fig.height=6, echo = FALSE}
library(aqp)
library(sharpshootR)

set.seed(1010)
x <- lapply(sprintf("Sim %01d", 1:10), random_profile, n = c(5:8), n_prop = 3, method = 'LPP', SPC = TRUE)
x <- combine(x)
site(x)$group <- 'L1 Median'

x <- trunc(x, 0, 100)

# aggregation formula
fm <- group ~ p1 + p2 + p3

# L1 median by group
z <- L1_profiles(x, fm = fm, method = 'constant', maxDepthConstant = 100)

# combine L1 + source data
s <- combine(x, z)

# compute PCA
pc <- princomp(horizons(s)[, c('p1', 'p2', 'p3')], cor = TRUE)
pc.scores <- predict(pc, horizons(s)[, c('p1', 'p2', 'p3')])

# save to SPC
s$pc1 <- pc.scores[, 1]

## mixture of color is not intuitive
## consider using actual soil color
# s$pc_color <- rgb(scales::rescale(pc.scores, to = c(0, 1)))

# num classification
d <- NCSP(s, vars = c('p1', 'p2', 'p3'), maxDepth = 100, rescaleResult = TRUE)

# simple clustering for viz
h <- hclust(d, method = 'ward.D2')

# viz
par(mar = c(0, 0, 3, 1.5))
plotProfileDendrogram(s, clust = h, color = 'pc1', y.offset = 0.2, scaling.factor = 0.018, width = 0.3, name = NA, depth.axis = list(cex = 0.85), divide.hz = FALSE)
```




# Demonstration
Demonstrate with real data.
```{r fig.width = 8, fig.height=6}
library(aqp)
library(soilDB)
library(sharpshootR)
library(cluster)
library(reshape2)
library(latticeExtra)
library(Gmedian)
library(MASS)
library(tactile)

# get some interesting data
x <- fetchKSSL('plainfield')

# remove partial-profile records
x <- subset(x, pedon_completeness_index > 3)
x <- subset(x, pedon_key != '6678')

# this is rather drastic, consider other possible "repairs"
# fillHzGaps(x), accumulateDepths(x), repairMissingHzDepths(x)
# remove problematic horizonation
x <- HzDepthLogicSubset(x)

# normalize taxonname for use as a grouping variable
x$taxonname <- tolower(x$taxonname)

# compute weighted-mean particle diameter (mm) for later
x$wmpd <- with(horizons(x), ((vcs * 1.5) + (cs * 0.75) + (ms * 0.375) + (fs * 0.175) + (vfs *0.075) + (silt * 0.026) + (clay * 0.0015)) / (vcs + cs + ms + fs + vfs + silt + clay))
```

# Aggregation
```{r fig.width = 8, fig.height=6} 
site(x)$group <- 'L1 Median'

# aggregation formula
fm <- group ~ sand + silt + clay + estimated_ph_h2o + bs82 + wmpd

# L1 medians
z <- L1_profiles(x, fm = fm, method = 'constant', maxDepthConstant = 150)

# marginal medians
a <- slab(x, fm = fm)
```

## Sanity Check
```{r fig.width = 8, fig.height=6}
# check that sand + silt + clay sum to 100
(z$ssc_total <- z$sand + z$silt + z$clay)
```

## Merge Marginal + L1
```{r fig.width = 8, fig.height=6}
## TODO: this involves a lot of dumb re-naming
vars <- c("sand", "silt", "clay", "estimated_ph_h2o", "bs82", "wmpd")
z.long <- reshape2::melt(as(z, 'data.frame'), id.vars = c(idname(z), horizonDepths(z)), measure.vars = vars)

new.vars <- c('variable', 'group', 'top', 'bottom', 'p.q50', 'p.q25', 'p.q75', 'contributing_fraction')

# re-name and adjust to match slab output
names(z.long)[1] <- 'group'
names(z.long)[5] <- 'p.q50'
z.long$p.q25 <- NA
z.long$p.q75 <- NA
# not sure how we can best utilize this
z.long$contributing_fraction <- 0
z.long <- z.long[, new.vars]

g <- make.groups(
  marginal = a[, new.vars],
  L1 = z.long
)
```

## Graphical Comparison
```{r fig.width = 12, fig.height=6}
# plotting style
tps <- tactile.theme(superpose.line = list(col = c('RoyalBlue', 'DarkRed', 'DarkGreen'), lwd = 2))

# fancy panel names
levels(g$variable) <- c('Sand (%)', 'Silt (%)', 'Clay (%)', 'pH 1:1 H2O', 'BS at pH 8.2 (%)', 'WMPD (mm)')

# plot grouped, aggregate data
xyplot(top ~ p.q50 | variable, groups = which, data=g, ylab='Depth',
       main = 'Plainfield Lab Data',
       xlab='median bounded by 25th and 75th percentiles',
       lower = g$p.q25, upper = g$p.q75, ylim = c(155,-5),
       panel = panel.depth_function, alpha=0.25, sync.colors=TRUE,
       prepanel = prepanel.depth_function,
       # cf = g$contributing_fraction,
       par.strip.text=list(cex=0.8),
       strip=strip.custom(bg=grey(0.85)),
       layout=c(6,1), scales=list(x=list(alternating=1, relation='free'), y=list(alternating=3)),
       auto.key = list(lines = TRUE, points = FALSE, columns = 2),
       par.settings = tps
)
```

# Pair-Wise Distances
```{r fig.width = 12, fig.height=6}
# merge SPCs: original + L1 profile
x.z <- combine(list(x, z))

# calculate soil texture class
x.z$texture <- ssc_to_texcl(sand = x.z$sand, clay = x.z$clay)
x.z$texture <- factor(x.z$texture, levels = SoilTextureLevels(which = 'codes'))

# truncate the new collection to 150cm
x.z <- trunc(x.z, 0, 150)

# pair-wise eval of distances
d <- NCSP(x.z, vars = vars, maxDepth = 150, k = 0)

# divisive hierarchical clustering
hc <- diana(d)
```

## Hierarchical Clustering
```{r fig.width = 12, fig.height=6}
par(mar = c(0, 0, 3, 1))
plotProfileDendrogram(x.z, clust = hc, scaling.factor = 0.85, y.offset = 5, width = 0.3, divide.hz = FALSE, color = 'texture', name.style = 'center-center', id.style = 'side')

plotProfileDendrogram(x.z, clust = hc, scaling.factor = 0.85, y.offset = 5, width = 0.3, divide.hz = FALSE, color = 'wmpd', name.style = 'center-center', id.style = 'side')

plotProfileDendrogram(x.z, clust = hc, scaling.factor = 0.85, y.offset = 5, width = 0.3, divide.hz = FALSE, color = 'bs82', name.style = 'center-center', id.style = 'side')
```


## nMDS
The L1 profile is located very close to the center of the ordination.
```{r fig.width = 8, fig.height=8}
# non-metric multidimensional scaling of distance matrix
mds <- sammon(d)

# viz
par(mar = c(1, 1, 1, 1))
plot(mds$points, type = 'n', axes = FALSE, xlab = '', ylab = '')
abline(h = 0, v = 0, lty = 2, col = 'grey')
text(mds$points, row.names(mds$points), font = 2, cex = 0.66)
box()
```

----------------------------
This document is based on `aqp` version `r utils::packageDescription("aqp", field="Version")`.


