---
title: "Mixing Munsell Colors in {aqp}"
author: "D.E. Beaudette"
date: "`r Sys.Date()`"
output:
  html_document:
    smart: no
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: false
---
  
```{r setup, echo=FALSE, results='hide', warning=FALSE}
# setup
library(knitr, quietly = TRUE)
opts_chunk$set(message=FALSE, warning=FALSE, background='#F7F7F7', fig.retina=2, fig.align = 'center', dev='png', tidy=FALSE, verbose=FALSE, cache = FALSE)
options(width=100, stringsAsFactors=FALSE, cache=FALSE)
```


# Background

The [Munsell system](https://en.wikipedia.org/wiki/Munsell_color_system) provides a quantitative (numeric) description of color that is approximately linear in terms of average human perception, and, conveniently isolates the independent qualities of color into hue, value, and chroma. Moving from one chip to the next represents a linear traversal through a continuous, 3D (hue, value, chroma) color space. The presentation (soil color book) and notation (e.g. `10YR 3/3`) that we are familiar with is based on a discretization of that space into perceptually-compact units (the color chips) that offers sufficient contrast for field identification without an overwhelming number of possibilities. For example, our ability to perceive contrast in lightness (Munsell value) is roughly 2-4x greater than our ability to perceive contrast in chroma: hence the lack of chips representing `/5`, `/7`, `/9`, etc..

Given that the Munsell system is quantitative and roughly linear in terms of human color perception, it is entirely possible to:

 * treat value and chroma as numeric entities
 * interpolate between chips (hue, value, or chroma)
 * simulate mixtures of two or more colors in Munsell notation

I'd recommend [SSSA special publication #31](https://acsess.onlinelibrary.wiley.com/doi/book/10.2136/sssaspecpub31), any book on color science, or [one](http://www.applepainter.com/) of several [excellent websites](http://www.munsellcolourscienceforpainters.com/). Our definition of ["soil color contrast"](https://www.nrcs.usda.gov/wps/portal/nrcs/detail/soils/ref/?cid=nrcs142p2_053569) is based on the quantitative, linear nature of the Munsell system.


An accurate simulation of pigment mixtures ("subtractive" color mixtures) is incredibly complex due to factors that aren't easily measured or controlled: pigment solubility, pigment particle size distribution, water content, substrate composition, and physical obstruction to name a few. That said, it is possible to simulate reasonable, subtractive color mixtures given a reference spectra library (350-800nm) and some assumptions about pigment qualities and lighting. For the purposes of estimating a mixture of soil colors (these are pigments after all) we can relax these assumptions and assume a standard light source. The only missing piece is the spectral library for all Munsell chips in our color books.

Thankfully, [Scott Burns has outlined the entire process](https://arxiv.org/ftp/arxiv/papers/1710/1710.06364.pdf), and Paul Centore has provided a Munsell color chip [reflectance spectra library](http://www.munsellcolourscienceforpainters.com/). The estimation of a subtractive mixture of soil colors can proceed as follows:

 1. lookup the associated spectra for each color
 2. computed the weighted (area proportion) geometric mean of the spectra
 3. search for the closest matching spectra in the reference library
 4. suggest that Munsell chip as the best candidate for the simulated mixture

Key assumptions include:

 * similar particle size distribution
 * similar mineralogy (i.e. pigmentation qualities)
 * similar water content. 
 
For the purposes of estimating (for example) a "mixed soil color within the top 18cm of soil" these assumptions are usually valid. Again, these are estimates that are ultimately "snapped" to the nearest chip and not do not need to approach the accuracy of paint-matching systems.


So what is a reasonable answer to the following soil color mixture question: <span style="padding: 5px; background-color: `r aqp::parseMunsell('5YR 2/2')`; color: white;">`50% 5YR 2/2`</span> + <span style="padding: 5px; background-color: `r aqp::parseMunsell('10YR 4/4')`; color: white;">`50% 10YR 4/4`</span> = <b>?</b>.


## Munsell Spectral Library
```{r fig.height=5.5, fig.width=9}
library(aqp)
library(lattice)
library(tactile)
library(sharpshootR)
library(gower)
library(venn)

# local copy of the Munsell chip spectral library
# c/o http://www.munsellcolourscienceforpainters.com/
# odd chroma spectra via interpolation
# see ?munsell.spectra for details
# try aqp:::.summarizeMunsellSpectraRanges()
data(munsell.spectra)

# all hues, limit to specific hue / chroma slice
x <- munsell.spectra[munsell.spectra$value  == 6 & munsell.spectra$chroma == 8, ]

# each Munsell chip has a 36-element spectra
# ranging from 380-730 nm
# table(x$munsell)

# spectra IDs
x$ID <- factor(x$munsell)
# create a color / chip
cols <- parseMunsell(as.character(levels(x$ID)))

# plot style
tps <- tactile.theme(superpose.line = list(col = cols, lwd = 2))

# final figure
xyplot(
  reflectance ~ wavelength, groups = ID, data = x, 
  par.settings = tps,
  main = 'Value 6 / Chroma 8',
  type = c('l', 'g'),
  ylab = 'Reflectance',
  xlab = 'Wavelength (nm)',
  scales = list(tick.number = 12),
  xlim = c(370, 740)
)
```

## Implementation
Required packages, be sure to ge the [latest {aqp} from GitHub](https://github.com/ncss-tech/aqp).

```{r}
library(aqp)
library(lattice)
library(tactile)
library(sharpshootR)
library(gower)
library(venn)
```

Create a vector of colors to mix, and evaluate [color contrast metrics](https://ncss-tech.github.io/AQP/aqp/color-contrast.html).
```{r fig.height=3.5, fig.width=6}
colors <- c('5YR 2/2', '10YR 4/4')
colorContrast(m1 = colors[1], m2 = colors[2])
```
We can see that these colors are 2 hue, 2 value, and 2 chroma "chips" apart from eachother. This evaluates to a *prominent* [soil color contrast class](https://www.nrcs.usda.gov/wps/portal/nrcs/detail/soils/ref/?cid=nrcs142p2_053569) and [$\Delta{E_{00}}$](http://zschuessler.github.io/DeltaE/learn/) of about 20.


A picture is far more intuitive.
```{r fig.height=3.5, fig.width=4}
par(bg = 'black', fg = 'white', mar = c(0, 0, 0, 0))
colorContrastPlot(m1 = colors[1], m2 = colors[2])
```

Simulate a mixture via `mixMunsell`. See the [online documentation](http://ncss-tech.github.io/aqp/docs/reference/mixMunsell.html) for details on the method and [relevant citations](http://ncss-tech.github.io/aqp/docs/reference/munsell.spectra.html).
```{r fig.height=2.5, fig.width=5}
# just the closest match
mixed <- mixMunsell(colors)

# top 5 matches, decreasing order
mixed.5 <- mixMunsell(colors, n = 5)
```

Left to right: original colors-to-mix and final mixture candidate (`r mixed$munsell`). This is the Munsell chip associated with the reference spectra that is *closest* to the [weighted geometric mean](https://en.wikipedia.org/wiki/Weighted_geometric_mean) of the source spectra.
```{r fig.height=2.5, fig.width=6}
par(bg = 'black', fg = 'white', mar = c(0, 0, 0, 0))
soilPalette(parseMunsell(c(colors, mixed$munsell)), lab = c(colors, mixed$munsell))
```

Evaluate the top 5 mixture candidates (ranked by spectral distance) using metrics of [color contrast](http://ncss-tech.github.io/aqp/docs/reference/colorContrast.html). Note that color contrast does not always correlate with spectral distance.
```{r fig.height=4, fig.width=8}
par(bg = 'black', fg = 'white')
colorContrastPlot(
  m1 = rep(mixed$munsell, times = nrow(mixed.5)),
  m2 = mixed.5$munsell,
  labels = c('closest match', 'top 5 matches')
  )
```

## Results

Demonstrate reference spectra for the source colors and final mixture. The dashed line is the weighted geometric mean spectra.
```{r fig.height=5.5, fig.width=9}
plotColorMixture(colors, showMixedSpec = TRUE)
```

<span style="padding: 5px; background-color: `r aqp::parseMunsell('5YR 2/2')`; color: white;">`50% 5YR 2/2`</span> + <span style="padding: 5px; background-color: `r aqp::parseMunsell('10YR 4/4')`; color: white;">`50% 10YR 4/4`</span> = <span style="padding: 5px; background-color: `r aqp::parseMunsell(mixed$munsell)`; color: white;">`r mixed$munsell`</span>.

What do you think, is the mixture plausible? Experiments with real soil material + Nix Pro sensor measurements pending.

### Experimenting with Color Mixtures
The `colorMixtureVenn` function from the sharpshootR package is a handy way to view simulated mixtures of 2-7 colors.
```{r fig.height=5, fig.width=5}
colorMixtureVenn(colors)

colorMixtureVenn(c('10YR 8/1', '2.5YR 3/6', '10YR 2/2'))
```





# Additional Examples

## Tinkering with `plotColorMixture`

```{r fig.height=5.5, fig.width=9}
plotColorMixture(c('10YR 5/3', '10YR 3/2'), swatch.cex = 5, showMixedSpec = TRUE)

plotColorMixture(c('10YR 6/2', '5YR 5/6'), swatch.cex = 5, showMixedSpec = TRUE)

# is this right? can the resulting spectra be "higher" than the source?
plotColorMixture(c('10YR 6/2', '5YR 5/6'), w = c(2,1), swatch.cex = 5, showMixedSpec = TRUE)

# label collision
plotColorMixture(c('10YR 5/3', '10YR 3/2', '5R 2/2'), swatch.cex = 5, showMixedSpec = TRUE)
plotColorMixture(c('10YR 5/3', '10YR 3/2', '5R 2/2'), swatch.cex = 4, label.cex = 0.65, showMixedSpec = TRUE)


plotColorMixture(c('10YR 4/6', '2.5Y 6/2', '5Y 2/2'), w = c(1, 1, 2), swatch.cex = 5, showMixedSpec = TRUE)

plotColorMixture(c('5G 6/5', '5R 5/4'), w = c(1, 2))
plotColorMixture(c('5G 6/5', '5R 5/4'), w = c(3, 1), swatch.cex = 4, label.cex = 0.65, showMixedSpec = TRUE)
```

Check: "blue" + "yellow" = "green", yes. Note weighted geometric mean spectra (dotted line): closest chip isn't a perfect fit but close enough.
```{r fig.height=5.5, fig.width=9}
plotColorMixture(c('5B 5/10', '5Y 8/8'), swatch.cex = 4, label.cex = 0.65, showMixedSpec = TRUE)
```


Top 3 mixture candidates
```{r fig.height=5.5, fig.width=9}
plotColorMixture(c('5B 5/10', '5Y 8/8'), swatch.cex = 4, label.cex = 0.65, showMixedSpec = TRUE, n = 3)
```

## Simulation of a continuous mixture
Seems plausible.
```{r fig.height=2.5, fig.width=10}
s <- seq(0, 1, by = 0.2)
z <- lapply(s, function(i) {
  
  m <- mixMunsell(c('5B 5/10', '5Y 8/8'), w = c(1-i, i))
  
  return(m$munsell)
})

z <- do.call('c', z)

par(bg = 'black', fg = 'white', mar = c(0,0,0,0))
soilPalette(parseMunsell(z), sprintf('%s (%s / %s))', z, 1 - s, s))
```



## Interpreting Spectral (Gower) Distances
Should the mixture of two `10YR` colors always result in a `10YR` mixture candidate? Not necessarily (blue + yellow = green).
```{r}
m <- c('10YR 8/6', '10YR 2/2')
x <- mixMunsell(m, n = 20)

# it isn't
x[1, ]
```

The weighted geometric mean of the two spectra seems plausible:
```{r fig.height=5.5, fig.width=9}
plotColorMixture(c('10YR 8/6', '10YR 2/2'), swatch.cex = 5, showMixedSpec = TRUE)
```


Evaluate the top 20 mixture candidates in terms of spectral distance and $\Delta{E_{00}}$ relative to the most likely match.
```{r}
z <- expand.grid(m1 = x$munsell[1], m2 = x$munsell)
d <- colorContrast(z[[1]], z[[2]])

n <- nrow(x)
s <- 1:n
cols <- parseMunsell(x$munsell)
```

Prepare some figures to assist with the exploration.
```{r fig.width=8, fig.height=5}
par(mar = c(4.5, 4.5, 2, 1), bg = 'black', fg = 'white')

plot(s, x$distance, type = 'n', ylab = 'Spectral Distance', xlab = 'Match Candidate', col.axis = 'white', col.lab = 'white', axes = FALSE)
grid()
lines(s, x$distance, type = 's')
points(s, x$distance, pch = 15, col = cols, cex= 4)
text(s, x$distance, gsub(' ', '\n', x$munsell), cex = 0.5, col = invertLabelColor(cols))
axis(side = 1, at = s, col.axis = 'white')
axis(side = 2, col.axis = 'white', las = 1, cex.axis = 0.75)
title(sprintf("Mixture: %s + %s", m[1], m[2]), col.main = 'white')
```


```{r fig.width=8, fig.height=5}
par(mar = c(4.5, 4.5, 2, 1), bg = 'black', fg = 'white')

plot(s, d$dE00, type = 'n', ylab = expression(~Delta*E['00']), xlab = 'Match Candidate', col.axis = 'white', col.lab = 'white', axes = FALSE)
grid()
lines(s, d$dE00, type = 's')
points(s, d$dE00, pch = 15, col = cols, cex= 4)
text(s, d$dE00, gsub(' ', '\n', x$munsell), cex = 0.5, col = invertLabelColor(cols))
axis(side = 1, at = s, col.axis = 'white')
axis(side = 2, at = 0:15, col.axis = 'white', las = 1, cex.axis = 0.75)
title(sprintf("Mixture: %s + %s", m[1], m[2]), col.main = 'white')
```


The relationship between spectral distance and $\Delta{E_{00}}$ is complex.
```{r fig.width=8, fig.height=5}
par(mar = c(4.5, 4.5, 2, 1), bg = 'black', fg = 'white')

plot(x$distance, d$dE00, type = 'n', xlab = 'Spectral Distance', ylab = expression(~Delta*E['00']), col.axis = 'white', col.lab = 'white', axes = FALSE)
grid()
points(x$distance, d$dE00, pch = 15, col = cols, cex= 4.5)
text(x$distance, d$dE00, gsub(' ', '\n', x$munsell), cex = 0.66, col = invertLabelColor(cols))
axis(side = 1, col.axis = 'white')
axis(side = 2, at = 0:15, col.axis = 'white', las = 1, cex.axis = 0.75)
title(sprintf("Mixture: %s + %s", m[1], m[2]), col.main = 'white')
```

Try again, this time using the entire spectral library.
```{r fig.width=8, fig.height=8}
m <- c('10YR 8/6', '10YR 2/2')

# return distances to entire spectral library
x <- mixMunsell(m, n = 2486)

# do not compare 1st mixture candidate to itself (dE00 = 0)
z <- expand.grid(m1 = x$munsell[1], m2 = x$munsell[-1])
d <- colorContrast(z[[1]], z[[2]])

# color vector, note that we are skipping the first color
cols <- parseMunsell(x$munsell[-1])

par(mar = c(4.5, 4.5, 2, 1), bg = 'black', fg = 'white')

# skip first mixture candidate
plot(x$distance[-1], d$dE00, xlab = 'Spectral Distance', ylab = expression(~Delta*E['00']), col.axis = 'white', col.lab = 'white', las = 1, type = 'n', ylim = c(0, 50))

grid()

# skip first mixture candidate
points(x$distance[-1], d$dE00, pch = 15, col = cols, cex = 0.5)
title(sprintf("Mixture Candidates: %s + %s", m[1], m[2]), col.main = 'white')

# add first mixture candidate
points(x$distance[1], d$dE00[1], bg = cols[1], pch = 22)
```


Reasonable thresholds on spectral distance between mixture and mixture candidates.
```{r fig.height=5.5, fig.width=10}
# assuming threshold of dE00
reasonable.matches <- d$m2[which(d$dE00 <= 5)]

# find these colors in the mixture candidates
idx <- which(x$munsell %in% reasonable.matches)

# combine mixture candidates + contrast metrics
z <- merge(x[idx, ], d, by.x = 'munsell', by.y = 'm2', sort = FALSE, incomparables = FALSE)

# display
colorContrastPlot(z$m1, z$munsell, labels = c('First Mixture Canidate', 'Subsequent Candidates'))
```

So: spectral distances to the first mixture candidate of $> 0.015$ may be unreliable. TBD.

### Spectral Distance vs. dE00
TODO: pair-wise dE00 vs. pair-wise spectral distances.
```{r fig.width = 8.5, fig.height=5}
library(cluster)

# spectral library
data("munsell.spectra.wide")

# remove 'wavelength' column and transpose
z <- t(munsell.spectra.wide[, -1])

# all pair-wise distances, using Gower's distance metric
d <- daisy(z, metric = 'gower', stand = FALSE)
# reduce distance matrix to vector
d <- as.vector(d)

# quick plot
par(mfcol = c(1, 2), mar = c(4.5, 4, 3, 1))
hist(d, main = 'All Spectra', axes = FALSE, xlab = 'Spectral Distance (Gower)', ylab = 'Frequency', breaks = 50)
box()
axis(1, at = seq(0, 1, by = 0.1), cex.axis = 0.75)
axis(2, las = 1, cex.axis = 0.75)

hist(d[d < 0.1], main = 'Spectra d < 0.1', axes = FALSE, xlab = 'Spectral Distance (Gower)', ylab = '', breaks = 50)
box()
axis(1, at = seq(0, 0.1, by = 0.01), cex.axis = 0.75)
axis(2, las = 1, cex.axis = 0.75)
```



# Alternative color mixing strategies
How does `soilDB::estimateColorMixture` (weighted mean of CIELAB coordinates) compare? Not bad (~2 chips off), but the subtractive mixture seems more likely.
```{r fig.height=3.5, fig.width=6}
library(soilDB)

colors <- c('10YR 6/2', '7.5YR 3/3')

d <- cbind(
  parseMunsell(colors, convertColors=FALSE),
  parseMunsell(colors, return_triplets=TRUE, returnLAB=TRUE),
  pct=c(0.5, 0.5),
  col=parseMunsell(colors, convertColors=TRUE)
)

e1 <- estimateColorMixture(d, backTransform = TRUE)
m1 <- mixMunsell(colors)
```

```{r fig.height=5.5, fig.width=9}
plotColorMixture(colors)
```

```{r fig.height=3.5, fig.width=6}
par(bg = 'black', fg = 'white', mar = c(0,0,0,0))
colorContrastPlot(
  m1$munsell,
  sprintf('%s %s/%s', e1$colorhue, e1$colorvalue, e1$colorchroma),
  labels = c('Subtractive Mixture', 'Weighted Average CIELAB')  
  )
```

What does a $\Delta E_{00}$ of **4.55** mean in terms of "color chips" in a simulated color book?
```{r fig.height=5, fig.width=10}
contrastChart('10YR 6/2', hues = c('10YR', '7.5YR'), thresh = 4.55)
```





----------------------------
This document is based on `aqp` version `r utils::packageDescription("aqp", field="Version")`.
