---
output:
  html_vignette:
    mathjax: null
    jquery: null
    smart: no
---

```{r setup, echo=FALSE, results='hide', warning=FALSE}
library(knitr, quietly = TRUE)

opts_chunk$set(message=FALSE, warning=FALSE, out.width="100%", fig.retina = 2, fig.align='center', dev='png', dev.args=list(pointsize=10, antialias='cleartype'), tidy=TRUE)
options(width=100, stringsAsFactors=FALSE)

knit_hooks$set(htmlcap = function(before, options, envir) {
  if(!before) {
    paste('<p class="caption" style="font-size:85%; font-style: italic; font-weight: bold;">',options$htmlcap,"</p><hr>",sep="")
    }
    })
```


Pair-Wise Distance Between Profiles<br>Mixed Variable Types
=======================================
D.E. Beaudette
<br>
`r format(Sys.time(), "%Y-%m-%d")`
<br>
This document is based on `aqp` version `r utils::packageDescription("aqp", field="Version")` and `sharpshootR` version `r utils::packageDescription("sharpshootR", field="Version")`.


# Introduction




# Setup R Envionment
If you have never used the [aqp](http://ncss-tech.github.io/AQP/aqp/aqp-intro.html) or [soilDB](http://ncss-tech.github.io/AQP/soilDB/soilDB-Intro.html) packages before, you will likely need to install them. This only needs to be done once. 
```{r install-packages, eval=FALSE}
# stable versions + deps
install.packages('aqp', dep=TRUE)
install.packages('soilDB', dep=TRUE)
install.packages('sharpshootR', dep=TRUE)

# latest versions
devtools::install_github("ncss-tech/aqp", dependencies=FALSE, upgrade_dependencies=FALSE)
devtools::install_github("ncss-tech/sharpshootR", dependencies=FALSE, upgrade_dependencies=FALSE)
```

Once you have all of the R packages on which  this document depends, it is a good idea to load them. R packages must be **installed** anytime you change versions of R (e.g., after an upgrade) and **loaded** anytime you want to access functions from within those packages.

```{r load-packages}
library(aqp)
library(soilDB)
library(sharpshootR)
library(cluster)
library(RColorBrewer)
```

# Sample Data
While the methods outlined in this document can be applied to any collection of pedons, it is convenient to work with a standardized set of data. You can follow along with the analysis by copying code from the following blocks and running it in your **R** session. The sample data used in this document is based on soil profiles that have been correlated to the [Loafercreek](https://soilseries.sc.egov.usda.gov/OSD_Docs/L/LOAFERCREEK.html) soil series from the Sierra Nevada Foothill Region of California. Note that the internal structure of the `loafercreek` data is identical to the structure returned by [`fetchNASIS()` from the soilDB package](http://ncss-tech.github.io/AQP/soilDB/soilDB-Intro.html). All horizon-level values are pulled from the pedon horizon table of the pedons being analyzed.

```{r load-data, fig.width=9, fig.height=4, htmlcap='Pedons correlated to the Loafercreek soil series.'}
# load sample data from the soilDB package
data(loafercreek, package = 'soilDB')

# get a subset of profiles to work with
x <- loafercreek[20:30, ]
```


# Methods

## Check Data
```{r fig.width=10, fig.height=6, htmlcap=''}
# check existing texture
table(x$texture_class)

# order field texture according to plant available water low -> high
# note: leaving out "in-lieu-of" texture classes for organic or bedrock
x$texture_class <- factor(x$texture_class, ordered = TRUE, levels = c('sl', 'scl', 'l', 'sil', 'c', 'sic', 'cl', 'sicl'))

# graphical check
par(mar=c(0,1,3,1))
plot(x, color='texture_class', label='pedon_id', col.palette=brewer.pal(10, 'Spectral'))
addVolumeFraction(x, 'total_frags_pct')

# make a fake name with texture class + RF volume:
x$fake.name <- ifelse(!is.na(x$texture_class), paste0(x$texture_class, ' - ', x$total_frags_pct),  '')
plot(x, color='texture_class', name='fake.name', label='pedon_id', cex.name=0.65, col.palette=brewer.pal(10, 'Spectral'))
addVolumeFraction(x, 'total_frags_pct')
```

## Horizon Level Attributes
```{r fig.width=10, fig.height=6, htmlcap=''}
# make an index to soil horizons (anything other than O|Cr|R)
idx <- grep('O|Cr|R', x$hzname, invert = TRUE)

# compute pair-wise distances using texture class (ordered factor) and rock frag volume
d.hz <- profile_compare(x, vars=c('texture_class', 'total_frags_pct'), max_d=100, k=0, filter=idx, replace_na=TRUE, add_soil_flag=TRUE)
# divisive hierarchical clustering
dd.hz <- diana(d.hz)

# plot dendrogram + profiles
par(mar=c(2,1,3,1))
plotProfileDendrogram(x, dd.hz, scaling.factor = 0.75, y.offset = 5, width=0.15, color='texture_class', name='fake.name', label='pedon_id', cex.name=0.75, col.palette=brewer.pal(10, 'Spectral'))
addVolumeFraction(x, 'total_frags_pct')
```


## Site and Horizon Level Attributes
```{r fig.width=10, fig.height=6, htmlcap=''}
# fix some missing data
x$elev_field[x$elev_field == 0] <- 380

# graphical check: order pedons by slope
par(mar=c(0,1,3,1))
plot(x, color='texture_class', label='pedon_id', col.palette=brewer.pal(10, 'Spectral'), plot.order=order(x$slope_field))
addVolumeFraction(x, 'total_frags_pct')
axis(side=1, at=1:length(x), labels=x$slope_field[order(x$slope_field)], line=-3)
mtext(text='Field Described Slope (%)', side=1, line=-4)

# graphical check: order pedons by elevation
par(mar=c(0,1,3,1))
plot(x, color='texture_class', label='pedon_id', col.palette=brewer.pal(10, 'Spectral'), plot.order=order(x$elev_field))
addVolumeFraction(x, 'total_frags_pct')
axis(side=1, at=1:length(x), labels=x$elev_field[order(x$elev_field)], line=-3)
mtext(text='Field Described Slope (m)', side=1, line=-4)
```

```{r fig.width=10, fig.height=6, htmlcap=''}
# compute pair-wise distances using:
# hz attributes: texture class (ordered factor) and rock frag volume
# site attributes: surface slope and elevation
d.hz.site <- profile_compare(x, vars=c('texture_class', 'total_frags_pct', 'slope_field', 'elev_field'), max_d=100, k=0, filter=idx, replace_na=TRUE, add_soil_flag=TRUE)
# divisive hierarchical clustering
dd.hz.site <- diana(d.hz.site)

# plot dendrogram + profiles
par(mar=c(2,1,3,1))
plotProfileDendrogram(x, dd.hz.site, scaling.factor = 0.008, y.offset = 0.1, width=0.15, color='texture_class', name='fake.name', label='pedon_id', cex.name=0.75, col.palette=brewer.pal(10, 'Spectral'))
addVolumeFraction(x, 'total_frags_pct')
```
