---
title: "Pair-Wise Distance Between Profiles and Mixed Variable Types"
author: "D.E. Beaudette"
date: "`r Sys.Date()`"
output:
  html_document:
  mathjax: null
jquery: null
smart: no
keep_md: no
---
  
```{r setup, echo=FALSE, results='hide', warning=FALSE}
library(knitr)
opts_chunk$set(message=FALSE, warning=FALSE, background='#F7F7F7', fig.retina=2, dev='png', tidy=FALSE, verbose=FALSE, cache = FALSE)
options(width=100, stringsAsFactors=FALSE, cache=FALSE)
```



# Introduction


# Setup R Envionment
If you have never used the [aqp](http://ncss-tech.github.io/AQP/aqp/aqp-intro.html) or [soilDB](http://ncss-tech.github.io/AQP/soilDB/soilDB-Intro.html) packages before, you will likely need to install them. This only needs to be done once. 
```{r install-packages, eval=FALSE}
# stable versions + deps
install.packages('aqp', dep=TRUE)
install.packages('soilDB', dep=TRUE)
install.packages('sharpshootR', dep=TRUE)

# latest versions
devtools::install_github("ncss-tech/aqp", dependencies=FALSE, upgrade_dependencies=FALSE)
devtools::install_github("ncss-tech/sharpshootR", dependencies=FALSE, upgrade_dependencies=FALSE)
```

Once you have all of the R packages on which  this document depends, it is a good idea to load them. R packages must be **installed** anytime you change versions of R (e.g., after an upgrade) and **loaded** anytime you want to access functions from within those packages.

```{r load-packages}
library(aqp)
library(soilDB)
library(sharpshootR)
library(cluster)
library(RColorBrewer)
library(ape)
```


# Simple Example via Soil Texture
```{r}
# soil textures, sorted according to field book
data("ROSETTA.centroids")

# ordered vs. nominal texture classes
tex.ordered <- ROSETTA.centroids$texture
tex.nominal <- factor(ROSETTA.centroids$texture, levels=sort(as.character(ROSETTA.centroids$texture)), ordered = FALSE)

# assemble into DF, note that stringsAsFactors=FALSE required
x.1 <- data.frame(
  p1=tex.ordered,
  stringsAsFactors = FALSE
  )

x.2 <- data.frame(
  p1=tex.nominal,
  stringsAsFactors = FALSE
)

# copy texture classes to row names for automatic distance matrix / clustering labels
row.names(x.1) <- x.1$p1                  
row.names(x.2) <- x.2$p1

# 1D clustering
d.1 <- daisy(x.1[, 1, drop=FALSE], metric = 'gower') 
d.2 <- daisy(x.2[, 1, drop=FALSE], metric = 'gower')

## note differences
# d.1
# d.2

# check: pair-wise distances should reflect information contained in ranks of ordinal representation
par(mar=c(1,1,3,1), mfcol=c(1,2))
plot(as.phylo(as.hclust(diana(d.1))), main='Ordinal', label.offset=0.01)
plot(as.phylo(as.hclust(diana(d.2))), main='Nominal', label.offset=0.01)

```


# Sample Data
While the methods outlined in this document can be applied to any collection of pedons, it is convenient to work with a standardized set of data. You can follow along with the analysis by copying code from the following blocks and running it in your **R** session. The sample data used in this document is based on soil profiles that have been correlated to the [Loafercreek](https://soilseries.sc.egov.usda.gov/OSD_Docs/L/LOAFERCREEK.html) soil series from the Sierra Nevada Foothill Region of California. Note that the internal structure of the `loafercreek` data is identical to the structure returned by [`fetchNASIS()` from the soilDB package](http://ncss-tech.github.io/AQP/soilDB/soilDB-Intro.html). All horizon-level values are pulled from the pedon horizon table of the pedons being analyzed.

```{r load-data, fig.width=9, fig.height=4}
# load sample data from the soilDB package
data(loafercreek, package = 'soilDB')

# get a subset of profiles to work with
x <- loafercreek[10:20, ]
```


# Methods

## Check Data
```{r fig.width=10, fig.height=6}
# check existing textures
table(x$texture_class)

# order field texture according to particle size
# note: leaving out "in-lieu-of" texture classes for organic soil material and bedrock
x$texture_class <- factor(x$texture_class, ordered = TRUE, levels = SoilTextureLevels())
x$texture_class <- droplevels(x$texture_class)

# make a copy without ordering
x$texture_class_nominal <- factor(x$texture_class, levels=sort(as.character(unique(x$texture_class)), decreasing = TRUE), ordered = FALSE)

# graphical check
par(mar=c(0,1,3,1))
plot(x, color='texture_class', label='pedon_id')
plot(x, color='texture_class_nominal', label='pedon_id')

# make a fake horizon-level attr, until #7 is fixed
x$constant <- rep(1, times=nrow(x))
```

## Horizon Level Attributes
```{r fig.width=10, fig.height=6, htmlcap=''}

# compute pair-wise distances using texture class (ordered factor)
d.hz <- profile_compare(x, vars=c('texture_class', 'constant'), max_d=100, k=0, rescale.result=TRUE)
# divisive hierarchical clustering
dd.hz <- diana(d.hz)

# plot dendrogram + profiles
par(mar=c(1,0,3,1))
plotProfileDendrogram(x, dd.hz,  width=0.25, color='texture_class', label='pedon_id', cex.name=0.65)
```


```{r fig.width=10, fig.height=6, htmlcap=''}
# compute pair-wise distances using texture class (nominal factor)
d.hz <- profile_compare(x, vars=c('texture_class_nominal', 'constant'), max_d=100, k=0, rescale.result=TRUE)
# divisive hierarchical clustering
dd.hz <- diana(d.hz)

# plot dendrogram + profiles
par(mar=c(1,0,3,1))
plotProfileDendrogram(x, dd.hz,  width=0.25, color='texture_class_nominal', label='pedon_id', cex.name=0.65)
```



## Site and Horizon Level Attributes
```{r fig.width=10, fig.height=6, htmlcap=''}
# fix some missing data
x$elev_field[x$elev_field == 0] <- 380

# graphical check: order pedons by slope
par(mar=c(0,1,3,1))
plot(x, color='texture_class', label='pedon_id', col.palette=brewer.pal(10, 'Spectral'), plot.order=order(x$slope_field))
addVolumeFraction(x, 'total_frags_pct')
axis(side=1, at=1:length(x), labels=x$slope_field[order(x$slope_field)], line=-3)
mtext(text='Field Described Slope (%)', side=1, line=-4)

# graphical check: order pedons by elevation
par(mar=c(0,1,3,1))
plot(x, color='texture_class', label='pedon_id', col.palette=brewer.pal(10, 'Spectral'), plot.order=order(x$elev_field))
addVolumeFraction(x, 'total_frags_pct')
axis(side=1, at=1:length(x), labels=x$elev_field[order(x$elev_field)], line=-3)
mtext(text='Field Described Slope (m)', side=1, line=-4)
```

```{r fig.width=10, fig.height=6, htmlcap=''}
# compute pair-wise distances using:
# hz attributes: texture class (ordered factor) and rock frag volume
# site attributes: surface slope and elevation
d.hz.site <- profile_compare(x, vars=c('texture_class', 'total_frags_pct', 'slope_field', 'elev_field'), max_d=100, k=0)
# divisive hierarchical clustering
dd.hz.site <- diana(d.hz.site)

# plot dendrogram + profiles
par(mar=c(2,1,3,1))
plotProfileDendrogram(x, dd.hz.site, scaling.factor = 0.008, y.offset = 0.1, width=0.15, color='texture_class', name='fake.name', label='pedon_id', cex.name=0.75, col.palette=brewer.pal(10, 'Spectral'))
addVolumeFraction(x, 'total_frags_pct')
```


----------------------------
This document is based on `aqp` version `r utils::packageDescription("aqp", field="Version")`, `soilDB` version `r utils::packageDescription("soilDB", field="Version")`, and `sharpshootR` version `r utils::packageDescription("sharpshootR", field="Version")`.
