---
output:
  html_vignette:
    mathjax: null
    jquery: null
    smart: no
---

```{r setup, echo=FALSE, results='hide', warning=FALSE}
library(knitr, quietly = TRUE)
library(printr, quietly = TRUE)
opts_chunk$set(message=FALSE, warning=FALSE, out.width="100%", fig.retina = 2, fig.align='center', dev='png', dev.args=list(pointsize=10, antialias='cleartype'), tidy=TRUE)
options(width=100, stringsAsFactors=FALSE)

knit_hooks$set(htmlcap = function(before, options, envir) {
  if(!before) {
    paste('<p class="caption" style="font-size:85%; font-style: italic; font-weight: bold;">',options$htmlcap,"</p><hr>",sep="")
    }
    })
```


Soil Color Transition Probability
=======================================
D.E. Beaudette
<br>
`r format(Sys.time(), "%Y-%m-%d")`
<br>
This document is based on `aqp` version `r utils::packageDescription("aqp", field="Version")`, `soilDB` version `r utils::packageDescription("soilDB", field="Version")`, and `sharpshootR` version `r utils::packageDescription("sharpshootR", field="Version")`.

Details pending...


Use pedon data associated with the Pierre soil series.
```{r}
library(soilDB)
library(sharpshootR)
library(igraph)
library(plyr)
library(reshape2)
library(markovchain)

# get lab / morphologic data for the Pierre soil series
x <- fetchKSSL(series='pierre', returnMorphologicData = TRUE)

# extract pedons into SoilProfileCollection
s <- x$SPC

# extract horizon data from SPC
h <- horizons(s)

# simplify color data: 1 row / horizon, from morphologic data tables
x.colors <- simplifyColorData(x$morph$phcolor, id.var = 'labsampnum')

# merge color data into SPC
h <- join(h, x.colors, by='labsampnum', type='left', match='first')

# remove horizons that are missing moist colors
h <- subset(h, h$m_hue != '' & ! is.na(h$m_hue) & ! is.na(h$m_value) & ! is.na(h$m_chroma))

# re-assemble Munsell color notation for moist color
h$color <- paste0(h$m_hue, ' ', h$m_value, '/', h$m_chroma)

# pack horizon data back into SPC
horizons(s) <- h
```


Alternate data source.
```{r, eval=FALSE}
# load AQP sample dataset 5
data(sp5)
s <- sp5

# convert soil color to RGB
s.rgb <- t(col2rgb(s$soil_color))
s.rgb <- s.rgb / 254

# convert soil color to closest Munsell chip
s.munsell <- rgb2munsell(horizons(s)[, c('R25', 'G25', 'B25')])

s$color <- paste0(s.munsell$hue, ' ', s.munsell$value, '/', s.munsell$chroma)

s$moist_soil_color <- s$soil_color
```


```{r, fig.width=9, fig.height=9}
# generate TP matrix from horizon moist colors
tp <- hzTransitionProbabilities(s, "color", loopTerminalStates = FALSE)

# greate graph object
par(mar = c(1, 1, 1, 1))
g <- plotSoilRelationGraph(tp, graph.mode = "directed", vertex.scaling.factor=2, edge.arrow.size = 0.5, edge.scaling.factor = 2.5, vertex.label.cex = 0.75, vertex.label.family = "sans")


# join Munsell color notation to graph nodes
d <- data.frame(color=V(g)$name, stringsAsFactors = FALSE)
d <- join(d, horizons(s), by = 'color', type='left', match='first')
V(g)$color <- d$moist_soil_color

# prepare labels by converting to HSV
hsv.cols <- t(rgb2hsv(col2rgb(V(g)$color)))
hsv.cols[, 1] <- 0
hsv.cols[, 2] <- 0
hsv.cols[, 3] <- ifelse(hsv.cols[, 3] > 0.5, 0, 1)
V(g)$label.color <- hsv(hsv.cols[, 1], hsv.cols[, 2], hsv.cols[, 3])

# remove loops from graph, retain duplicate paths
g <- simplify(g, remove.loops = TRUE, remove.multiple=FALSE)

# final plot
par(mar=c(0,0,1,0), bg=grey(0.85))
set.seed(1010101)
plot(g, edge.arrow.size = 0.5, vertex.label.cex = 0.55, vertex.label.family = "sans", vertex.label.font=2, edge.color='black')
```

