---
title: "Soil Profile Sketches"
author: "D.E. Beaudette"
date: "`r Sys.Date()`"
output:
  html_document:
    mathjax: null
    jquery: null
    smart: no
    number_sections: no
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, echo=FALSE, results='hide', warning=FALSE}
library(knitr, quietly=TRUE)

opts_chunk$set(
  message = FALSE, 
  warning = FALSE, 
  background = '#F7F7F7', 
  fig.align = 'center', 
  dev = 'svglite', 
  tidy = FALSE, 
  verbose = FALSE
)

options(width=100, stringsAsFactors=FALSE)
```


# Introduction
This is an expanded description of the arguments to `plotSPC()` and tips on making soil profile sketches look the way you want them to. Be sure to get the latest version of [aqp](https://github.com/ncss-tech/aqp/).

# Example Data
```{r}
library(aqp)
library(soilDB)

# some interesting soil series
s <- c('leon', 'musick', 'clarksville', 'pardee', 'lucy', 'pierre', 'drummer', 'zook', 'san joaquin')

# get basic morphology and extended data from SoilWeb cache
osds.full <- fetchOSD(s, extended = TRUE)

# save copy of SoilProfileCollection for later
osds <- osds.full$SPC
```

# Canvas
See the [SPC introduction for the basics](http://ncss-tech.github.io/AQP/aqp/aqp-intro.html#10_Plotting_SoilProfileCollection_Objects).
```{r fig.width=10, fig.height=6}
explainPlotSPC(osds)
```



# Arguments to `plotSPC`

## `x`
A `SoilProfileCollection` object with one or more profiles. Large collections (>25 profiles) may be hard to interpret on screen. Consider save the output directly to a PNG or PDF, for example:

```{r eval = FALSE}
# start output device, dimensions are in pixels
png(filename = 'output.png', width = 1600, height = 800)

# plot
plotSPC(reallyBigSPC)

# close device, write file
dev.off()
```

Turning off ID and horizon designation annotation is another strategy for getting a quick glimpse of a large collection. Turning off profile and horizon boundary outlines with `divide.hz = FALSE` can help too.
```{r fig.width=10, fig.height=4}
# a single profile from our example
musick <- subset(osds, id == 'MUSICK')

# standard deviation of horizon boundary variation
horizons(musick)$hzb <- 10

# simulate 50 profiles
musick.sim <- perturb(musick, n = 50, boundary.attr = 'hzb')

# no margins
par(mar = c(0, 0, 0, 0))
# note other arguments used to improve legibility
plotSPC(musick.sim, print.id = FALSE, name = NA, plot.depth.axis = TRUE, width = 0.4, divide.hz = FALSE, axis.line.offset = -4, cex.depth.axis = 0.75)
```




## Profile Arrangement

### `n`
A single integer describing the amount of space along x-axis to allocate, defaults to `length(x)`.

```{r fig.width=10, fig.height=6.5}
explainPlotSPC(osds, cex.names = 0.66, width = 0.3, n = 15, print.id = FALSE, name.style = 'center-center')

text(x = 10.5, y = 90, label = "additional space allocated by 'n = 15'", adj = 0, cex = 0.75)
arrows(x0 = length(osds) + 1, x1 = 15, y0 = 100, y1 = 100, code = 3, length = 0.1)
```

### `plot.order`
integer vector describing the order in which individual soil profiles should be plotted

### `relative.pos`
vector of relative positions along the x-axis, within {1, n}, ignores plot.order see details


### `add`
When `TRUE`, sketches and annotations (like a depth axis) are added to the current output device. With careful use of `x.idx.offset` it is possible to combine several sets of profiles, such as the results of a simulation (see [`perturb()`](http://ncss-tech.github.io/aqp/reference/perturb.html)). With careful use of `scaling.factor`, profiles can be added to any figure.

```{r fig.width=10, fig.height=6}
# select first profile
x <- osds[1, ]

# convert horizon distinctness codes 
# to a horizon boundary "variability" index
horizons(x)$hzd <- hzDistinctnessCodeToOffset(x$distinctness) * 2

# generate 10 simulations
x.sim <- perturb(x, n = 10, boundary.attr = 'hzd')

# margins
par(mar = c(0, 0, 0, 0))

# plot the source profile, leaving room for 11 profiles (n = 11)
plotSPC(x, cex.names = 0.66, name.style = 'center-center', width = 0.3, plot.depth.axis = FALSE, hz.depths = TRUE, hz.depths.offset = 0.05, n = 11, id.style = 'side', cex.id = 0.66, fixLabelCollisions = TRUE, max.depth = 145)

# add 10 simulations, offset 1 position to the right with x.idx.offset = 1
plotSPC(x.sim, cex.names = 0.66, name.style = 'center-center', width = 0.3, plot.depth.axis = FALSE, hz.depths = TRUE, hz.depths.offset = 0.05, id.style = 'top', add = TRUE, x.idx.offset = 1, fixLabelCollisions = TRUE)

# annotate simulations
arrows(x0 = 2, x1 = 11, y0 = -15, y1 = -15, length = 0.1, code = 3)
mtext('Simulation', side = 3, at = 5.5, line = -2.5, font = 2, adj = 0)
```


### `x.idx.offset`
Horizontal offset from 0 (left-hand edge), used to shift all profiles left or right within the sketch. Sometimes this is useful for making small, aesthetic adjustments especially when making sketches of large collections. When used together with the `add` argument, it is possible to align multiple versions of the same data by shifting profiles left or right. The following example adds profile sketches that use dry soil colors next to the original (moist soil colors) sketches.

```{r fig.width=10, fig.height=6.5}
# convert dry colors to hex notation
osds$dry_color <- munsell2rgb(osds$dry_hue, osds$dry_value, osds$dry_chroma)

# adjust margins
par(mar = c(1.5, 0, 2, 1))

# original data, supress horizon designations
plotSPC(osds, width = 0.2, name = NA)

# add and shift dry color sketches to the left (-0.4 units)
plotSPC(osds, width = 0.15, cex.names = 0.66, plot.depth.axis = FALSE, print.id = FALSE, name = NA, x.idx.offset = -0.4, add = TRUE, color = 'dry_color')

# a title
title('Moist and Dry Colors')

# demonstrate shift with an arrow below each profile
s <- 1:length(osds)
bottoms <- profileApply(osds, max) + 15
arrows(x0 = s, y0 = bottoms, x1 = s - 0.4, y1 = bottoms, length = 0.1)

# annotate moist vs. dry
text(x = s, y = bottoms, labels = 'M', pos = 3, font = 3, cex = 0.66)
text(x = s - 0.4, y = bottoms, labels = 'D', pos = 3, font = 3, cex = 0.66)
```

### `y.offset`
Either a single value or numeric vector of vertical offsets with `length == length(x)`, in depth units of `x`. Profile sketches and depth axis are shifted when using a single y offset value. The depth axis is disabled when using multiple y offsets. Note: `y.offset` is automatically re-ordered by `plot.order`.

Shift all profile sketches down 300cm, leaving room for additional annotation above. Right-hand axis added for clarity.
```{r fig.width=10, fig.height=10}
par(mar = c(0, 0, 0, 0))

plotSPC(osds, cex.names = 0.66, name.style = 'center-center', width = 0.3, y.offset = 300, hz.depths = TRUE, fixLabelCollisions = TRUE, hz.depths.offset = 0.05, plot.depth.axis = FALSE)

# fake data
rect(xleft = (1:length(osds)) - 0.25, xright = (1:length(osds)) + 0.25, ybottom = 250, ytop = runif(n = length(osds), min = 0, max = 250), col = 4)

# demonstrate full y-range within figure
axis(side = 4, line = -3, las = 1)
```

Shift each profile sketch a different amount. Transect or cross-section figures can be manually created in this way. See also `scaling.factor` for applying a vertical scaling to all profiles.
```{r fig.width=10, fig.height=10}
# manually create a nice looking set of relative elevations along a pretend landform
.yshift <- c(0, 1, 4, 8, 15, 22, 30, 35, 40) * 5

# compute the Hurst Redness Index for each horizon
# this is a horizon-level attribute
osds$hr <- hurst.redness(osds$hue, osds$value, osds$chroma)

# compute the profile median of Hurst Redness Index
# this is now a site-level attribute
osds$profile.median.hr <- profileApply(osds, function(i) median(i$hr, na.rm = TRUE))

# rank profile median, Hurst Redness Index
# use this to sort profiles
o <- order(osds$profile.median.hr)

# however, this will complicate a manual specification of the landform surface
# now we need an index back to the original order
idx <- match(1:length(osds), o)

# margins
par(mar = c(0, 0, 0, 0))

# apply profile-specific y-offsets
# re-arrange horizontal position with 'plot.order = o'
plotSPC(osds, cex.id = 0.66, cex.names = 0.66, name.style = 'center-center', width = 0.3, y.offset = .yshift[idx], hz.depths = TRUE, fixLabelCollisions = TRUE, hz.depths.offset = 0.05, plot.order = o)

# demonstrate full y-range within figure
axis(side = 4, line = -3, las = 1)

# add idealized land surface
.xland <- 1:length(osds)
.yland <- .yshift - 30

# smooth via interpolation
.landSurface <- splinefun(.xland, .yland)
.s <- seq(0.5, length(osds) + 0.5, by = 0.1)

# add smoothed, idealized land surface
lines(x = .s, y = .landSurface(.s), lwd = 2)

# remind ourselves that this is fake
title('A Fake Diagram of Real Soils', line= -1.5)
```



### `max.depth`
This is a suggested lower depth for the sketches, data are not modified and profiles may extend beyond figure margins.



## Profile Styling

### `width`
Each profile is given 1 horizontal unit of space in the figure. The `width` argument is given as 1/2 of this space: typically ranging from 0.1--0.4. Values >= 0.5 will cause overlap.

```{r fig.width=10, fig.height=5}
par(mar = c(0, 0, 0, 1))

plotSPC(osds, cex.names = 0.66, print.id = FALSE, name = NA)

# annotate horizontal space allocated to each profile
abline(v = seq(0.5, length(osds) + 0.5, by = 1), lty = 2)
```


```{r fig.width=10, fig.height=5}
par(mar = c(0, 0, 0, 1))

# too skinny, but lots of room for extra annotation
plotSPC(osds, width = 0.1, cex.names = 0.66, print.id = FALSE)

# works OK without horizon designations on the right side
plotSPC(osds, width = 0.45, cex.names = 0.66, name = NA)

# about right
plotSPC(osds, width = 0.3, cex.names = 0.66)
```


### `scaling.factor`
vertical scaling of profile depths, useful for adding profiles to an existing figure

```{r fig.width=10, fig.height=6}
# margins
par(mar = c(0, 0, 0, 0))

# plot the source profile, leaving room for 11 profiles (n = 11)
plotSPC(osds, cex.names = 0.66, name = NA, width = 0.25, plot.depth.axis = FALSE, hz.depths = TRUE, n = length(osds) * 2, print.id = FALSE, fixLabelCollisions = TRUE)

plotSPC(osds, cex.names = 0.66, name = NA, width = 0.25, plot.depth.axis = FALSE, hz.depths = TRUE, print.id = FALSE, add = TRUE, x.idx.offset = length(osds), scaling.factor = 0.5, fixLabelCollisions = TRUE)

arrows(x0 = length(osds) + 1, x1 = length(osds) * 2, y0 = -15, y1 = -15, length = 0.1, code = 3)
mtext('scaling.factor = 0.5', side = 3, at = 12.5, line = -2.25, font = 2, adj = 0, cex = 0.85)
```


### `lwd`
line width multiplier used for sketches

### `lty`
line style used for sketches




## Horizon Styling

### `color`
A quoted column name containing R-compatible color descriptions (e.g. `#7A5C37FF`, approximately 10YR 4/4), or numeric / categorical data to be displayed thematically.

Numeric values.
```{r fig.width=10, fig.height=5}
# leave room at the top for a legend, and on the right hand side for depth axis.
par(mar = c(0, 0, 3, 1))

# horizon top depth
plotSPC(osds, color = 'top', col.label = 'Horizon Top Depth (cm)', cex.names = 0.66)

plotSPC(osds, color = 'value', col.label = 'Munsell Value (moist)', cex.names = 0.66)
```

Categorical values.
```{r fig.width=10, fig.height=5}
# set ordering of Munsell hue according to hue position
osds$hue_ordered <- factor(osds$hue, levels = huePosition(returnHues = TRUE))

# leave room at the top for a legend, and on the right hand side for depth axis.
par(mar = c(0, 0, 3, 1))

plotSPC(osds, color = 'hue_ordered', col.label = 'Munsell Value (moist)', cex.names = 0.66)

# reaction class is already an ordered factor, set in fetchOSD()
plotSPC(osds, color = 'pH_class', col.label = 'Reaction Class', cex.names = 0.66, col.legend.cex = 0.66)
```

User-defined colors in hex notation.
```{r fig.width=10, fig.height=5}
# assign each horizon a color of the rainbow
horizons(osds)$rainbowColors <- rainbow(n = nrow(osds))

par(mar = c(0, 0, 0, 0))
plotSPC(osds, color = 'rainbowColors', cex.names = 0.66, col.legend.cex = 0.66)

# color gradient from 2.5YR 6/8 -> 5Y 3/3
cr <- colorRampPalette(parseMunsell(c('2.5YR 6/8', '5Y 3/3')), space = 'Lab', interpolate = 'spline')

horizons(osds)$gradientColors <- cr(n = nrow(osds))

plotSPC(osds, color = 'gradientColors', cex.names = 0.66, col.legend.cex = 0.66)
```


### `density`
fill density used for horizon color shading, either a single integer or a quoted column name (horizon-level attribute) containing integer values (default is NULL, no shading)

### `show.legend`
logical, show legend? (default is TRUE)

### `col.label`
thematic legend title

### `col.palette`
color palette used for thematic sketches (default is rev(brewer.pal(10, 'Spectral')))

### `col.palette.bias`
color ramp bias (skew), see colorRamp

### `col.legend.cex`
scaling of thematic legend

### `n.legend`
approximate number of classes used in numeric legend, max number of items per row in categorical legend

### `default.color`
default horizon fill color used when color attribute is NA





## Annotation / Labeling


### `name`
Quoted column name of the (horizon-level) attribute containing horizon designations or labels, if missing `hzdesgnname(x)` is used. Suppress horizon name printing by setting `name = NA` or `name = ''`.

### `name.style`
Horizon designation label style, one of `c('right-center', 'center-center', 'left-center', 'left-top', 'center-top'))`. The `right-center` position has been the default in **aqp** since 2010. I prefer the `center-center` style. Note that horizon designation label colors are automatically adjusted for maximum contrast (by `invertLabelColor()`) when placed inside of a profile sketch.

```{r fig.width=10, fig.height=16}
par(mar=c(0,0,1,0), mfcol=c(5,1))

for(i in c('right-center', 'center-center', 'left-center', 'left-top', 'center-top')) {
  plotSPC(osds, width = 0.33, cex.names = 0.9, shrink=TRUE, print.id = FALSE, plot.depth.axis=FALSE, name.style=i)
  mtext(text = i, at = 0.5, side=1, line=-4, cex=1, adj = 0, font = 2)
}
```

### `shrink`
logical, reduce character scaling for 'long' horizon by 80%

### `shrink.cutoff`
character length defining 'long' horizon names

### `shrink.thin`
integer, horizon thickness threshold for shrinking horizon names by 80%, only activated when shrink = TRUE (NULL = no shrinkage)

### `abbr`
logical, abbreviate label

### `abbr.cutoff`
suggested minimum length for abbreviated label


### `label`
A single (quoted) site-level attribute name that will be used to label profiles. This defaults to `idname(x)` (e.g. the profile ID), but can be anything in `site(x)`. See `id.style` argument for adjusting side vs. top vs. automatic label positioning. Suppress printing of profile labels with `print.id = FALSE`.

```{r fig.width=10, fig.height=5}
par(mar = c(0, 0, 0, 1))
plotSPC(osds, cex.names = 0.66, label = 'subgroup', name.style = 'center-center', width = 0.3)
title('Subgroup Labels', line = -1)
```


### `hz.depths`
When `TRUE`, small horizon depth labels are placed just to the right of horizon top depths (right justified). The results may be illegible if there are too many profiles in the collection or if profiles are too wide. Works best with `name.style = 'center-center'` and `plot.depth.axis = FALSE`.

```{r fig.width=10, fig.height=5}
par(mar = c(0, 0, 0, 0))
plotSPC(osds, cex.names = 0.66, name.style = 'center-center', width = 0.3, plot.depth.axis = FALSE, hz.depths = TRUE)
```

### `hz.depths.offset`
Numeric values between 0.01--0.15 shift horizon depth annotation to the right of profile sketches.

```{r fig.width=10, fig.height=5}
par(mar = c(0, 0, 0, 0))
plotSPC(osds, cex.names = 0.66, name.style = 'center-center', width = 0.3, plot.depth.axis = FALSE, hz.depths = TRUE, hz.depths.offset = 0.05)
```


### `hz.depths.lines`
When `TRUE`, segments are drawn between horizon depth labels and actual horizon depth; this is useful when including horizon boundary distinctness and/or when `fixLabelCollisions = TRUE`.

```{r fig.width=10, fig.height=5}
par(mar = c(0, 0, 0, 0))
plotSPC(osds, cex.names = 0.66, name.style = 'center-center', width = 0.3, plot.depth.axis = FALSE, hz.depths = TRUE, hz.depths.offset = 0.05, hz.depths.lines = TRUE)
```

### `fixLabelCollisions`
When `TRUE`, use `aqp::fixOverlap()` to attempt fixing horizon depth labeling collisions. This will slow plotting of large collections. Enabling fixes also sets `hz.depths.lines = TRUE`. Adjustments made by `aqp::fixOverlap()` involve random permutations, so output is different each time. Use `set.seed` when you need consistent results.

```{r fig.width=10, fig.height=5}
par(mar = c(0, 0, 0, 0))
plotSPC(osds, cex.names = 0.66, name.style = 'center-center', width = 0.3, plot.depth.axis = FALSE, hz.depths = TRUE, hz.depths.offset = 0.05, fixLabelCollisions = TRUE)
```


### `alt.label` and `alt.label.col`
A single (quoted) site-level attribute name that will be used as secondary annotation, inset just below the top of each profile. Note that the `alt.label.col` argument will probably have to be used for legibility.

```{r fig.width=10, fig.height=5}
par(mar = c(0, 0, 0, 1))
plotSPC(osds, cex.names = 0.66, alt.label = 'tax_tempcl', width = 0.3, alt.label.col = 'white')
```

### `cex.names`
This is a baseline scaling factor applied to all labels in the figure: ID labels, horizon designations, horizon depths, and depth axis. Values around `0.6` are usually just about right.


### `cex.id`
Scaling factor applied to the profile IDs.

### `font.id`
Font style applied to the profile IDs, default is `2` (bold).


### `id.style`
Profile ID annotation style. The default value of `auto` uses a simple heuristic used to select from:

 * `top`: centered above each profile 
 * `side`: along the top-left edge of profiles



## Horizon Boundaries

### `divide.hz`
logical, divide horizons with line segment? (TRUE), see details

### `hz.distinctness.offset`
NULL, or quoted column name (horizon-level attribute) containing vertical offsets used to depict horizon boundary distinctness (same units as profiles), see details and codehzDistinctnessCodeToOffset; consider setting hz.depths.lines = TRUE when used in conjunction with hz.depths = TRUE

### `hz.topography.offset`
NULL, or quoted column name (horizon-level attribute) containing offsets used to depict horizon boundary topography (same units as profiles), see details and codehzTopographyCodeToOffset

### `hz.boundary.lty`
quoted column name (horizon-level attribute) containing line style (integers) used to encode horizon topography



## Depth Axis

### `plot.depth.axis`
Logical, include a depth axis on the right-hand side of the figure. You may have to increase the right-hand margin.

### `axis.line.offset`
Horizontal offset applied to depth axis (default is -2.5, larger numbers move the axis to the right).

### `cex.depth.axis`
Scaling factor applied specifically to the depth axis.
```{r fig.width=10, fig.height=5}
par(mar = c(0, 0, 0, 3))
plotSPC(osds, cex.names = 0.6, cex.depth.axis = 1.5, name.style = 'center-center', width = 0.3, plot.depth.axis = TRUE, hz.depths = TRUE, hz.depths.offset = 0.05, hz.depths.lines = TRUE)
```

### `n.depth.ticks`
Suggested number of ticks in depth axis.

















```{r eval = FALSE, echo = FALSE}

# encode horizon boundarydistinctness via vertical offset
osds$hd <- hzDistinctnessCodeToOffset(osds$distinctness)

# encode horizon boundary topography via vertical offset
osds$hzto <- hzTopographyCodeToOffset(osds$topography)

# also encode horizon boundary topography las line type
osds$hzto.lty <- hzTopographyCodeToLineType(osds$topography)



osds$bnd.code <- sprintf(
  "%s%s",
  substr(osds$distinctness, 1, 1),
  substr(osds$topography, 1, 1)
)

# remove missing (NA) labels
osds$bnd.code <- gsub('NANA', '', osds$bnd.code)

# ok
par(mar = c(0, 0, 0, 0))

plot(osds)

plotSPC(osds)

plotSPC(osds, id.style = 'side')

plotSPC(osds, id.style = 'top', width = 0.1)
plotSPC(osds, id.style = 'top', width = 0.5)

plotSPC(osds, id.style = 'top', width = 0.35)

plotSPC(osds, id.style = 'top', width = 0.35, name.style = 'center-center')

plotSPC(osds, id.style = 'top', width = 0.35, name.style = 'center-center', plot.depth.axis = FALSE, hz.depths = TRUE)

plotSPC(osds, id.style = 'top', width = 0.35, name.style = 'center-center', plot.depth.axis = FALSE, hz.depths = TRUE, cex.names = 0.66, cex.id = 0.5)

plotSPC(osds, id.style = 'top', width = 0.3, name.style = 'center-center', plot.depth.axis = FALSE, hz.depths = TRUE, cex.names = 0.66, cex.id = 0.5, fixLabelCollisions = TRUE)

plotSPC(osds, id.style = 'top', width = 0.3, name.style = 'center-center', plot.depth.axis = FALSE, hz.depths = TRUE, cex.names = 0.66, cex.id = 0.5, fixLabelCollisions = TRUE, hz.depths.offset = 0.05)


png(filename = 'figure.png', width = 1200, height = 900)

# pdf(file = 'poster.pdf', width = 96, height = 42)

plotSPC(osds, id.style = 'top', width = 0.3, name.style = 'center-center', plot.depth.axis = FALSE, hz.depths = TRUE, cex.names = 0.66, cex.id = 0.5, fixLabelCollisions = TRUE, hz.depths.offset = 0.05, shrink = TRUE)

dev.off()



par(mar = c(0, 0, 3, 0))

plotSPC(osds, id.style = 'top', width = 0.35, name.style = 'center-center', color = 'value', col.label = 'Munsell Value (moist)', cex.names = 0.66, cex.id = 0.5, shrink = TRUE)

plotSPC(osds, id.style = 'top', width = 0.35, name.style = 'center-center', color = 'hue', col.label = 'Munsell Value (moist)', cex.names = 0.66, cex.id = 0.5, shrink = TRUE)


huePosition(returnHues = TRUE)
huePositionCircle()

osds$hue_ordered <- factor(osds$hue, levels = huePosition(returnHues = TRUE))

plotSPC(osds, id.style = 'top', width = 0.35, name.style = 'center-center', color = 'hue_ordered', col.label = 'Munsell Value (moist)', cex.names = 0.66, cex.id = 0.5, shrink = TRUE)





par(mar = c(0, 0, 0, 2))

plotSPC(osds, width = 0.3, hz.distinctness.offset = 'hd', cex.id = 0.5, cex.names = 0.66, axis.line.offset = -1) 

plotSPC(osds, width = 0.3, hz.topography.offset = 'hzto', cex.id = 0.5, cex.names = 0.66, axis.line.offset = -1) 

plotSPC(osds, width = 0.3, hz.distinctness.offset = 'hd', hz.topography.offset = 'hzto', cex.id = 0.5, cex.names = 0.66, name = 'bnd.code', axis.line.offset = -1) 


plotSPC(osds, width = 0.3, hz.distinctness.offset = 'hd', hz.topography.offset = 'hzto', cex.id = 0.5, cex.names = 0.66, name = 'bnd.code', hz.boundary.lty = 'hzto.lty') 

legend('bottomleft', horiz = TRUE, legend = c('Smooth', 'Wavy', 'Irregular', 'Broken'), lty = 1:4, inset = 0.05, bty = 'n', cex = 0.85)



```
