---
title: "Soil Texture Vizualization Ideas"
author: "D.E. Beaudette"
date: "`r Sys.Date()`"
output:
  html_document:
    mathjax: null
    jquery: null
    smart: no
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, echo=FALSE, results='hide', warning=FALSE}
# setup
library(knitr, quietly = TRUE)
opts_chunk$set(message=FALSE, warning=FALSE, background='#F7F7F7', fig.retina=2, dev='png', tidy=FALSE, verbose=FALSE, cache = FALSE)
options(width=100, stringsAsFactors=FALSE, cache=FALSE)
```

# Examples with the Loafercreek sample data

Load required packages, sample data, and subset of `horizon` attributes.
```{r fig.height=7, fig.width=7}
# SoilProfileCollection methods
# you will need the latest version of aqp
# https://github.com/ncss-tech/aqp/
library(aqp)
# data-getting / sample data
library(soilDB)
# soil texture triangle
library(soiltexture)
# convenient alpha() function
library(scales)

# load sample data from soilDB package
data("loafercreek")

# extract just horizon data
h <- horizons(loafercreek)
```


Subset `sand`, `silt`, `clay` percentages and filter missing records or inconsistent fractions (e.g. add to more or less than 100%). Column names are adjusted to accommodate subsequent usage. 
```{r fig.height=7, fig.width=7}
# extract components of field texture, removing rows with missing data
ssc <- h[, c('genhz', 'sand', 'silt', 'clay')]
ssc <- na.omit(ssc)

# adjust names for plotting with TT.plot()
# names must be SAND, SILT, CLAY
names(ssc) <- toupper(names(ssc))

# test of bogus data
ssc$sum <- rowSums(ssc[, c('SAND', 'SILT', 'CLAY')])
# > 5% deviation from 100%
idx <- which(abs(ssc$sum - 100) > 5)

# check errors
ssc[idx, ]

# ommit errors
# maybe fix these later?
ssc <- ssc[-idx, ]

# check: OK
head(ssc)
```

```{r fig.height=3, fig.width=7}
# split by generalized horizon name into a list of data.frames
ssc.list <- split(ssc, ssc$GENHZ)

record.count <- rev(sapply(ssc.list, nrow))

par(mar = c(4.5, 3, 3, 1))
dotchart(record.count, pch = 21, bg = 'royalblue', pt.cex = 1.5, xlab = 'Number of records', main = 'Loafercreek\nRecords per Generalized Horizon Label')
```

# Visualization via `soiltexture` Package

An empty soil texture triangle, based particle size classes and abbreviations used by the US. Department of Agriculture, Natural Resources Conservation Service (USDA-NRCS) and National Cooperative Soil Survey (NCSS). This interface to ternary diagrams is provided by the excellent [`soiltexture`](https://CRAN.R-project.org/package=soiltexture) package by Julien Moeys.
```{r fig.height=7, fig.width=7}
TT.plot(
  class.sys = "USDA-NCSS.TT",    # use "our" texture triangle
  main = 'USDA-NRCS / NCSS',          # title
  cex.lab = 0.75,                 # scaling of label text
  cex.axis = 0.75,                # scaling of axis
  frame.bg.col = 'white',         # background color
  class.lab.col = 'black',        # color for texture class labels
  lwd.axis = 1.5,
  lwd.lab = 2,
  arrows.show=TRUE
)
```

Create a new figure, this  time saving the ternary diagram geometry for later (`TT` object). Data can be added as needed with the `TT.` family of functions.
```{r fig.height=7, fig.width=7}
# init figure and save geometry
TT <- soiltexture::TT.plot(
  class.sys = "USDA-NCSS.TT",    # use "our" texture triangle
  main = 'Loafercreek\nall horizons',          # title
  tri.sum.tst = FALSE,            # do not test for exact sum(sand, silt, clay) == 100
  cex.lab = 0.75,                 # scaling of label text
  cex.axis = 0.75,                # scaling of axis
  frame.bg.col = 'white',         # background color
  class.lab.col = 'black',        # color for texture class labels
  lwd.axis = 1.5,
  lwd.lab = 2,
  arrows.show=TRUE
)

# add data
# screen coordinates are available for later use
xy <- TT.points(tri.data = ssc, geo = TT, tri.sum.tst = FALSE, cex=0.5, col=alpha('royalblue', 0.5))

head(xy)
```


Subset data to `A` horizons, all with one call to `TT.plot`.
```{r fig.height=7, fig.width=7}
# local copy of `A` horizons
x <- ssc.list$A

# compose title using the current genhz
this.hz <- x$GENHZ[1]
title.text <- sprintf("Loafercreek: %s Horizons", this.hz)

TT.plot(
  class.sys= "USDA-NCSS.TT",    # use "our" texture triangle
  tri.data=x,                 # 
  main=title.text,    # title
  tri.sum.tst=FALSE,            # do not test for exact sum(sand, silt, clay) == 100
  cex.lab=0.75,                 # scaling of label text
  cex.axis=0.75,                # scaling of axis
  cex=0.5,                      # scaling of point symbols
  col=alpha('darkgreen', 0.5),  # color of point symbols, with 50% transparency
  frame.bg.col='white',         # background color
  class.lab.col='black',        # color for texture class labels
  lwd.axis=2                    # line thickness for axis
)
```


# Describing Central Tendency and Spread

A demonstration of [`textureTriangleSummary()`](http://ncss-tech.github.io/aqp/docs/reference/textureTriangleSummary.html), which displays approximate "ranges" (based on user-defined percentiles) in sand, silt, and clay percentages on the soil texture triangle in the form of a shaded polygon. The outline of the polygon is created by connecting (marginal) percentiles (typically 5th and 95th) of sand, silt, and clay fractions. The "Sample RV" is based on the (marginal) 50th percentiles (e.g. median sand, silt, and clay fractions).


```{r fig.height=7, fig.width=7}
# local copy of `A` horizons
x <- ssc.list$A

# visualize area bound by marginal percentiles of sand, silt, clay
stats <- textureTriangleSummary(
    x, pch = 1, cex = 0.5, 
    range.alpha = 50, col = grey(0.5), legend = FALSE
  )

# marginal percentiles
round(stats)
```


# Simulation
Simulation of sand, silt, clay as a **composition** via [`bootstrapSoilTexture()`](http://ncss-tech.github.io/aqp/docs/reference/bootstrapSoilTexture.html).

```{r fig.height=7, fig.width=7}
# simulate some data and try again
s <- bootstrapSoilTexture(x, n = 100, method = 'dirichlet')$samples

# make the figure, ignore results
textureTriangleSummary(
  s, pch = 1, cex = 0.5, 
  range.alpha = 75,
  col = grey(0.5), legend = FALSE,
  main = 'Dirichlet'
)

# get last triangle geometry
TT <- TT.geo.get()
TT.points(tri.data = x, geo = TT, tri.sum.tst = FALSE, cex=0.25, col='royalblue')

# simple legend
legend('top', 
       legend = c('Source', 'Simulated'), 
       pch = c(21, 1), 
       col = c('black', 'black'), 
       pt.bg = c('royalblue', NA), 
       horiz = TRUE, bty = 'n'
)
```

```{r fig.height=7, fig.width=7}
# simulate some data and try again
s <- bootstrapSoilTexture(x, n = 100, method = 'normal')$samples

# make the figure, ignore results
textureTriangleSummary(
  s, pch = 1, cex = 0.5, 
  range.alpha = 50, col = grey(0.5), legend = FALSE,
  main = 'Normal'
)

# get last triangle geometry
TT <- TT.geo.get()
TT.points(tri.data = x, geo = TT, tri.sum.tst = FALSE, cex=0.25, col='royalblue')

# simple legend
legend('top', 
       legend = c('Source', 'Simulated'), 
       pch = c(21, 1), 
       col = c('black', 'black'), 
       pt.bg = c('royalblue', NA), 
       horiz = TRUE, bty = 'n'
)
```

## Investigate Simulation Stability
To be continued.

```{r fig.height=7, fig.width=7}
# empty figure and geometry for later use
TT <- soiltexture::TT.plot(
  class.sys = "USDA-NCSS.TT",    # use "our" texture triangle
  main = 'Simulation Stability\nNormal Distribution',          # title
  tri.sum.tst = FALSE,            # do not test for exact sum(sand, silt, clay) == 100
  cex.lab = 0.75,                 # scaling of label text
  cex.axis = 0.75,                # scaling of axis
  frame.bg.col = 'white',         # background color
  class.lab.col = 'black',        # color for texture class labels
  lwd.axis = 1.5,
  lwd.lab = 2,
  arrows.show=TRUE
)

# 100 replications
s.n <- replicate(
  100, 
  bootstrapSoilTexture(x, n = 100, method = 'normal')$samples, simplify = FALSE
)

s.d <- replicate(
  100, 
  bootstrapSoilTexture(x, n = 100, method = 'dirichlet')$samples, simplify = FALSE
)

# internal function used to generate margin stats + bouding polygon
res.n <- lapply(s.n, function(i) {
  aqp:::.get.ssc.low.rv.high(ssc = i, p = c(0.05, 0.5, 0.95), delta = 1, TT.obj = TT)
})

res.d <- lapply(s.d, function(i) {
  aqp:::.get.ssc.low.rv.high(ssc = i, p = c(0.05, 0.5, 0.95), delta = 1, TT.obj = TT)
})


# depict denisty via overlay with transparency
trash <- sapply(res.n, function(i) {

  # extract marginal median
  .ssc <- as.data.frame(i$stats[2, 1:3, drop = FALSE])
  
  # add median as RV
  TT.points(tri.data = .ssc, geo = TT, tri.sum.tst = FALSE, cex=0.5, col = alpha('firebrick', 0.2))
  
  # add bounding polygon
  polygon(i$range$x, i$range$y, col=alpha('royalblue', 0.01), border = alpha('royalblue', 0.01))  
})


TT <- soiltexture::TT.plot(
  class.sys = "USDA-NCSS.TT",    # use "our" texture triangle
  main = 'Simulation Stability\nDirichlet Distribution',          # title
  tri.sum.tst = FALSE,            # do not test for exact sum(sand, silt, clay) == 100
  cex.lab = 0.75,                 # scaling of label text
  cex.axis = 0.75,                # scaling of axis
  frame.bg.col = 'white',         # background color
  class.lab.col = 'black',        # color for texture class labels
  lwd.axis = 1.5,
  lwd.lab = 2,
  arrows.show=TRUE
)

trash <- sapply(res.d, function(i) {

  # extract marginal median
  .ssc <- as.data.frame(i$stats[2, 1:3, drop = FALSE])
  
  # add median as RV
  TT.points(tri.data = .ssc, geo = TT, tri.sum.tst = FALSE, cex=0.5, col = alpha('firebrick', 0.2))
  
  # add bounding polygon
  polygon(i$range$x, i$range$y, col=alpha('royalblue', 0.01), border = alpha('royalblue', 0.01))  
})
```



----------------------------
This document is based on `aqp` version `r utils::packageDescription("aqp", field="Version")` and `soilDB` version `r utils::packageDescription("soilDB", field="Version")`.

