<!--
manually process like this:
library(knitr); library(markdown)
knit('vertical_vs_perpendicular.Rmd')
markdownToHTML('vertical_vs_perpendicular.md', 'vertical_vs_perpendicular.html')
-->

```{r setup, echo=FALSE, results='hide'}
opts_chunk$set(message=FALSE, warning=FALSE, background='#F7F7F7', dpi=100, fig.align='center', dev='CairoPNG', tidy=FALSE)
options(width=100, stringsAsFactors=FALSE)
```

### Introduction
This is a simple demonstration of how to do some soils-related stuff in **R** with the `aqp` package.


### A Visual Demonstration of Horizon Depths Measured Vertically vs. Perpendicular to the Surface Slope

The relationship between perpendicular horizon depth ($T_{p}$) and vertical horizon depth ($T_{v}$) is defined as: 

$$T_{p} = T_{v} cos(\theta)$$

where $\theta$ is the surface slope measured in radians. Surface slope measured in percent ($s$) can be converted to radians with $\theta = atan(s / 100)$.

### Setup environment by loading required packages, and user-defined functions
```{r setup-env}
# load required packages, you may have to install these if missing:
# install.packages('Hmisc', dep=TRUE)
library(lattice)
library(Hmisc)
library(Cairo)
library(aqp)
library(xtable)

# convert angle in percent to radians
pct2rad <- function(pct) {
	atan(pct / 100)
}

# compute difference in thickness as a function of slope pct
thickDiff <- function(pct, Tv) {
	Tp <- Tv * cos(pct2rad(pct))
	return(Tv - Tp)
}

# transform vertical to perpendicular depths
# operates on a single SPC object
VtoP <- function(i, which) {
	# get vector of bottom or top depths
	b <- horizons(i)[[which]]
	# convert vertical to perpendicular depths and round to an integer
	b.p <- round(b * cos(pct2rad(i$slope)))
	return(b.p)
}
```





### Generate a sample data set for subsequent demonstrations
```{r generate-pedons}
# generate a single profile, keep only the relevant stuff
set.seed(54321)
p <- random_profile(1)[, c('id', 'top', 'bottom', 'name')]

# stack 10 copies of these data
pp <- rbind(p, p, p, p, p, p, p, p, p, p)

# add a unique ID to each chunk copy
pp$id <- rep(1:10, each=nrow(p))

# upgrade our data.frame into a SoilProfileCollection object
depths(pp) <- id ~ top + bottom

# generate a sequence of (10) simulated slope values
s <- data.frame(id=1:10, slope=seq(from=1, to=70, length.out=10))
# assign these values into the 'site' attribute table of our pedons
site(pp) <- s

# assume data were generated with vertical horizon depths
# compute corresponding perpendicular horizon bottom depths
pp$Tp <- profileApply(pp, VtoP, which='bottom')

# check our new object:
print(pp)
```

### Plot differences in horizon depth as a function of slope
```{r plot-pedons-with-arrows, fig.width=8, fig.height=4, out.width='100%'}
# adjust figure margins
par(mar=c(2,0,0.75,0))
# plot SoilProfileCollection
plot(pp, n.depth.ticks=10)

# add a title
title('Vertical Horizon Depths (black) vs. Perpendicular Horizon Depths (blue)', cex.main=0.75)

# add axis with slope values
axis(side=1, cex.axis=0.75, line=-1.25, at=1:10, labels=round(s$slope))
mtext(side=1, 'Slope (%)', line=0.5, cex=0.75)

# iterate over profiles and add offset hz boundaries
for(i in 1:length(pp)) {
	pp.i <- pp[i, ]
	# use color to show only those offsets that are > 1 cm
	cols <- ifelse(abs(pp.i$bottom - pp.i$Tp) > 1, 'RoyalBlue', NA)
	arrows(x0=i, x1=i, y0=pp.i$bottom, y1=pp.i$Tp, length=0.08, col=cols)
	segments(x0=i-0.2, x1=i+0.2, y0=pp.i$Tp, y1=pp.i$Tp, col=cols, lty=2)
}
```

### This time plot after converting horizon depths from vertical basis to perpendicular basis
```{r change-depths, fig.width=8, fig.height=4, out.width='100%'}
# make a copy of the original profiles
pp.perp <- pp

# convert horizon top and bottom depths from vertical to perpendicular basis
pp.perp$top <- profileApply(pp.perp, VtoP, which='top')
pp.perp$bottom <- profileApply(pp.perp, VtoP, which='bottom')

# setup margins
par(mar=c(2,0,0.75,0))
# plot profiles
plot(pp.perp, n.depth.ticks=10)
# add title
title('Changes in Horizon Depth as a Function of Surface Slope', cex.main=0.75)
# add slope axis
axis(side=1, cex.axis=0.75, line=-1.25, at=1:10, labels=round(s$slope))
mtext(side=1, 'Slope (%)', line=0.5, cex=0.75)
# add horizontal lines to assist with perception of depth
abline(h=seq(from=5, to=75, by=5), lty=3, col='grey')
```

### Tabulate the mean and cumulative difference in horizon depths by pedon
```{r make-depth-diff-table, results='asis'}
# compute difference in horizon thickness
diff.thick <- (pp$bottom - pp$top) - (pp.perp$bottom - pp.perp$top)
theID <- horizons(pp)$id # extract IDs
# compute mean and total difference by pedon
diff.by.pedon <- by(diff.thick, theID, function(i) cbind(mean.diff.cm=mean(i), total.diff.cm=sum(i)))
# composite into a data.frame and add slope values
diff.by.pedon.df <- data.frame(do.call('rbind', diff.by.pedon), slope=round(s$slope))
# convert tabular output into HTML
print(xtable(diff.by.pedon.df), type='html')
```

### Another visualization of differences in vertical vs. perpendicular depths as a function of slope
```{r diff-vs-slope-figure-setup}
# make a grid of horizon thickness and slope percents
d <- expand.grid(thick=c(5,10,20,50,100), slope=1:70)

# compute the difference between vertical and perpendicular depths
d$thickDiff <- thickDiff(d$slope, d$thick)

# check the results
head(d)
```

```{r plot-diff-by-slope, fig.width=10, fig.height=5, out.width='100%'}
# plot these data
# note the different syntax-- this is lattice-style plotting
xYplot(thickDiff ~ slope, data=d, groups=factor(thick, labels=paste(c(5,10,20,50,100), 'cm depth')), type='l',
ylab='(Vertical - Perpendicular) Hz Thickness', xlab='Slope (%)', 
main='Difference in Apparent Horizon Thickness as a Function of Slope', 
scales=list(alternating=3, tick.number=10), 
par.settings=list(superpose.line=list(col='black', lwd=1.5)), 
label.curves=list(cex=1), 
panel=function(...) {
	panel.abline(v=seq(0, 70, by=10), h=seq(0, 20, by=2), col=grey(0.75), lty=3)
	panel.xYplot(...)
	})
```
