---
title: "Theoretical Water Retention Curves"
author: "D.E. Beaudette"
date: "`r Sys.Date()`"
output:
  html_document:
    mathjax: null
    jquery: null
    smart: no
    number_sections: no
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: false
---
  
```{r setup, echo=FALSE, results='hide', warning=FALSE}
# setup
library(knitr, quietly=TRUE)
opts_chunk$set(message=FALSE, warning=FALSE, background='#F7F7F7', fig.align='center', dev='svglite', tidy=FALSE, verbose=FALSE)
options(width=100, stringsAsFactors=FALSE)
```

# Setup
```{r}
library(aqp)
library(soilDB)

library(dendextend)
library(cluster)
library(ape)

library(latticeExtra)
library(grid)
library(tactile)
library(viridis)
```

# Theoretical Water Retention Curves
Average soil hydraulic parameters by USDA soil texture class, extracted from the [ROSETTA documentation](https://www.ars.usda.gov/pacific-west-area/riverside-ca/agricultural-water-efficiency-and-salinity-research-unit/docs/model/rosetta-class-average-hydraulic-parameters/). Approximate water-retention data have been computed for an idealized volume of homogenous soil material via van Genuchten model.

```{r, echo=FALSE}
data("ROSETTA.centroids")
kableExtra::kable_styling(knitr::kable(ROSETTA.centroids, row.names = FALSE, digits = 3), full_width = FALSE, font_size = 10)
```


Develop theoretical water retention curves via van Genuchten model and texture class centroids.
```{r}
# load average soil hydraulic parameters
data("ROSETTA.centroids")

# re-level texture class by approximate AWC
ll <- ROSETTA.centroids$texture[order(ROSETTA.centroids$awc)]
ROSETTA.centroids$texture <- factor(ROSETTA.centroids$texture, levels=ll)

# iterate over horizons and generate VG model curve
res <- lapply(1:nrow(ROSETTA.centroids), function(i) {
  # fit model using parameters from centroids
  # model bounds are given in kPA of suction
  vg <- KSSL_VG_model(VG_params = ROSETTA.centroids[i, ], phi_min = 10^-3, phi_max=10^6)
  
  # extract curve and add texture ID
  m <- vg$VG_curve
  m$texture <- ROSETTA.centroids$texture[i]
  
  return(m)
})

# copy over lab sample number as ID
res <- do.call('rbind', res)

# check: OK
head(res)
```

```{r fig.width = 9, fig.height = 5}
tps <- tactile.theme(superpose.line=list(col=viridis::viridis(12), lwd=2))

xyplot(
  phi ~ theta, groups=texture, data=res, 
  type=c('l', 'g'), 
  scales=list(alternating=3, x=list(tick.number=10), y=list(log=10, tick.number=10)), 
  yscale.components=yscale.components.logpower, 
  ylab=expression(Suction~~(kPa)), 
  xlab=expression(Volumetric~Water~Content~~(cm^3/cm^3)), 
  par.settings = tps, 
  strip=strip.custom(bg=grey(0.85)), 
  auto.key=list(columns=4, lines=TRUE, points=FALSE, cex=0.8)
)
```


Plot water retention curves and annotate with permanent wilting point (approximately 1500kPa suction), field capacity (approximately 33kPa suction), and saturation (close to 0 kPa suction). Note that matric potential (typically negative pressure by convention) are displayed as suction (positive values).
```{r fig.width=10, fig.height=8}

trellis.par.set(superpose.symbol=list(col=c('black', 'black', 'black'), fill=c('firebrick', 'orange', 'royalblue'), cex=1, pch=22))

sk <- simpleKey(text=c('permanent wilting point', 'field capacity', 'saturation'), points = TRUE, columns=3)

xyplot(
  phi ~ theta | texture, data=res, 
  type=c('l', 'g'), 
  scales=list(alternating=3, x=list(tick.number=6), y=list(log=10, tick.number=6)), 
  yscale.components=yscale.components.logpower, 
  ylab=expression(Matric~~Potential~~(-kPa)), 
  xlab=expression(Volumetric~Water~Content~~(cm^3/cm^3)), 
  par.settings = tactile.theme(plot.line=list(col='black', lwd=2)), 
  strip=strip.custom(bg=grey(0.85)), 
  as.table=TRUE,
  key=sk,
  main='Idealized Water Retention\nUSDA-ARS ROSETTA Model Centroids', 
  # sub='van Genuchten model',
  subscripts = TRUE,
  panel= function(x, y, subscripts=subscripts, ...) {
    # idealized water retention curve
    panel.xyplot(x, y, ...)
    
    # get texture (i.e. the current panel)
    tx <- res$texture[subscripts][1]
    # look-up centroid data for this texture
    centroid.data <- ROSETTA.centroids[match(tx, ROSETTA.centroids$texture), ]
    
    # add PWP, note that y-scale is log_10-transformed
    panel.points(x=centroid.data$pwp, y=log(1500, base = 10), col='black', fill='firebrick', cex=0.85, pch=22)
    # add PWP, note that y-scale is log_10-transformed
    panel.points(x=centroid.data$fc, y=log(33, base = 10), col='black', fill='orange', cex=0.85, pch=22)
    # add SAT, note that y-scale is log_10-transformed
    panel.points(x=centroid.data$sat, y=log(0.1, base = 10), col='black', fill='royalblue', cex=0.85, pch=22)
    
    # annotate with approx AWC
    grid.text(sprintf('AWC: %s', round(centroid.data$awc, 3)), x = unit(0.95, 'npc'), y = unit(0.95, 'npc'), gp=gpar(cex=0.75), hjust=1)
    
  }
)
```


# Soil Texture Class

Align (order) soil texture classes based on (average) estimated water retention parameters.
```{r fig.width = 5, fig.height = 4}
data("ROSETTA.centroids")

# local copy
x <- ROSETTA.centroids

# cluster based on parameters
m <- ROSETTA.centroids[, c('sat', 'fc', 'pwp')]
row.names(m) <- x$texture
h <- as.hclust(agnes(daisy(m)))

# attemp to align with total AWC
h <- dendextend::rotate(h, order(x$awc))
h2 <- as.phylo(h)

# check: OK!
par(mar = c(0, 0, 0, 0))
plot(h2, label.offset = 0.005)
h$labels[h$order]
```

Explore several sorting strategies.
```{r}
# clustering approach
x$texture_sort1 <- factor(x$texture, levels = rev(h$labels[h$order]))

# just AWC
x$texture_sort2 <- factor(x$texture, levels = x$texture[order(x$awc)])
```

Develop figures.
```{r fig.width = 8, fig.height = 5}
p.1 <- segplot(texture ~ pwp + fc, data=x, 
        horizontal = TRUE, 
        level = awc, col.regions=viridis,
        main='Available Water Holding Capacity\nUSDA-ARS ROSETTA Model Centroids', 
        sub='Sorted According to Field Book v3.0',
        xlab=expression(Volumetric~Water~Content~~(cm^3/cm^3)),
        par.settings = tactile.theme()
)

p.2 <- segplot(texture_sort1 ~ pwp + fc, data=x, 
        horizontal = TRUE, level = awc, col.regions=viridis,
        main='Available Water Holding Capacity\nUSDA-ARS ROSETTA Model Centroids',
        sub='Sorted According to {PWP, FC, SAT} Values',
        xlab=expression(Volumetric~Water~Content~~(cm^3/cm^3)),
        par.settings = tactile.theme()
)

p.3 <- segplot(texture_sort2 ~ pwp + fc, data=x, horizontal = TRUE, 
        level = awc, col.regions=viridis,
        main='Available Water Holding Capacity\nUSDA-ARS ROSETTA Model Centroids',
        sub='Sorted According to AWC',
        xlab=expression(Volumetric~Water~Content~~(cm^3/cm^3)),
        par.settings = tactile.theme()
        )
```

```{r fig.width = 8, fig.height = 5, echo = FALSE}
print(p.1)
print(p.2)
print(p.3)
```






----------------------------
This document is based on `aqp` version `r utils::packageDescription("aqp", field="Version")`.


