---
title: "Exploring Geomorphic Summaries"
author: "D.E. Beaudette"
date: "`r Sys.Date()`"
output:
  html_document:
    mathjax: null
    jquery: null
    smart: no
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: false
---


```{r setup, echo=FALSE, results='hide', warning=FALSE}
# setup
library(knitr, quietly=TRUE)
library(kableExtra, quietly=TRUE)
opts_chunk$set(message=FALSE, warning=FALSE, background='#F7F7F7', fig.align='center', fig.retina=2, dev='png', tidy=FALSE, verbose=FALSE)
options(width=100, stringsAsFactors=FALSE)
```

# Setup
```{r install-deps, eval=FALSE}
# stable packages from CRAN
install.packages('aqp')
install.packages('soilDB')
install.packages('sharpshootR')
install.packages('FactoMineR')
install.packages('cowplot')

# latest versions from GitHub
remotes::install_github("ncss-tech/aqp", dependencies=FALSE, upgrade_dependencies=FALSE)
remotes::install_github("ncss-tech/soilDB", dependencies=FALSE, upgrade_dependencies=FALSE)
remotes::install_github("ncss-tech/sharpshootR", dependencies=FALSE, upgrade_dependencies=FALSE)
```



# Hillslope Position / Soil Morphology
```{r eval=TRUE}
library(aqp)
library(soilDB)
library(sharpshootR)

# soil of interest
# https://casoilresource.lawr.ucdavis.edu/gmap/?loc=44.10219,-122.38589,z14
s <- 'peavine'

# all siblings
sib <- siblings(s)

# all soils of interest
soils <- c(s, sib$sib$sibling)

# get morphology + extended summaries
osds <- fetchOSD(soils, extended = TRUE)
```

```{r fig.width=10, fig.height=6}
res <- vizHillslopePosition(osds$hillpos)

# re-order hillslope proportions according to clustering
hp <- osds$hillpos[res$order, ]
nm <- names(hp[, 2:6])
```

```{r echo = FALSE}
kable_styling(kable(osds$hillpos[res$order, ], digits = 2, row.names = FALSE, format = 'html'), font_size = 10, full_width = FALSE)
```

```{r fig.width=10, fig.height=6}
print(res$fig)
```

```{r fig.width=12, fig.height=6}
par(mar=c(0,0,1,1))
SoilTaxonomyDendrogram(osds$SPC, scaling.factor = 0.015, width = 0.28, name.style = 'center-center')
```

```{r fig.width=6, fig.height=1.5}
# colors
par(mar = c(1, 1, 1, 1))
hp.cols <- RColorBrewer::brewer.pal(n = 5, name = 'Set1')[c(2, 3, 4, 5, 1)]
soilPalette(hp.cols, lab = nm)
```

## Composite Figures

### Version 1: Lines
```{r fig.width=12, fig.height=8}
par(mar = c(0.5, 0, 0, 2))
layout(matrix(c(1,2)), widths = c(1,1), heights = c(2,1))
plotProfileDendrogram(osds$SPC, res$clust, dend.y.scale = 3, scaling.factor = 0.012, y.offset = 0.2, width = 0.32, name.style = 'center-center', cex.names = 0.7, shrink = TRUE, cex.id = 0.55)

matplot(y = hp[, 2:6], type = 'b', lty = 1, pch = 16, axes = FALSE, col = hp.cols, xlab = '', ylab = '', xlim = c(0.5, length(osds$SPC) + 1))
# grid(nx = 0, ny = NULL)
axis(side = 4, line = -1, las = 1, cex.axis = 0.7)
# axis(side = 2, line = -3, las = 1, cex.axis = 0.7)
legend('topleft', legend = rev(nm), col = rev(hp.cols), pch = 16, bty = 'n', cex = 0.8, pt.cex = 2, horiz = TRUE, inset = c(0.01, 0.01))
mtext('Probability', side = 2, line = -2, font = 2)
```


### Version 2: Stacked Bars
```{r fig.width=10, fig.height=8}
par(mar = c(0.5, 0, 0, 2))
layout(matrix(c(1,2)), widths = c(1,1), heights = c(2,1))
plotProfileDendrogram(osds$SPC, res$clust, dend.y.scale = 3, scaling.factor = 0.012, y.offset = 0.2, width = 0.32, name.style = 'center-center', cex.names = 0.7, shrink = TRUE, cex.id = 0.55)

sp <- c(1.5, rep(1, times = length(osds$SPC) - 1))
barplot(height = t(as.matrix(hp[, 2:6])), beside = FALSE, width = 0.5, space = sp, col = hp.cols,  axes = FALSE, xlab = '', ylab = '', xlim = c(0.5, length(osds$SPC) + 1), ylim = c(0, 1.2))

legend(x = 0.75, y = 1.2, legend = rev(nm), col = rev(hp.cols), pch = 15, bty = 'n', cex = 0.8, pt.cex = 1.25, horiz = TRUE)
mtext('Probability', side = 2, line = -2, font = 2)

```



# Correspondance Analysis
```{r}
library(aqp)
library(soilDB)
library(sharpshootR)

library(FactoMineR)
library(cowplot)

# a classic catena
soils <- c('cecil', 'altavista', 'lloyd', 'wickham', 'wilkes',  'chewacla', 'congaree')
s <- fetchOSD(soils, extended = TRUE)

# create visualizations
hs <- vizHillslopePosition(s$hillpos)
gc <- vizGeomorphicComponent(s$geomcomp)
```

```{r fig.width=10, fig.height=8}
# montage figures
plot_grid(hs$fig, gc$fig, nrow = 2, rel_heights = c(0.5, 0.5))
```

## CA from Soil Series Geomorphic Summaries
Data are from SSURGO components.
```{r fig.width=8, fig.height=8}
hs.tab <- (s$hillpos[, 2:6])
row.names(hs.tab) <- s$hillpos$series

gc.tab <- (s$geomcomp[, 2:7])
row.names(gc.tab) <- s$geomcomp$series

# perform CA
hs.ca <- CA(hs.tab, graph = FALSE)
gc.ca <- CA(gc.tab, graph = FALSE)
```

```{r fig.width=8, fig.height=8}
# plot
plot(hs.ca, autoLab='yes', title='Hillslope Position', cex=0.75, col.col='firebrick', col.row='royalblue')
plot(gc.ca, autoLab='yes', title='Geomorphic Component', cex=0.75, col.col='firebrick', col.row='royalblue')
```

```{r fig.width=10, fig.height=10}
## combine lattice + ggplot graphics

p2 <- plot(hs.ca, autoLab='yes', title='Hillslope Position', cex=0.75, col.col='firebrick', col.row='royalblue')

plot_grid(hs$fig, p2, nrow = 2, rel_heights = c(0.4, 0.6))

```



# Next Steps
Work with source data, cross-tabulate 2D / 3D proportions when possible. Not possible with flattened summaries delivered by `fetchOSD`.
```{r echo = FALSE, eval=FALSE}

hs.tab <- (s$hillpos[, 2:7])
gc.tab <- (s$geomcomp[, 2:8])


hs.tab <- round(sweep(hs.tab[, -6], MARGIN = 1, STATS = hs.tab$n, FUN = '*'))
hs.tab$series <- s$hillpos$series

gc.tab <- round(sweep(gc.tab[, -7], MARGIN = 1, STATS = gc.tab$n, FUN = '*'))
gc.tab$series <- s$geomcomp$series

# g <- merge(hs.tab, gc.tab, by = 'series', sort = FALSE)


hs.counts <- lapply(split(hs.tab, hs.tab$series), function(i) {
  i.s <- i$series
  i$series <- NULL
  u <- rep(names(i), times = as.vector(unlist(i)))
  data.frame(series = i.s, u)
})

gc.counts <- lapply(split(gc.tab, gc.tab$series), function(i) {
  i.s <- i$series
  i$series <- NULL
  u <- rep(names(i), times = as.vector(unlist(i)))
  data.frame(series = i.s, u)
})


# this won't work... we need paired observations
tab <- table(hs.counts$ALTAVISTA$u, gc.counts$ALTAVISTA$u)

# perform CA as series at a time
ca <- CA(tab, graph = FALSE)
```


----------------------------
This document is based on `aqp` version `r utils::packageDescription("aqp", field="Version")`, `soilDB` version `r utils::packageDescription("soilDB", field="Version")`, and `sharpshootR` version `r utils::packageDescription("sharpshootR", field="Version")`.
