---
title: "Exploring Geomorphic Summaries"
author: "D.E. Beaudette"
date: "`r Sys.Date()`"
output:
  html_document:
    mathjax: null
    jquery: null
    smart: no
    number_sections: no
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: false
---


```{r setup, echo=FALSE, results='hide', warning=FALSE}
# setup
library(knitr, quietly=TRUE)
library(kableExtra, quietly=TRUE)
opts_chunk$set(message=FALSE, warning=FALSE, background='#F7F7F7', fig.align='center', fig.retina = 1, dev='png', tidy=FALSE, verbose=FALSE)
options(width=100, stringsAsFactors=FALSE)
```

# Setup
```{r install-deps, eval=FALSE}
# stable packages from CRAN
install.packages('aqp')
install.packages('soilDB')
install.packages('sharpshootR')
install.packages('FactoMineR')
install.packages('cowplot')

# latest versions from GitHub
remotes::install_github("ncss-tech/aqp", dependencies=FALSE, upgrade_dependencies=FALSE)
remotes::install_github("ncss-tech/soilDB", dependencies=FALSE, upgrade_dependencies=FALSE)
remotes::install_github("ncss-tech/sharpshootR", dependencies=FALSE, upgrade_dependencies=FALSE)
```



# Hillslope Position / Soil Morphology
```{r eval=TRUE}
library(aqp)
library(soilDB)
library(sharpshootR)

# soil of interest
# https://casoilresource.lawr.ucdavis.edu/gmap/?loc=44.10219,-122.38589,z14
s <- 'peavine'
# s <- 'clarion'

# all siblings
sib <- siblings(s)

# all soils of interest
soils <- c(s, sib$sib$sibling)

# get morphology + extended summaries
osds <- fetchOSD(soils, extended = TRUE)
```



There may be missing series from either / both SPC and geomorphic summaries. TBC.
```{r eval = FALSE}
# there may be records missing from SPC / hill slope position
nm <- intersect(profile_id(osds$SPC), osds$hillpos$series)

# keep only those series that exist in both
sub <- subset(osds$SPC, profile_id(osds$SPC) %in% nm)

## inverse problem: extra records in hill slope summaries
# subset hillpos
hillpos.sub <- subset(osds$hillpos, subset = series %in% profile_id(sub))

## TODO: use subset objects below
```


```{r fig.width=10, fig.height=6}
res <- vizHillslopePosition(osds$hillpos)

# re-order hillslope proportions according to clustering
hp <- osds$hillpos[res$order, ]
nm <- names(hp[, 2:6])
```

```{r echo = FALSE}
kable_styling(kable(osds$hillpos[res$order, ], digits = 2, row.names = FALSE, format = 'html'), font_size = 10, full_width = FALSE)
```

```{r fig.width=10, fig.height=6}
print(res$fig)
```

```{r fig.width=12, fig.height=6}
par(mar=c(0,0,1,1))
SoilTaxonomyDendrogram(osds$SPC, scaling.factor = 0.015, width = 0.28, name.style = 'center-center')
```

```{r fig.width=6, fig.height=1.5}
# colors
par(mar = c(1, 1, 1, 1))
hp.cols <- RColorBrewer::brewer.pal(n = 5, name = 'Set1')[c(2, 3, 4, 5, 1)]
soilPalette(hp.cols, lab = nm)
```

## Composite Figures

### Version 1: Lines
```{r fig.width=12, fig.height=8}
par(mar = c(0.5, 0, 0, 2))
layout(matrix(c(1,2)), widths = c(1,1), heights = c(2,1))
plotProfileDendrogram(osds$SPC, res$clust, dend.y.scale = 3, scaling.factor = 0.012, y.offset = 0.2, width = 0.32, name.style = 'center-center', cex.names = 0.7, shrink = TRUE, cex.id = 0.55)

matplot(y = hp[, 2:6], type = 'b', lty = 1, pch = 16, axes = FALSE, col = hp.cols, xlab = '', ylab = '', xlim = c(0.5, length(osds$SPC) + 1))
# grid(nx = 0, ny = NULL)
axis(side = 4, line = -1, las = 1, cex.axis = 0.7)
# axis(side = 2, line = -3, las = 1, cex.axis = 0.7)
legend('topleft', legend = rev(nm), col = rev(hp.cols), pch = 16, bty = 'n', cex = 0.8, pt.cex = 2, horiz = TRUE, inset = c(0.01, 0.01))
mtext('Probability', side = 2, line = -2, font = 2)
```


### Version 2: Stacked Bars
```{r fig.width=10, fig.height=8}
par(mar = c(0.5, 0, 0, 2))
layout(matrix(c(1,2)), widths = c(1,1), heights = c(2,1))
plotProfileDendrogram(osds$SPC, res$clust, dend.y.scale = 3, scaling.factor = 0.012, y.offset = 0.2, width = 0.32, name.style = 'center-center', cex.names = 0.7, shrink = TRUE, cex.id = 0.55)

sp <- c(1.5, rep(1, times = length(osds$SPC) - 1))
barplot(height = t(as.matrix(hp[, 2:6])), beside = FALSE, width = 0.5, space = sp, col = hp.cols,  axes = FALSE, xlab = '', ylab = '', xlim = c(0.5, length(osds$SPC) + 1), ylim = c(0, 1.2))

legend(x = 0.75, y = 1.2, legend = rev(nm), col = rev(hp.cols), pch = 15, bty = 'n', cex = 0.8, pt.cex = 1.25, horiz = TRUE)
mtext('Probability', side = 2, line = -2, font = 2)

```


# 3D Surface Shape
```{r fig.width=10, fig.height=6}
res <- vizSurfaceShape(osds$shape_across, title = 'Surface Shape (Across)')

# re-order proportions according to clustering
ss <- osds$shape_across[res$order, ]
nm <- names(ss[, 2:6])
```

```{r echo = FALSE}
kable_styling(kable(osds$shape_across[res$order, ], digits = 2, row.names = FALSE, format = 'html'), font_size = 10, full_width = FALSE)
```

```{r fig.width=10, fig.height=6}
print(res$fig)
```

```{r fig.width=12, fig.height=6}
par(mar = c(0.5, 0, 1.5, 2))
plotProfileDendrogram(osds$SPC, res$clust, dend.y.scale = 3, scaling.factor = 0.012, y.offset = 0.2, width = 0.32, name.style = 'center-center', cex.names = 0.7, shrink = TRUE, cex.id = 0.55)
title('Surface Shape (Across)')
```




# Correspondance Analysis
```{r}
library(aqp)
library(soilDB)
library(sharpshootR)

library(FactoMineR)
library(cowplot)

# a classic catena
soils <- c('cecil', 'altavista', 'lloyd', 'wickham', 'wilkes',  'chewacla', 'congaree')
s <- fetchOSD(soils, extended = TRUE)

# create visualizations
hs <- vizHillslopePosition(s$hillpos)
gc <- vizGeomorphicComponent(s$geomcomp)
```

```{r fig.width=10, fig.height=8}
# montage figures
plot_grid(hs$fig, gc$fig, nrow = 2, rel_heights = c(0.5, 0.5))
```

## CA from Soil Series Geomorphic Summaries
Data are from SSURGO components.
```{r fig.width=8, fig.height=8}
# row-wise proportions add to 1
hs.tab <- (s$hillpos[, 2:6])
row.names(hs.tab) <- s$hillpos$series

# row-wise proportions add to 1
gc.tab <- (s$geomcomp[, 2:7])
row.names(gc.tab) <- s$geomcomp$series

## TODO: think about implications:
# * row-wise proportions ignore relationships to other rows
# * table-wise proportions assume equal weights
# * converting to row-wise counts would over-weight series of great extent

# convert row-wise proportions -> table-wise proportions
hs.tab <- hs.tab / sum(hs.tab)
gc.tab <- gc.tab / sum(gc.tab)

# perform CA
hs.ca <- CA(hs.tab, graph = FALSE)
gc.ca <- CA(gc.tab, graph = FALSE)
```

```{r fig.width=8, fig.height=8}
# plot
plot(hs.ca, autoLab='yes', title='Hillslope Position', cex=0.75, col.col='firebrick', col.row='royalblue')
plot(gc.ca, autoLab='yes', title='Geomorphic Component', cex=0.75, col.col='firebrick', col.row='royalblue')
```

```{r fig.width=10, fig.height=10}
## combine lattice + ggplot graphics
p2 <- plot(hs.ca, autoLab='yes', title='Hillslope Position', cex=0.75, col.col='firebrick', col.row='royalblue')

plot_grid(hs$fig, p2, nrow = 2, rel_heights = c(0.4, 0.6))
```



# Multiple Factor Analysis
The results are interesting, but contaminated by missing data: ALTAVISTA and CHEWACLA have < 4 records in the hills geomorphic component summary. These two soils are better described in the terrace summary.

https://journal.r-project.org/archive/2013/RJ-2013-003/RJ-2013-003.pdf
```{r fig.width=10, fig.height=10}

# prepare 2D table
hs.tab <- (s$hillpos[, 1:6])
# prepare 3D table
gc.tab <- (s$geomcomp[, 1:7])

# inner join
# this is the column-wise concatenation of two cross-tabulations
g <- merge(hs.tab, gc.tab, by = 'series', sort = FALSE)
row.names(g) <- g$series
g$series <- NULL


# multiple factor analysis
m <- MFA(g, group = c(5, 6), type = c('f', 'f'), name.group = c('2D', '3D'), graph = FALSE)

plot(m, choix = 'freq', palette = c('black', 'firebrick', 'royalblue'))
```


# Combined 3D Geomorphic Descriptions
```{r fig.width=8, fig.height=8}

sib <- siblings('drummer', only.major = FALSE)
soils <- c('drummer', sib$sib$sibling)

s <- fetchOSD(soils, extended = TRUE)

# TODO: conditionally convert to counts and join if nrow > 1

hl <- data.frame(series = s$geomcomp$series, s$geomcomp[, 2:7] * s$geomcomp$n)

tr <- data.frame(series = s$terrace$series, s$terrace[, 2:3] * s$terrace$n)

fl <- data.frame(series = s$flats$series, s$flats[, 2:3] * s$flats$n)

g <- merge(hl, tr, by = 'series', sort = FALSE, all.x = TRUE, all.y = TRUE)
g <- merge(g, fl, by = 'series', sort = FALSE, all.x = TRUE, all.y = TRUE)

m <- as.matrix(g[, -1])
m[is.na(m)] <- 0

m <- sweep(m, MARGIN = 1, STATS = rowSums(m), FUN = '/')

g <- data.frame(series = g$series, m)

tab <- g[, -1]
row.names(tab) <- g$series

g.ca <- CA(tab, graph = FALSE)


plot(g.ca, autoLab = 'yes', title = 'Combined 3D Geomorphic Description', cex = 0.75, col.col = 'firebrick', col.row = 'royalblue')
```

```{r}
summary(g.ca)

dimdesc(g.ca)

knitr::kable(g, digits = 2)
```

# relate entropy to CA ordination
```{r eval=FALSE}
g$H <- apply(g[, -1], 1, aqp::shannonEntropy)
```


----------------------------
This document is based on `aqp` version `r utils::packageDescription("aqp", field="Version")`, `soilDB` version `r utils::packageDescription("soilDB", field="Version")`, and `sharpshootR` version `r utils::packageDescription("sharpshootR", field="Version")`.
