---
title: "Getting and Comparing KSSL Data"
author: "D.E. Beaudette"
date: "`r Sys.Date()`"
output:
  html_document:
    mathjax: null
    jquery: null
    smart: no
---

```{r setup, echo=FALSE, results='hide', warning=FALSE}
# setup
library(knitr, quietly=TRUE)
opts_chunk$set(message=FALSE, warning=FALSE, background='#F7F7F7', fig.align='center', fig.retina=2, dev='png', tidy=FALSE, verbose=FALSE)
options(width=100, stringsAsFactors=FALSE)
```

# Introduction
This document demonstrates how to use the [soilDB](http://ncss-tech.github.io/AQP/soilDB/soilDB-Intro.html) package to download [KSSL](http://ncsslabdatamart.sc.egov.usda.gov/) data from SoilWeb. These data are from the [September 2018 snapshot](https://github.com/dylanbeaudette/process-kssl-snapshot), and will be updated as future snapshots are released. Comparisons are made via graphical summaries of key soil properties with depth, using [data structures and functions](hhttp://ncss-tech.github.io/AQP/aqp/aqp-intro.html) from the [aqp](http://ncss-tech.github.io/AQP/) package.

## Installation
With a recent version of R (>= 3.5), it is possible to get all of the packages that this tutorial depends on via:
```{r install-deps, eval=FALSE}
# stable packages from CRAN
install.packages('maps', dep=TRUE)
install.packages('aqp', dep=TRUE)
install.packages('soilDB', dep=TRUE)

# latest versions from GitHub
remotes::install_github("ncss-tech/aqp", dependencies=FALSE, upgrade_dependencies=FALSE)
remotes::install_github("ncss-tech/soilDB", dependencies=FALSE, upgrade_dependencies=FALSE)
```


# Quick Example: getting lab characterization (KSSL) and basic morphology (NASIS)
KSSL and NASIS data are from the September 2018 snapshot. Details pending.
```{r}
library(aqp)
library(soilDB)

# get lab and morphologic data
# simplify colors to 1 color (moist and dry) per horizon via mixing
# see ?mix_and_clean_colors for details.
s <- fetchKSSL(series='auburn', returnMorphologicData = TRUE, simplifyColors = TRUE)

# extract pedons into SoilProfileCollection
pedons <- s$SPC
```


```{r}
# the result is a list, check it out
str(s, 2)

# check out the "raw" morphologic data:
lapply(s$morph, head)
```


```{r fig.width=10, fig.height=6}
# check
par(mar=c(0,0,0,0))
plotSPC(pedons, color='moist_soil_color', print.id=FALSE, name='hzn_desgn')
```


Simplification of fragment data is still a manual process.
```{r fig.width=10, fig.height=6}
# simplify fragment data
s.frags <- simplifyFragmentData(s$morph$phfrags, id.var='labsampnum')

# merge fragment data into SPC
h <- horizons(pedons)
h <- merge(h, s.frags, by='labsampnum', all.x=TRUE, sort=FALSE)
horizons(pedons) <- h


# check
par(mar=c(0,0,3,0))
plotSPC(pedons, color='total_frags_pct', print.id=FALSE, name='hzn_desgn')
addVolumeFraction(pedons, 'total_frags_pct', pch=1, cex.min=0.25, cex.max = 0.5)
```

## Fragment Volume >= 100%
There are cases where errors in the source data (NASIS `Pedon Horizon Fragments` table) suggest coarse fragment volume > 100%. As of `soilDB` 2.0.2 and `aqp` 1.15.4, the `simplifyFragmentData()` function will report such errors and the `addVolumeFraction()` function will fail gracefully by truncating fragment volume at 100%.

Give it a try with an example (thanks to Brian Gardner for finding it).
```{r fig.width=6, fig.height=4, warning=TRUE}
# get lab and morphologic data
s <- fetchKSSL(series='kettenbach', returnMorphologicData = TRUE)

# extract pedons into SoilProfileCollection
pedons <- s$SPC

# simplify fragment data
s.frags <- simplifyFragmentData(s$morph$phfrags, id.var='labsampnum')

# merge fragment data into SPC
h <- horizons(pedons)
h <- merge(h, s.frags, by='labsampnum', all.x=TRUE, sort=FALSE)
horizons(pedons) <- h

# flag horizons with >= 100% total fragment volume
pedons$fragment_errors <- factor(pedons$total_frags_pct >= 100)

# label pedons with user pedon ID
# label horizons with total fragment %
par(mar=c(0,0,3,5))
plotSPC(pedons, color='fragment_errors', name='total_frags_pct', label='pedon_id', cex.name=0.85)
addVolumeFraction(pedons, 'total_frags_pct', pch=1, cex.min=0.25, cex.max = 0.5)
```


Check the original fragment data, based on the warnings issued by `simplifyFragmentData()`. Looks like multiple versions of the same data were accidentally entered. Errors like this can be corrected in NASIS, but may require assistance from the regional staff depending on KSSL pedon ownership. `fetchKSSL()` is based on a snapshot of NASIS and LIMS data, therefore corrections may not be available until after the next release cycle.
```{r}
# identify and print affected rows
idx <- which(s$morph$phfrags$labsampnum %in% c('82P02200', '82P02204'))
s$morph$phfrags[idx, ]
```

# Fetch Data Associated with a Set of IDs
As of `soilDB` version 2.5.3, `fetchKSSL()` is fully vectorized with the exception of `bbox`. Note that it is much more efficient to work with the KSSL snapshot when attempting to load large collections of data by ID.

```{r, results='hide'}
# pedons indexed by "pedlabsampnum" ID
pls <- c("04N0610", "04N0611", "04N0612", "04N0613")

# vectorization = implicit iteration
res <- fetchKSSL(pedlabsampnum = pls, returnMorphologicData = TRUE, simplifyColors = TRUE)

# check: OK
str(res, 2)

# extract SPC
pedons <- res$SPC

# check: looks good
plotSPC(pedons, color='estimated_ph_h2o', name='hzn_desgn')

# check: looks good
plotSPC(pedons, color='moist_soil_color', name='hzn_desgn')
```


# Example Application
Data can be queried by "taxonname"" (typically a soil series), using a geographic bounding-box in WGS84 referenced coordinates, or by MLRA. Taxonname matching is case-insensitive and based on the most current taxonname in NASIS. See `?fetchKSSL()` or [this GitHub repository](https://github.com/dylanbeaudette/process-kssl-snapshot) for details.

In this example, we are downloading KSSL data for the [musick](https://soilseriessc.egov.usda.gov/OSD_Docs/M/MUSIC.html), [chaix](https://soilseries.sc.egov.usda.gov/OSD_Docs/C/CHAIX.html), and [holland](https://soilseries.sc.egov.usda.gov/OSD_Docs/H/HOLLAND.html) soils; commonly found on granitic rocks of the Sierra Nevada Mountain region of California.
```{r fig.width=6, fig.height=7}
# load libraries
library(aqp)
library(soilDB)
library(lattice)
library(maps)

# fetch KSSL data by series name
sn <- c('musick', 'holland', 'chaix')
g <- fetchKSSL(series = sn, progress = FALSE)

# compute weighted-mean particle diameter for later
g$wmpd <- with(horizons(g), ((vcs * 1.5) + (cs * 0.75) + (ms * 0.375) + (fs * 0.175) + (vfs *0.075) + (silt * 0.026) + (clay * 0.0015)) / (vcs + cs + ms + fs + vfs + silt + clay))

# estimate soil depth based on horizon designations
sdc <- getSoilDepthClass(g, name='hzn_desgn', top='hzn_top', bottom='hzn_bot')

# splice-into site data
site(g) <- sdc

# summarize soil depth by taxonname
tapply(g$depth, g$taxonname, summary)
```


Check for variation in `taxonname` and normalize accordingly. In this case we can just convert to all lower-case.
```{r fig.width=6, fig.height=7}
# check
table(g$taxonname)

# normalize via lower-case
g$taxonname <- tolower(g$taxonname)

# OK
table(g$taxonname)
```


Sometimes it is useful to split a collection into individual objects. Do this now by `filter`-ing on taxonname.
```{r fig.width=6, fig.height=7}
musick <- filter(g, taxonname == 'musick')
holland <- filter(g, taxonname == 'holland')
chaix <- filter(g, taxonname == 'chaix')
```



```{r fig.width=6, fig.height=7}
# generate a basemap of northern California, with county boundaries
map('county', 'California', xlim=c(-123.25, -118.75), ylim=c(36.5, 41))
# add long/lat axes
map.axes()
# add locations of musick
points(y ~ x, data=site(musick), pch=21, bg='RoyalBlue')
# add locations of holland
points(y ~ x, data=site(holland), pch=21, bg='DarkRed')
# add locations of chaix
points(y ~ x, data=site(chaix), pch=21, bg='DarkGreen')
# add a simple legend
legend('topright', pch=21, pt.bg=c('RoyalBlue', 'DarkRed', 'DarkGreen'), legend=c('musick', 'holland', 'chaix'), bty='n')
```




## Profile Plots
Generate "thematic" profile sketches by coloring horizons according to clay content, and gouped by taxonname
```{r fig.width=15, fig.height=6}
par(mar=c(0,0,3,1))

# convert taxonname to a factor for grouping
g$taxonname <- factor(g$taxonname)

# suppress printing IDs and horizon designations
groupedProfilePlot(g, groups = 'taxonname', color='clay', print.id=FALSE, name=NA)
groupedProfilePlot(g, groups = 'taxonname', color='estimated_ph_h2o', print.id=FALSE, name=NA)
groupedProfilePlot(g, groups = 'taxonname', color='wmpd', print.id=FALSE, name=NA)
```



## Aggregation
The `slab()` function is used to aggregate selected variables within collections of soil profiles along depth-slices. In this case, we aggregate clay, pH (1:1 H2O, estimated from saturated paste pH when missing), and base saturation at pH 8.2 along 1-cm thick slices and within groups defined by the variable 'taxonname'. See `?slab()` for details on how this function can be used.
```{r aggregate}
g.slab <- slab(g, taxonname ~ clay + estimated_ph_h2o + bs82 + wmpd)

# inspect stacked data structure
str(g.slab)
```


## Refine Labels
It is convenient to know how many pedons there are within each collection--therefore, we append this value to the series name. Using the same approach, we can rename the soil properties with more useful descriptions and units of measure.
```{r rename}
# re-name soils with series name + number of pedons-- order is critical !
new.levels <- c('musick', 'holland', 'chaix')
new.labels <- paste(new.levels, ' [', c(length(musick), length(holland), length(chaix)), ' pedons]', sep='')
g.slab$taxonname <- factor(g.slab$taxonname, levels=new.levels, labels=new.labels)

# new names should match the order in:
levels(g.slab$variable)

# re-name soil property labels-- order is critical !
levels(g.slab$variable) <- c('Clay (%)', 'pH 1:1 H2O', 'Base Saturation at pH 8.2 (%)', 'WMPD (mm)')
```


## Graphically Compare
The slice-wise median and 25th/75th percentiles are reasonable estimations of central tendency and spread. "Contributing fraction" values (% of pedons with data at a given depth) are printed along the right-hand side of each panel. These values provide both an indication of how deep the soils are, and, how much confidence can be placed in the aggregate data at any given depth.
```{r plot, fig.width=10, fig.height=6}
# define plotting style
tps <- list(superpose.line=list(col=c('RoyalBlue', 'DarkRed', 'DarkGreen'), lwd=2))

# plot grouped, aggregate data
xyplot(top ~ p.q50 | variable, groups=taxonname, data=g.slab, ylab='Depth',
       xlab='median bounded by 25th and 75th percentiles',
       lower=g.slab$p.q25, upper=g.slab$p.q75, ylim=c(155,-5),
       panel=panel.depth_function, alpha=0.25, sync.colors=TRUE,
       prepanel=prepanel.depth_function,
       cf=g.slab$contributing_fraction,
       par.strip.text=list(cex=0.8),
       strip=strip.custom(bg=grey(0.85)),
       layout=c(4,1), scales=list(x=list(alternating=1, relation='free'), y=list(alternating=3)),
       par.settings=tps,
       auto.key=list(columns=3, lines=TRUE, points=FALSE)
)
```





----------------------------
This document is based on `aqp` version `r utils::packageDescription("aqp", field="Version")` and `soilDB` version `r utils::packageDescription("soilDB", field="Version")`.
