---
title: "Simulating a Mixture of Soil Colors"
author: "D.E. Beaudette"
date: "`r Sys.Date()`"
output:
  html_document:
    mathjax: null
    jquery: null
    smart: no
---
  
```{r setup, echo=FALSE, results='hide', warning=FALSE}
library(knitr, quietly=TRUE)
opts_chunk$set(message=FALSE, warning=FALSE, background='#F7F7F7', fig.retina=2, dev='png', tidy=FALSE, verbose=FALSE)
options(width=100, stringsAsFactors=FALSE)
```

# Background

The following code is only a *rough estimation* of mixing soil material and not a physically-based model of subtractive mixing of pigments. [We will get there eventually](https://github.com/ncss-tech/aqp/issues/10).


# Example

Required packages, be sure to ge the [latest versions from GitHub](https://github.com/ncss-tech/aqp).
```{r}
library(aqp)
library(soilDB)
```


Pick some colors and assign weights:

<span style="padding: 5px; background-color: `r parseMunsell('10YR 3/5')`; color: white;">`60% 10YR 3/5`</span> + <span style="padding: 5px; background-color: `r parseMunsell('5YR 5/6')`; color: white;">`30% 5YR 5/6`</span> + <span style="padding: 5px; background-color: `r parseMunsell('2.5Y 5/6')`; color: white;">`10% 2.5Y 5/6`</span> = <b>?</b>.

The `parseMunsell` function is used 3 times to: 

 1. split Munsell notation into hue/value/chroma components
 2. convert to sRGB / CIELAB coordinates
 3. convert to sRGB hex notation.
```{r}
colors <- c('10YR 3/5', '5YR 5/6', '2.5Y 5/6')
weights <- c(60, 30, 10)

# combine into a data.frame and convert to sRGB + CIE LAB
d <- cbind(
  parseMunsell(colors, convertColors=FALSE),
  parseMunsell(colors, return_triplets=TRUE, returnLAB=TRUE),
  pct=weights,
  col=parseMunsell(colors, convertColors=TRUE)
)

d
```


Estimate mixture via weighted average in CIELAB color space. This is a nearly-perceptual color system so the weighted average is a *reasonable* estimate. Again, this is not as accurate as subtractive mixing of pigments via spectral data.
```{r}
(mx <- estimateColorMixture(d, wt = 'pct', backTransform = TRUE))
mx.munsell <- sprintf("%s %s/%s", mx$colorhue, mx$colorvalue, mx$colorchroma)
```

<span style="padding: 5px; background-color: `r parseMunsell('10YR 3/5')`; color: white;">`60% 10YR 3/5`</span> + <span style="padding: 5px; background-color: `r parseMunsell('5YR 5/6')`; color: white;">`30% 5YR 5/6`</span> + <span style="padding: 5px; background-color: `r parseMunsell('2.5Y 5/6')`; color: white;">`10% 2.5Y 5/6`</span> = <span style="padding: 5px; background-color: `r parseMunsell(mx.munsell)`; color: white;">`r mx.munsell`</span>.


## Visualization Ideas
Demonstrate mixing soil colors in CIE LAB color spaces via `estimateColorMixture()`.
```{r, fig.width=7, fig.height=2.5}
mixColorDemo <- function(x) {
  x <- x[order(x$pct, decreasing = TRUE), ]
  n.cols <- nrow(x)
  x.vals <- 1:n.cols
  y.vals <- rep(0, times=n.cols)
  cols <- rgb(x[, c('r', 'g', 'b')])
  col.txt <- paste0(x$hue, ' ', x$value , '/', x$chroma, '\n(', x$pct, '%)')
  
  plot(x.vals, y.vals, col=cols, pch=15, cex=10, axes=FALSE, ylab='', xlab='', xlim=c(0.5, n.cols + 2.5))
  text(x.vals[-length(x.vals)] + 0.5, y.vals, '+', cex=2, lwd=2)
  text(max(x.vals) + 0.5, y.vals[1], '=', cex=2, lwd=2)
  
  text(x.vals, y.vals, col.txt, pos=1, offset=2.5, cex=0.75)
  
  # mix in CIELAB
  mixed.lab <- estimateColorMixture(x, wt = 'pct', backTransform = TRUE)
  
  # add mixed colors
  points(max(x.vals)+1, y.vals[1], col=rgb(mixed.lab[, c('r', 'g', 'b')]), cex=10, pch=15)
  col.txt <- paste0(mixed.lab$colorhue, ' ', mixed.lab$colorvalue , '/', mixed.lab$colorchroma)
  text(max(x.vals)+1, y.vals[1], col.txt, pos=1, offset=2.5, cex=0.75)
}


par(mar=c(1,1,1,1))
mixColorDemo(d)
```

### Alternative Representation
```{r, fig.width=9, fig.height=4}
# simulate mixture
mixed.lab <- estimateColorMixture(d, wt = 'pct', backTransform = TRUE)

# color vector
cols <- c(
  d$col, 
  rgb(mixed.lab[, c('r', 'g', 'b')])
  )

# label vector
labs <- c(
  colors,
  paste0(mixed.lab$colorhue, ' ', mixed.lab$colorvalue , '/', mixed.lab$colorchroma)
)

# add weights
labs <- sprintf("%s      (%s%%)", labs, c(weights, 100))

# plot
soilPalette(
  colors = cols,
  lab = labs,
  lab.cex = 1
  )
```


# Mixing via Reference Spectra
```{r fig.width=9, fig.height=7}
colors <- c('10YR 3/5', '5YR 5/6', '2.5Y 5/6')
weights <- c(60, 30, 10)

plotColorMixture(x = colors, w = weights)
```


# Mix a Lot

This is very inefficient code.
```{r fig.width=12, fig.height=4}
s <- seq(from = 0, to = 100, by = 10)
colors <- c('10YR 3/3', '2.5Y 6/4')

z <- lapply(s, function(i) {
  
  wts <- c(100 - i, i)
  
  d <- cbind(
  parseMunsell(colors, convertColors=FALSE),
  parseMunsell(colors, return_triplets=TRUE, returnLAB=TRUE),
  pct = wts,
  col = parseMunsell(colors, convertColors=TRUE)
)
  
  mx <- estimateColorMixture(d, wt = 'pct', backTransform = TRUE)
  
  res <- data.frame(
    m.x = colors[1],
    m.y = colors[2],
    col.x = parseMunsell(colors[1]),
    col.y = parseMunsell(colors[2]),
    wt.x = wts[1],
    wt.y = wts[2],
    mixed.chip = sprintf('%s %s/%s', mx$colorhue, mx$colorvalue, mx$colorchroma),
    mixed.col = rgb(mx$r, mx$g, mx$b),
    stringsAsFactors = FALSE
  )
  
  res
  
  return(res)
})

z <- do.call('rbind', z)


par(mar = c(1,1,1,1))
plot(1:nrow(z), rep(1:nrow(z)), xlim = c(0.25, nrow(z) + 0.25), ylim = c(0, 0.8), axes = FALSE, xlab = '', ylab = '', type = 'n')

rect(xleft = (1:nrow(z)) - 0.45, xright = (1:nrow(z)) + 0.45, ybottom = 0.25, ytop = 0.75, col = z$mixed.col, border = 'black', lwd = 2, lend = 2)

text(1:nrow(z), 0.2, z$mixed.chip)
text(1:nrow(z), 0.15, sprintf('%s/%s', z$wt.x, z$wt.y))
```

```{r fig.width=12, fig.height=4}
colorContrastPlot(z$m.y, z$mixed.chip)
colorContrastPlot(z$m.x, z$mixed.chip)
```

# Mix via Munsell Spectra
Hmmm, those colors look a little strange to me.

```{r fig.width=12, fig.height=4}
s <- seq(from = 0, to = 100, by = 10)
colors <- c('10YR 3/3', '2.5Y 6/4')

z <- lapply(s, function(i) {
  
  wts <- c(100 - i, i)
  mx <- mixMunsell(colors, w = wts)
  
  d <- data.frame(
    m.x = colors[1],
    m.y = colors[2],
    col.x = parseMunsell(colors[1]),
    col.y = parseMunsell(colors[2]),
    wt.x = wts[1],
    wt.y = wts[2],
    mixed.chip = mx$munsell,
    mixed.col = parseMunsell(mx$munsell),
    dist = mx$distance,
    stringsAsFactors = FALSE
  )
  
  return(d)
})

z <- do.call('rbind', z)

par(mar = c(1,1,1,1))
plot(1:nrow(z), rep(1:nrow(z)), xlim = c(0.25, nrow(z) + 0.25), ylim = c(0, 0.8), axes = FALSE, xlab = '', ylab = '', type = 'n')

rect(xleft = (1:nrow(z)) - 0.45, xright = (1:nrow(z)) + 0.45, ybottom = 0.25, ytop = 0.75, col = z$mixed.col, border = 'black', lwd = 2, lend = 2)

text(1:nrow(z), 0.2, z$mixed.chip)
text(1:nrow(z), 0.15, sprintf('%s/%s', z$wt.x, z$wt.y))

text(1:nrow(z), 0.5, round(z$dist, 3), cex = 0.75, col = invertLabelColor(z$mixed.col))
```

```{r fig.width=12, fig.height=4}
colorContrastPlot(z$m.y, z$mixed.chip)
colorContrastPlot(z$m.x, z$mixed.chip)
```


----------------------------
This document is based on `aqp` version `r utils::packageDescription("aqp", field="Version")` and `soilDB` version `r utils::packageDescription("soilDB", field="Version")`.

