---
title: "Getting Data from the Rapid Carbon Assessment Project"
author: "D.E. Beaudette and S. Wills"
date: "`r Sys.Date()`"
output:
  html_document:
    mathjax: null
    jquery: null
    smart: no
---
  
```{r setup, echo=FALSE, results='hide', warning=FALSE}
library(knitr, quietly=TRUE)
opts_chunk$set(message=FALSE, warning=FALSE, background='#F7F7F7', fig.retina=2, dev='png', tidy=FALSE, verbose=FALSE)
options(width=100, stringsAsFactors=FALSE)
```


# Introduction
This document demonstrates how to use the [soilDB](http://ncss-tech.github.io/AQP/soilDB/soilDB-Intro.html) package to download [RaCA](http://soils.usda.gov/survey/raca/index.html) via SoilWeb. Comparisons are made via graphical summaries of key soil properties with depth, using [data structures and functions](http://ncss-tech.github.io/AQP/aqp/aqp-intro.html) from the [aqp](http://ncss-tech.github.io/AQP/) package.


## The SoilWeb RaCA Query API

RaCA data are queried via HTTP request to the SoilWeb server, with query parameters encoded as [URL parameters](http://en.wikipedia.org/wiki/Query_string). For example, by appending several parameters to the base URL it is possible to query "site-level" data associated with soils correlated to the Auburn series like this:

```
http://casoilresource.lawr.ucdavis.edu/soil_web/rca/rca_query.php?what=site&series=auburn
```

There are 5 URL parameters that can be used to refine RaCA data queries:

  * what=**site** ------*query only the site-level data*
  
  * series=**auburn** ------*return only those data correlated to the Auburn series*
  
  * bbox=**-120,37,-122,38** ------*return only those data within the specified bounding box, constrained to a 5-degree block*
  
  * state=**ca** ------*return only those data within California*
  
  * plaintext=**1** ------*return data as plain text instead of [gzipped](http://en.wikipedia.org/wiki/Gzip) text*


The **what** parameter determines which chunk of data will be returned:
  * **site**: site and pedon level data
  * **horizon**: horizon level data
  * **veg**: vegetation information recorded at each site (relative abundence of 6 most common species)
  * **trees**: tree height(feet) and [DBH](http://en.wikipedia.org/wiki/Diameter_at_breast_height)(inches) measured on the first for trees [in](ftp://ftp-fc.sc.egov.usda.gov/NSSC/raca/RaCA_Field_Collection_Protocols.pdf) using RaCA procdures
  * **stock**: cumulative soil organic carbon stocks (Mg per ha) to standardized depths
  * **sample**: measured/modeled bulk density and carbon concentration at the sample-level
  * **vnir**: VNIR spectra [queries by soil series or BBOX only]


# Getting RaCA Data in R
The [soilDB](http://cran.at.r-project.org/package=soilDB) package abstracts URL-based queries for RaCA data into a single function, `fetchRaCA()`. Calling this function with user-defined query paramters, e.g. `fetchRaCA(series='auburn')`, will download all data (site, horizon, veg, trees, stock) associated with the Auburn series and return the results as a `list`. VNIR spectra can be very large and are therefore not downloaded by default.


## Installation
With a recent version of R (>= 2.15), it is possible to get all of the packages that this tutorial depends on via:
```{r install-deps, eval=FALSE}
# run these commands in the R console
install.packages('soilDB', dep=TRUE)
```

## Example Usage
Data can be queried by two-letter state code, fuzzy-matching of series name, or using a geographic bounding-box in WGS84 referenced coordinates. See `?fetchRaCA()` for details.
```{r get-data, fig.width=6, fig.height=6, results='hide', eval=TRUE}
# load libraries
library(soilDB)
require(aqp)
library(plyr)
library(maps)
library(lattice)
library(reshape2)
library(cluster)

# define a geographic bounding box
bbox <- c(-120, 37, -122, 38)

# fetch data by soil series name
r.series <- fetchRaCA(series='auburn')

# fetch data by bounding box
r.bbox <- fetchRaCA(bbox=bbox)

# fetch data by state
r.state <- fetchRaCA(state='ca')

# map each set of fetched data
par(mar=c(0,0,0,0))
map('county', 'california')
points(y ~ x, data=site(r.state$pedons), col='RoyalBlue', cex=0.5, pch=1)
points(y ~ x, data=site(r.bbox$pedons), col='DarkGreen', cex=0.8, pch=15)
points(y ~ x, data=site(r.series$pedons), col='DarkRed', cex=0.8, pch=16)
# add bounding box
rect(bbox[1], bbox[2], bbox[3], bbox[4])
# add long/lat axes
map.axes()
# add a simple legend
legend('topright', pch=c(1, 15, 16), col=c('RoyalBlue', 'DarkGreen', 'DarkRed'), legend=c('state', 'bbox', 'series'), bty='n', title='query by:')
```


## Exploring the Results
A list of objects and tables (`data.frame` objects) is returned by `fetchRaCA()`. Elements can be accessed using the `$` operator.
```{r explore-results}
# check structure
str(r.bbox, 2)
```


## Site/Pedon Data
Site and pedon data are returned as a [SoilProfileCollection object](https://r-forge.r-project.org/scm/viewvc.php/*checkout*/docs/aqp/aqp-intro.html?root=aqp) within the `pedons` list element.
```{r site-pedon, fig.width=8, fig.height=4, results='hide', eval=TRUE}
# extract site/pedon data
s <- r.series$pedons

# how many pedons in the collection?
length(s)

# how many pedons / RCA site?
table(s$rcasiteid)

# generate soil colors from moist sRGB
s$soil_color <- rep(NA, times=nrow(horizons(s)))
idx <- complete.cases(s$m_r)
s$soil_color[idx] <- with(horizons(s)[idx, ], rgb(m_r, m_g, m_b)) # moist colors

# plot first 20 pedons
par(mar=c(0,0,0,0))
plot(s[1:20, ], name='hzname', max.depth=100)
```


## Vegetation Data
```{r veg-trees}
# tabulate vegetation data across the entire collection of RaCA data queried by BBOX
sort(table(r.bbox$veg$plantnatvernm), decreasing=TRUE)

# aggregate tree information, by RaCA site
ddply(r.bbox$trees, 'siteiid', summarize, mean.dbh=mean(dbh), mean.height=mean(height), n=length(siteiid))
```


## Soil Organic Carbon Concentrations
```{r carbon-conc-raw-data, fig.width=8, fig.height=5, eval=TRUE}
# extract sample data horizon-level carbon and bulk density
s <- r.series$sample

# init SoilProfileCollection
depths(s) <- rcapid ~ sample_top + sample_bottom

# slice-wise aggregation over entire collection
a <- slab(s, ~ soc)

# plot
xyplot(
  top ~ p.q50, data=a, main='Auburn', ylab='Depth (cm)', xlab='Soil Organic Carbon Concentration (%)',
  lower=a$p.q5, upper=a$p.q95, sync.colors=TRUE, alpha=0.3,
  ylim=c(60,-5), asp=1.5,
  par.settings=list(superpose.line=list(lwd=2, col=c('RoyalBlue'))),
  panel=panel.depth_function, 
  prepanel=prepanel.depth_function,
  auto.key=list(columns=2, lines=TRUE, points=FALSE)
)
```


## Soil Organic Carbon Stock Data
```{r carbon-stock, fig.width=8, fig.height=5, eval=TRUE}
# join carbon stock estimates to SPC object
# these are site-level data
s <- r.bbox$pedons
site(s) <- r.bbox$stock

# compute mean SOC stock to 100cm by RCA site:
round(tapply(s$value_100cm, s$rcasiteid, mean))

# compute mean SOC stock to 100cm by series:
round(tapply(s$value_100cm, s$taxonname, mean))

# convert taxonname to a factor
s$taxonname <- factor(s$taxonname)

# re-level based on median C stock
stock.asc.order <- order(tapply(s$value_100cm, s$taxonname, median))
new.levels <- levels(s$taxonname)[stock.asc.order]
s$taxonname <- factor(s$taxonname, levels=new.levels)

# visualize
bwplot(taxonname ~ value_100cm, data=site(s), xlab='Carbon Stock to 100cm (Mg per ha)')
```


## Depth-Wise Aggregation of Carbon Stock Data
```{r stock-depth-aggregation, fig.width=7, fig.height=6, eval=TRUE}
m <- melt(r.series$stock, id.vars='rcapid', measure.vars=c(2,4,6,8,10,12))

# fix c stock variable name
names(m)[3] <- 'c.stock'

# generate depths from variable names
m$top <- NA
m$bottom <- NA

# sequence of possible top/bottom boundaries
m.top <- c(0,5,10,20,30,50)
m.bottom <- c(5,10,20,30,50,100)

# generate an index to real depths
# this only works because the factor levels are in the correct order
mm <- as.numeric(m$variable)

# assign real depths
m$top <- m.top[mm]
m$bottom <- m.bottom[mm]

# promote to SPC
depths(m) <- rcapid ~ top + bottom

# simple aggregation of the entire collection: note that this doesn't take into account error estimates...
a <- slab(m, ~ c.stock)

# plot aggregate data
xyplot(
  top ~ p.q50, data=a, main='Auburn', ylab='Depth (cm)', xlab='Cumulative Carbon Stock (Mg per ha)',
  lower=a$p.q5, upper=a$p.q95, sync.colors=TRUE, alpha=0.3,
  ylim=c(105,-5), asp=1.5,
  par.settings=list(superpose.line=list(lwd=2, col=c('RoyalBlue'))),
  panel=panel.depth_function, 
  prepanel=prepanel.depth_function,
  auto.key=list(columns=2, lines=TRUE, points=FALSE)
)
```



## VNIR Spectra
VNIR spectra are large, and therefore not downloaded by default. The `get.vnir=TRUE` argument to `fetchRaCA()` will download VNIR spectra associated with the query results.
```{r vnir, fig.width=8, fig.height=4, results='hide', eval=TRUE}
s <- fetchRaCA(series='auburn', get.vnir=TRUE)

# plot the spectra
par(mar=c(0,0,0,0))
matplot(t(s$spectra), type='l', lty=1, col=rgb(0, 0, 0, alpha=0.25), ylab='', xlab='', axes=FALSE)
box()

# Q-mode PCA of spectra
pca <- prcomp(s$spectra, center=TRUE, scale=TRUE)
summary(pca)

# cluster on first 3 principal components into 3 hard groupings via PAM
spectra.pam <- pam(predict(pca)[, 1:3], 3)

# re-plot spectra, colored by 3-class clustering
cols <- rgb(t(col2rgb(spectra.pam$clustering)), maxColorValue=255, alpha=100)
matplot(t(s$spectra), type='l', lty=1, col=cols, ylab='', xlab='', axes=FALSE)
box()
```


## Querying and loading spectra
This example demonstrates how to load RaCA data for a regional area, subset it to specific soil taxa based on pattern matching, and load the spectra for a specific set of taxa.  
```{r vnir1, fig.width=8, fig.height=4, results='hide', eval=TRUE}

#load data by state without spectra
r.state <- fetchRaCA(state='id')

# subset and get rcasiteid's for sites with volcanic ash, andic and vitrandic subgroups
idx <- grep('andic', r.state$pedons$tax_subgroup)
length(idx)
print(table(r.state$pedons$tax_subgroup[idx]))
print(table(r.state$pedons$rcasiteid[idx]))

rcaIDs <- r.state$pedons$rcasiteid[idx]
rcIDs <- paste(rcaIDs, sep=', ')
#r.state <- fetchRaCA(rcasiteid = rcIDs)

# load spectra for specific RaCA siteIDs for the subset above
s <- fetchRaCA(rcasiteid = rcIDs, get.vnir=TRUE)

```

## Working with VNIR spectra as SpectralDataFrames
A very useful construct called a `SpectralDataFrame` and associated operations can be found in the [spectacles](https://github.com/pierreroudier/spectacles) package.  We will promote the downloaded RaCA spectra to a SpectralDataFrame by joining in associated data and attribute variables.  Once promoted to a SpectralDataFrame they can be split by horizon designation name and then aggregated to a mean spectra for each set of similarly designated horizons. Applying a logical genetic horizonation order (from soil surface to subsurface) to the aggregated spectra makes it easier to compare how the spectra vary with depth. 
```{r vnir2, fig.width=8, fig.height=4, results='hide', eval=TRUE}
library(ggplot2)
library(spectacles)
library(RColorBrewer)

# convert spectra matrix to dataframe
d <- as.data.frame(s$spectra)

# assign rownames to id column
d$id <- rownames(d)

# create Spectra from the dataframe
spectra(d) <- id ~ 350:2500

# subset some data elements from the sample dataframe
s1 <- s$sample[ , c('sample_id','rcapid','hzname','sample_top','sample_bottom')]
rownames(s1) <- s1$sample_id
names(s1) <- c('id', 'rcapid', 'hzname','sample_top','sample_bottom')

# promote Spectra to a SpectralDataFrame object by joining in the associated data
features(d, key='id') <- s1

# split data into lists on horizon names (hzname)
d1 <- split(d, d$hzname)
print(table(d$hzname))
# plot subset of 'A' horizons
plot(d1$A)

# aggregate spectra by taking the mean values across spectra within each hzname group
d_ag <- aggregate_spectra(d, fun = mean, id = "hzname")

# melt format for plotting with ggplot
d2 <- melt_spectra(d_ag, attr = c('hzname','sample_top'))

# set a logical horizon order for plotting
hz_ordered <- c('Oi', 'Oe', 'A', 'Bw1', 'Bw2', '2Bt', '2Bw','2Bw2', '2Bw3', '2Bw4', '2BC', '2C', '2C1', '2C2')
d2$hzname <- factor(d2$hzname, levels = hz_ordered)

# plot
ggplot(d2) + geom_line(aes(x = wl, y = nir, colour = hzname)) 
```

----------------------------
This document is based on `aqp` version `r utils::packageDescription("aqp", field="Version")` and `soilDB` version `r utils::packageDescription("soilDB", field="Version")`.