---
title: "ROSETTA Model API"
author: "Dylan Beaudette, Richard Reid, Todd Skaggs"
date: "`r Sys.Date()`"
output:
  html_document:
    mathjax: null
    jquery: null
    smart: no
    number_sections: no
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: false
---
  
```{r setup, echo=FALSE, results='hide', warning=FALSE}
library(knitr, quietly=TRUE)
opts_chunk$set(message=FALSE, warning=FALSE, background='#F7F7F7', fig.retina=2, dev='png', tidy=FALSE, verbose=FALSE)
options(width=100, stringsAsFactors=FALSE)
```



# Introduction
This document outlines how to access a new, online, interfae to the ROSETTA model for predicting hydraulic parameters from soil properties. The ROSETTA API was developed by Dr. Todd Skaggs (USDA-ARS) and links to the work of Zhang and Schaap, (2017). The `ROSETTA` function in **soilDB** is one possible approach to using the API from R.

The ROSETTA model relies on a minimum of 3 soil properties, with increasing (expected) accuracy as additional properties are included:

  * required, sand, silt, clay: USDA soil texture separates (percentages) that sum to 100%

  * optional, bulk density (any moisture basis): mass per volume after accounting for >2mm fragments, units of gm/cm3

  * optional, volumetric water content at 33 kPa: roughly "field capacity" for most soils, units of cm^3/cm^3

  * optional, volumetric water content at 1500 kPa: roughly "permanent wilting point" for most plants, units of cm^3/cm^3

Soil properties must be described, in order, via `vars` argument. The API does not use the names but column ordering must follow: sand, silt, clay, bulk density, volumetric water content at 33kPa (1/3 bar), and volumetric water content at 1500kPa (15 bar).


Consider using the [interactive version](https://www.handbook60.org/rosetta), with copy/paste functionality.

## Related Documentation
  
  * [`ROSETTA()` function manual page](https://ncss-tech.github.io/soilDB/docs/reference/ROSETTA.html)
  * [Water Retention Data from KSSL](https://ncss-tech.github.io/AQP/soilDB/fetchKSSL-VG-demo.html)
  * [SCAN/SNOTEL](https://ncss-tech.github.io/AQP/soilDB/fetchSCAN-demo.html)
  * [Water Retention Curve Experiments](https://ncss-tech.github.io/AQP/aqp/water-retention-curves.html)
  * [Avergage ROSETTA Predictions by Soil Texture Class](https://ncss-tech.github.io/aqp/docs/reference/ROSETTA.centroids.html)


# Setup
Install all of the packages that this tutorial depends on via:
```{r eval=FALSE}
# run these commands in the R console
install.packages('httr', dependencies =  TRUE)
install.packages('jsonlite', dependencies =  TRUE)
install.packages('latticeExtra', dependencies = TRUE)
install.packages('tactile', dependencies =  TRUE)
install.packages('soilDB', dependencies = TRUE)
install.packages('aqp', dependencies = TRUE)
```

You will also need the latest version of `soilDB`:
```{r eval=FALSE}
install.packages('remotes', dep=TRUE)
remotes::install_github("ncss-tech/soilDB", dependencies=FALSE, upgrade_dependencies=FALSE)
```


# A Very Basic Example

The following demonstrates how to push soil property data from SDA through the ROSETTA API.
```{r}
# required libraries
library(soilDB)
library(httr)
library(jsonlite)

# SQL to be submitted to SDA
q <- "SELECT
-- contextual data
compname, comppct_r,
-- horizon morphology
hzname, hzdept_r, hzdepb_r, 
-- parameters used by ROSETTA, may contain NULL
sandtotal_r, silttotal_r, claytotal_r, dbthirdbar_r, 
wthirdbar_r/100 AS wthirdbar_decimal, wfifteenbar_r/100 AS wfifteenbar_decimal
-- tables of interest
FROM 
mapunit AS mu
-- implicit filtering 
INNER JOIN component AS co ON mu.mukey = co.mukey
INNER JOIN chorizon AS ch ON co.cokey = ch.cokey
-- single map unit
WHERE mu.mukey = '373596'
-- order for visual inspection
ORDER BY co.cokey, ch.hzdept_r ASC;"

# submit query to SDA
x <- SDA_query(q)

# attempting to use all possible soil properties
vars <- c('sandtotal_r', 'silttotal_r', 'claytotal_r', 'dbthirdbar_r', 'wthirdbar_decimal', 'wfifteenbar_decimal')

# call ROSETTA API 
# automatic, best-available model type (m = "0")
r <- ROSETTA(x, vars = vars, m = "0")
```

```{r echo = FALSE}
# check results
kableExtra::kable_styling(knitr::kable(r, row.names = FALSE, digits = 3), full_width = FALSE, font_size = 10)
```


# Theoretical Water Retention Curves
Develop theoretical water retention curves (van Genuchten model) from soil texture class centroids.
```{r}
# required libraries
library(aqp)
library(latticeExtra)
library(tactile)

# make a table of soil textures
# sand sand/silt/clay values at geometric centroids
tex <- SoilTextureLevels()
x <- texcl_to_ssc(tex)
x <- cbind(x, tex)
```

```{r echo = FALSE}
kableExtra::kable_styling(kable(head(x)), full_width = FALSE)
```

Submit example data to the ROSETTA API and generate theoretical water retention curves.
```{r}
r <- ROSETTA(x, vars = c('sand', 'silt', 'clay'))

# iterate over results and generate VG model curve
res <- lapply(1:nrow(r), function(i) {
  
  # model bounds are given in kPA of suction
  vg <- KSSL_VG_model(VG_params = r[i, ], phi_min = 10^-3, phi_max=10^6)
  
  # extract curve and add texture ID
  m <- vg$VG_curve
  m$texture <- r$tex[i]
  
  return(m)
})

# flatten to data.frame
res <- do.call('rbind', res)
```


```{r echo = FALSE}
kableExtra::kable_styling(kable(head(res), digits = 3), full_width = FALSE)
```


Plot water retention curves. Note that matric potential (typically negative pressure by convention) are displayed as suction (positive values).
```{r fig.width=12, fig.height=6}
tps <- tactile.theme(
  plot.line=list(col='royalblue', lwd=2)
  )

xyplot(
  phi ~ theta | texture, data = res, 
  type = c('l', 'g'), 
  scales = list(alternating=3, x=list(tick.number=6), y=list(log=10, tick.number=6)), 
  yscale.components = yscale.components.logpower, 
  ylab = expression(Matric~~Potential~~(-kPa)), 
  xlab = expression(Volumetric~Water~Content~~(cm^3/cm^3)), 
  par.settings = tps, 
  strip = strip.custom(bg=grey(0.85)), 
  as.table = TRUE,
  layout = c(7, 3),
  main='Idealized Water Retention'
)
```

# Automatic Model Selection
```{r}
# data from Texas
q <- "SELECT TOP 10 musym, co.cokey, compname, comppct_r,
    hzname, hzdept_r, hzdepb_r, sandtotal_r, silttotal_r, claytotal_r, dbthirdbar_r,
    wthirdbar_r/100 AS wthirdbar_decimal, wfifteenbar_r/100 AS wfifteenbar_decimal
    FROM legend
    INNER JOIN mapunit mu ON mu.lkey = legend.lkey
    INNER JOIN component co ON mu.mukey = co.mukey
    INNER JOIN chorizon ch ON co.cokey = ch.cokey
    WHERE legend.areasymbol LIKE 'TX%'
    ORDER BY musym, co.cokey, ch.hzdept_r ASC;"

x <- SDA_query(q)

# simulate missing data
x$dbthirdbar_r[1] <- NA
x$wthirdbar_decimal[2] <- NA
x$wfifteenbar_decimal[3] <- NA
x$sandtotal_r[9] <- NA
x[10, ] <- NA

# attempting to use all possible soil properties
vars <- c('sandtotal_r', 'silttotal_r', 'claytotal_r', 'dbthirdbar_r', 'wthirdbar_decimal', 'wfifteenbar_decimal')

# automatic model selection by API
# based on available data
# m = "0"
r <- ROSETTA(x, vars = vars)
```

```{r echo = FALSE}
# check results
kableExtra::kable_styling(knitr::kable(r, row.names = FALSE, digits = 3), full_width = FALSE, font_size = 10)
```


----------------------------
This document is based on `soilDB` version `r utils::packageDescription("soilDB", field="Version")`.

