---
title: "Soil Data Access (SDA) Tutorial"
author: "D.E. Beaudette, A.G. Brown, S.M. Roecker"
date: "`r Sys.Date()`"
output:
  html_document:
    mathjax: null
    jquery: null
    smart: no
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, echo=FALSE, results='hide', warning=FALSE}
# setup
library(knitr, quietly=TRUE)

opts_chunk$set(
  message = FALSE, 
  warning = FALSE, 
  background = '#F7F7F7', 
  fig.align = 'center', 
  dev = 'svglite', 
  tidy = FALSE, 
  verbose = FALSE
)

options(
  width = 100, 
  stringsAsFactors = FALSE, 
  scipen = 10
)
```



# Introduction
This is a short tutorial on how to interact with the [Soil Data Access (SDA) web-service](http://sdmdataaccess.nrcs.usda.gov/) using R. Queries are written using [a dialect of SQL](https://technet.microsoft.com/en-us/library/bb264565(v=sql.90).aspx). On first glance SQL appears similar to the language used to write NASIS queries and reports, however, these are two distinct languages. Soil Data Access is a "window" into the spatial and tabular data associated with the current SSURGO snapshot. Queries can contain spatial and tabular filters. If you are new to SDA or SQL, have a look at [this page](http://sdmdataaccess.nrcs.usda.gov/QueryHelp.aspx). Here is a [SSURGO-specific](https://nasis.sc.egov.usda.gov/NasisReportsWebSite/limsreport.aspx?report_name=SDA-SQL_Library_Home) library of SQL code, organized by task.

Spatial queries can be included in SQL statements submitted to SDA as long as the geometry has first been transformed to [WGS84 geographic](http://spatialreference.org/ref/epsg/wgs-84/) coordinates and formatted as "[well known text](https://en.wikipedia.org/wiki/Well-known_text)" (WKT). The `sf` and `terra` packages provide functionality for converting between coordinate systems. Coordinate reference system definitions (a "CRS") are typically provided using EPSG codes. You can search for various CRS definitions in a variety of formats using [spatialreference.org/](http://spatialreference.org/). 

The `soilDB` library for R provides a helper function (`SDA_query()`) for submitting queries to SDA, processing the result, and reformatting the results into a rectangular table (a `data.frame`). Most of the work required to use the `SDA_query()` function will be writing SQL to describe the columns you would like returned and how the data should be filtered and possibly grouped. The `SDA_spatialQuery()` function simplifies the process of using spatial data (`sf`, `spatVector`, or `Spatial*` objects) to query map unit polygons.

Follow along with the blocks of code below by copying / pasting into a new R "script" document. Each block of command can be run by pasting into the R console, or by "stepping through" lines of code by moving the cursor to the top of a block (in the R script panel) and repeatedly pressing ctrl + enter. Note that `knittr:kable` is used to create cleaner output in the tutorial and isn't required while following along.

If you are feeling adventurous, have a look at a draft [tutorial on queries that return geometry from SDA](http://ncss-tech.github.io/AQP/soilDB/SDA-tutorial-2.html). Additional tips on advanced spatial queries can be found [here](http://sdmdataaccess.sc.egov.usda.gov/documents/AdvancedSpatialDataAccess.pdf).


## Critical Notes 

### SSURGO vs. STATSGO
SSURGO (1:24k soil survey) and STATSGO (1:250k soil survey) records are stored *together* in SDA. Therefore, it is critical that evey query to SDA include some kind of filter for selecting the appropriate records. Filtering strategies include:

   * explicit exclusion of STATSGO records, via `legend.areasymbol != 'US'` in the WHERE clause
   * implicit exclusion of STATSGO records, via SSURGO areasymbol in the WHERE clause
   * spatial queries using SDA helper functions: e.g. `SDA_Get_Mukey_from_intersection_with_WktWgs84()`
   * explicit selection of SSURGO / STATSGO records by record ID: e.g. `mukey`, `cokey`, etc.


Explicit exclusion of STATSGO records:
```{sql, eval=FALSE}
SELECT [...]
FROM legend
INNER JOIN mapunit mu ON mu.lkey = legend.lkey
INNER JOIN component co ON mu.mukey = co.mukey
WHERE legend.areasymbol != 'US' ;
```

Implicit exclusion of STATSGO records:
```{sql, eval=FALSE}
SELECT [...]
FROM legend
INNER JOIN mapunit mu ON mu.lkey = legend.lkey
INNER JOIN component co ON mu.mukey = co.mukey
WHERE legend.areasymbol = 'CA113' ;
```

### Integer Columns
Be sure to explicitly `CAST()` integer columns when using operators such as multiplication, division, etc.. or when values may be very large. Queries that work with large sets of `muacres` from the `mapunit` table are one such case. Implicit casting is accomplished via operations that include higher precision values (e.g. include values to the right of the decimal point).


Integer overflow, result is `NULL`.
```{sql eval = FALSE}
SELECT 
SUM(muacres) AS ac 
FROM legend
INNER JOIN mapunit mu ON mu.lkey = legend.lkey
WHERE legend.areasymbol != 'US' ;
```

Casting `muacres` to numeric data type, result is 2,280,193,353 acres.
```{sql eval = FALSE}
SELECT 
SUM(CAST(muacres AS NUMERIC)) AS ac 
FROM legend
INNER JOIN mapunit mu ON mu.lkey = legend.lkey
WHERE legend.areasymbol != 'US' ;
```

Implicit casting to numeric via division by floating point value, result is 2280.193 million acres.
```{sql eval = FALSE}
SELECT 
SUM(muacres / 1000000.0) AS ac 
FROM legend
INNER JOIN mapunit mu ON mu.lkey = legend.lkey
WHERE legend.areasymbol != 'US' ;
```


Scaling of `muacres` by component percentage, as converted to a fraction. Result is **wrong**: 500 million acres.
```{sql eval = FALSE}
SELECT 
SUM((comppct_r / 100) * muacres) / 1000000 AS ac 
FROM legend
INNER JOIN mapunit mu ON mu.lkey = legend.lkey
INNER JOIN component AS co ON mu.mukey = co.mukey 
WHERE legend.areasymbol != 'US' ;
```


Casting `comppct_r` and `muacres` to numeric, result is correct: 2242.39 million acres. Note that there are still some components with NULL `comppct_r`, will be fixed eventually.
```{sql eval = FALSE}
SELECT 
SUM((CAST(comppct_r AS NUMERIC) / 100) * CAST(muacres AS NUMERIC)) / 1000000 AS ac 
FROM legend
INNER JOIN mapunit mu ON mu.lkey = legend.lkey
INNER JOIN component AS co ON mu.mukey = co.mukey 
WHERE legend.areasymbol != 'US' ;
```

Use this last pattern whenever converting component percentages to a fraction, or scaling the `muacres` field.


## Install Required R Packages
You only need to do this once. If you haven't installed these packages, then copy the code below and paste into the RStudio "console" pane.
```{r install-deps, eval=FALSE}
# run these commands in the R console
# stable version from CRAN + dependencies
install.packages("aqp", dep = TRUE)
install.packages("soilDB", dep = TRUE)
install.packages("terra", dep = TRUE)
install.packages("sf", dep = TRUE)
```


# Simple Queries 
Now that you have the required packages, load them into the current R session.
```{r load-libs}
library(aqp)
library(soilDB)
library(terra)
library(sf)
```


When was the CA653 survey area last exported to SSURGO?
```{r SDA-one-liner-1}
SDA_query("SELECT areasymbol, saverest FROM sacatalog WHERE areasymbol = 'CA653'")
```

Are there any survey areas that haven't been updated since Jan 1, 2019? Just STATSGO.
```{r SDA-one-liner-2}
SDA_query("SELECT areasymbol, saverest FROM sacatalog WHERE saverest < '01/01/2019'")
```

What is the most recently updated survey in CA?
```{r SDA-one-liner-3}
SDA_query("SELECT areasymbol, saverest FROM sacatalog WHERE areasymbol LIKE 'CA%' ORDER BY saverest DESC")[1, ]
```

A simple query from the component table, for a single map unit: `mukey = '461958'`. This is a SSURGO map unit key, therefore STATSGO records are implicitly removed from the results.
```{r basic-SDA-1}
q <- "SELECT 
mukey, cokey, comppct_r, compname, taxclname
FROM component
WHERE mukey = '461958'"

# run the query
res <- SDA_query(q)

# check
knitr::kable(head(res))
```

Get a list of map units that contain "Amador" as minor component. Note that this type of query requires explicit exclusion of STATSGO records.
```{r basic-SDA-2}
q <- "SELECT 
muname, mapunit.mukey, cokey, compname, comppct_r
FROM legend
INNER JOIN mapunit ON mapunit.lkey = legend.lkey
INNER JOIN component on mapunit.mukey = component.mukey
WHERE
-- exclude STATSGO
legend.areasymbol != 'US'
AND compname LIKE '%amador%'
AND majcompflag = 'No'"

# run the query
res <- SDA_query(q)

# check
knitr::kable(head(res))

# optionally save the results to CSV file
# write.csv(res, file='path-to-file.csv', row.names=FALSE)
```

Get basic map unit and component data for a single survey area, Yolo County (CA113). There is no need to exclude STATSGO records because we are specifying a SSURGO areasymbol in the WHERE clause.
```{r basic-SDA-3}
q <- "SELECT 
component.mukey, cokey, comppct_r, compname, taxclname, 
taxorder, taxsuborder, taxgrtgroup, taxsubgrp
FROM legend
INNER JOIN mapunit ON mapunit.lkey = legend.lkey
INNER JOIN component ON component.mukey = mapunit.mukey
WHERE legend.areasymbol = 'CA113'"

# run the query
res <- SDA_query(q)

# check
knitr::kable(head(res))
```

Cross tabulate the occurrence of components named "Auburn" and "Dunstone" with survey areasymbol. Note that this type of query requires explicit exclusion of STATSGO records.
```{r basic-SDA-4}
q <- "SELECT areasymbol, component.mukey, cokey, comppct_r, compname, compkind, taxclname
FROM legend
INNER JOIN mapunit ON mapunit.lkey = legend.lkey
INNER JOIN component ON component.mukey = mapunit.mukey
WHERE compname IN ('Auburn', 'Dunstone')
-- exclude STATSGO
AND areasymbol != 'US'
ORDER BY areasymbol, compname"

res <- SDA_query(q)

xtabs(~ areasymbol + compname, data=res)
```

# Queries Using Simple Spatial Filters

Get the map unit key and name at a single, manually-defined point (-121.77100 37.368402). Spatial queries using SDA helper functions automatically exclude STATSGO records.

```{r sda-spatial-query-basics-1}
q <- "SELECT mukey, muname
FROM mapunit
WHERE mukey IN (
SELECT * from SDA_Get_Mukey_from_intersection_with_WktWgs84('point(-121.77100 37.368402)')
)"

res <- SDA_query(q)
knitr::kable(res)
```

Get the map names and `mukey` values for a 1000m buffer around a manually-defined point (-121.77100 37.368402). The `buffer()` function from the {terra} package can conveniently derive planar buffers from geographic coordinates.

```{r sda-spatial-query-basics-1.1}
# the query point is in geographic coordinates / WGS84 datum
p <- vect(cbind(-121.77100, 37.368402), crs = 'epsg:4326')

# 1000m buffer (planar coordinates)
# result is still geographic coordinates
p.buff <- terra::buffer(p, 1000)

# convert to WKT
p.wkt <- as.data.frame(p.buff, geom = 'wkt')$geometry

q <- paste0("SELECT mukey, muname
FROM mapunit
WHERE mukey IN (
SELECT * from SDA_Get_Mukey_from_intersection_with_WktWgs84('", p.wkt, "')
)")

res <- SDA_query(q)
knitr::kable(head(res))
```

Get some component data for a manually-defined bounding box, defined in WGS84 geographic coordinates.
```{r}
# define a bounding box: xmin, xmax, ymin, ymax
#
#         +-------------------(ymax, xmax)
#         |                        |
#         |                        |
#     (ymin, xmin) ----------------+

# bbox extent
bb <- ext(-120.9, -120.8, 37.7, 37.8)

# convert to polygons
bb <- vect(bb, crs = 'epsg:4326')

# convert bounding box to WKT
bb <- as.data.frame(bb, geom = 'wkt')$geometry

# compose query, using WKT BBOX as filtering criteria
q <- paste0("SELECT mukey, cokey, compname, comppct_r
            FROM component 
            WHERE mukey IN (SELECT DISTINCT mukey FROM SDA_Get_Mukey_from_intersection_with_WktWgs84('", bb, "') )
            ORDER BY mukey, cokey, comppct_r DESC")

res <- SDA_query(q)

# check
knitr::kable(head(res))
```


Compute polygon area via SDA and double-check via `terra::expanse()`. Note the complex syntax required to convert polygons into "geography" type, thus invoking spherical geometry and resulting in area values with units of square meters. Computing polygon area server-side requires a more complex SQL statement, but is much more efficient because the associated polygon data do not need to be transferred from SDA to the local R session.

```{r}
# much faster (10x or more), because area is calculated server-side
a <- SDA_query("
SELECT mukey,
GEOGRAPHY::STGeomFromWKB(
  mupolygongeo.STUnion(
    mupolygongeo.STStartPoint()
  ).STAsBinary(), 4326
).STArea() * 0.000247105 AS area_ac
FROM mupolygon WHERE mukey = '418254'
")

# slow, polygon geometry must be converted to text and downloaded
x <- fetchSDA_spatial('418254')
x <- vect(x)
# compute area in sq. meters then convert to acres
x$area <-expanse(x) * 0.000247105

# note that the distribution of polygon area is identical
options(scipen = 10)
round(quantile(a$area_ac))
round(quantile(x$area))
```

# Helper Functions

## `fetchSDA()`

```{r fig.width=8, fig.height=5}
library(ggplot2)
library(gridExtra)

# query soil components by areasymbol and musym
test <- fetchSDA(WHERE = "areasymbol = 'IN005' AND musym = 'MnpB2'")

# profile plot
par(mar = c(0, 0, 0, 2.5))
plotSPC(test, name.style = 'center-center', cex.names = 0.8)
```

```{r fig.width=10, fig.height=7}
# convert the data for depth plot
clay_slice <- horizons(dice(test, 0:200 ~ claytotal_l + claytotal_r + claytotal_h))
names(clay_slice) <- gsub("claytotal_", "", names(clay_slice))

om_slice <- horizons(dice(test, 0:200 ~ om_l + om_r + om_h))
names(om_slice) <- gsub("om_", "", names(om_slice))

test2 <- rbind(data.frame(clay_slice, var = "clay"),
              data.frame(om_slice, var = "om")
)

h <- merge(test2, site(test)[c("nationalmusym", "cokey", "compname", "comppct_r")],
          by = "cokey",
          all.x = TRUE
)

# depth plot of clay content by soil component
gg_comp <- function(x) {
  ggplot(x) +
    geom_line(aes(y = r, x = hzdept_r)) +
    geom_line(aes(y = r, x = hzdept_r)) +
    geom_ribbon(aes(ymin = l, ymax = h, x = hzdept_r), alpha = 0.2) +
    xlim(200, 0) +
    xlab("depth (cm)") +
    facet_grid(var ~ nationalmusym + paste(compname, comppct_r)) +
    coord_flip()
}
g1 <- gg_comp(subset(h, var == "clay"))
g2 <- gg_comp(subset(h, var == "om"))

grid.arrange(g1, g2)
```

```{r eval = FALSE, echo = FALSE, fig.width=10, fig.height=7}
## TODO: this one isn't working, month not encoded as integers / factors

# query cosoilmoist (e.g. water table data) by mukey
x <- get_cosoilmoist_from_SDA(WHERE = "mukey = '328164'", stringsAsFactors = TRUE)

ggplot(x, aes(x = month, y = soimoistdept_r)) +
  geom_rect(aes(xmin = 0.5, xmax = 12.5,
                ymin = 0, ymax = max(x$soimoistdepb_r),
                fill = flodfreqcl)) +
  geom_line(cex = 1) +
  geom_point() +
  geom_ribbon(aes(ymin = soimoistdept_l, ymax = soimoistdept_h), alpha = 0.2) +
  ylim(max(x$soimoistdepb_r), 0) +
  xlab("month") + ylab("depth (cm)") +
  # scale_x_continuous(breaks = 1:12, labels = month.abb, name="Month") +
  facet_wrap(~ paste0(compname, ' (', comppct_r , ')')) +
  ggtitle(paste0(x$nationalmusym[1],
                 ': Water Table Levels from Component Soil Moisture Month Data'))
```

```{r fig.width=12, fig.height=10}
# query all Miami major components
s <- get_component_from_SDA(WHERE = "compname = 'Miami' \n
              AND majcompflag = 'Yes' AND areasymbol != 'US'")


# landform vs 3-D morphometry
test <- {
  subset(s, ! is.na(landform) | ! is.na(geompos)) ->.;
  split(., .$drainagecl, drop = TRUE) ->.;
  lapply(., function(x) {
    test = data.frame()
    test = as.data.frame(table(x$landform, x$geompos))
    test$compname   = x$compname[1]
    test$drainagecl = x$drainagecl[1]
    names(test)[1:2] <- c("landform", "geompos")
    return(test)
  }) ->.;
  do.call("rbind", .) ->.;
  .[.$Freq > 0, ] ->.;
  within(., {
    landform = reorder(factor(landform), Freq, max)
    geompos  = reorder(factor(geompos),  Freq, max)
    geompos  = factor(geompos, levels = rev(levels(geompos)))
  }) ->.;
}

test$Freq2 <- cut(test$Freq,
                  breaks = c(0, 5, 10, 25, 50, 100, 150),
                  labels = c("<5", "5-10", "10-25", "25-50", "50-100", "100-150")
)


ggplot(test, aes(x = geompos, y = landform, fill = Freq2)) +
  geom_tile(alpha = 0.5) + facet_wrap(~ paste0(compname, "\n", drainagecl)) +
  discrete_scale("colour", "viridis", function(n) viridisLite::viridis(n)) +
  theme(aspect.ratio = 1, axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
  ggtitle("Landform vs 3-D Morphometry for Miami Major Components on SDA")

```





## `SDA_spatialQuery()`
Spatial queries are usually simpler with a two step process via `SDA_spatialQuery()` to get a set of intersecting map unit keys, then `SDA_query()` to return additional tabular data. Some things to keep in mind when using `SDA_spatial_query()`:

  * use `what = 'mupolygon'` to return polygon features and computed area in acres (slower, higher resource usage)
  * use `what = 'mukey'` to return only map unit keys (faster, lower resource usage)
  * spatial filters must have a valid CRS defined
  * spatial queries for more than 1000 features should probably be done using a local copy of the map unit polygons
  
Long-running queries and queries that consume too many server resources will be canceled by SDA. See `?SDA_spatialQuery` for details.
```{r}
# query map unit keys intersecting with the 1000m buffered point defined above
mu <- SDA_spatialQuery(p.buff, what = 'mukey', geomIntersection = TRUE)

# create a query to get component data for resulting map unit keys
sql <- sprintf("SELECT mukey, cokey, compname, comppct_r
               FROM component
               WHERE mukey IN %s
               ORDER BY cokey, comppct_r DESC", format_SQL_in_statement(mu$mukey)
) 

co <- SDA_query(sql)

knitr::kable(head(co))
```


Try downloading some polygon geometry using the same buffered point. The `geomIntersection` argument to `SDA_spatialQuery()` toggles overlap (`FALSE`) vs. intersection (`TRUE`).
```{r}
# use the buffered point geometry to query SDA
# requesting spatial intersection vs. overlapping features
p.mu.polys <- SDA_spatialQuery(p.buff, what = 'mupolygon', geomIntersection = TRUE)
```

Graphical description of the previous steps: query point, 1000m buffer, buffer bounding box (BBOX), and intersection of buffered point / map unit polygons.
```{r results='hide', fig.width=6, fig.height=6}
# plot intersecting map unit polygons from SDA
plot(p.mu.polys, axes = FALSE, mar = c(1, 2, 0, 0))

# add buffered point
lines(p.buff, col = 'red', lwd = 2)

# add bounding box around buffered point
lines(ext(p.buff), col = 'RoyalBlue')

# add original point
points(p, col='orange', pch = 15, cex = 2)

legend('bottomleft', legend=c('query point', '1000m buffer', 'buffer BBOX', 'polygon intersection'), col=c('orange', 'red', 'royalblue', 'black'), lwd = c(NA, 2, 2, 2), pch=c(15, NA, NA, NA), xpd = FALSE, cex = 0.8)
```






Let's apply the `SDA_spatialQuery()` function to some real data; KSSL pedons correlated to "Auburn". Not all of these pedons have coordinates, so we will have to do some filtering first. See the in-line comments for details on each line of code.
```{r sda-spatial-kssl, results='hide'}
# get KSSL pedons with taxonname = Auburn
# result is a SoilProfileCollection object
# coordinates will be WGS84 GCS
auburn <- fetchKSSL('auburn')

# keep only those pedons with valid coordinates
auburn <- subset(auburn, subset = !is.na(x) & !is.na(y))

# extract site data with coordinates
s <- site(auburn)

# init spatVector
sp <- vect(s, geom = c('x', 'y'), crs = 'epsg:4326')

# perform SDA query on collection of points
mu.data <- SDA_spatialQuery(sp, what = 'mupolygon')

# use local spatial intersection to link source + SDA data
# row-order is preserved
sp$mukey <- terra::extract(mu.data, sp)$mukey

# join results to original SoilProfileCollection using 'pedlabsampnum'
site(auburn) <- as.data.frame(sp)[, c('pedlabsampnum', 'mukey')]
```


Check the results and plot the "Auburn" KSSL pedons, grouped by intersecting map unit key and coloring horizons according to clay content.
```{r fig.width=12, fig.height=5}
# plot profiles, grouped by mukey
# color horizons with clay content
par(mar = c(0, 2, 4, 2))

groupedProfilePlot(auburn, groups = 'mukey', name.style = 'center-center', group.name.cex = 0.65, color = 'clay', id.style = 'side', label = 'pedon_id', max.depth = 100)

# describe IDs
mtext('user pedon ID', side = 2, line = -1.5)
mtext('mukey', side = 3, line = -1, at = c(0, 0), adj = 0)
```


----------------------------
This document is based on `soilDB` (`r utils::packageDescription("soilDB", field="Version")`) and `aqp` (`r utils::packageDescription("aqp", field="Version")`).
