---
output:
  html_document:
    fig_caption: yes
    jquery: null
    mathjax: null
    smart: no
---

```{r setup, echo=FALSE, results='hide', warning=FALSE}
# setup
library(knitr, quietly=TRUE)
# library(printr, quietly=TRUE)
opts_chunk$set(message=FALSE, warning=FALSE, background='#F7F7F7', fig.align='center', fig.retina=1, dev='png', tidy=FALSE, verbose=FALSE)
options(width=100, stringsAsFactors=FALSE)
```

Soil Data Access (SDA2) Tutorial
================================
`r format(Sys.time(), "%Y-%m-%d")`
<br>Dylan Beaudette

## Introduction
This is a short tutorial on how to interact with the [Soil Data Access (SDA) web-service](http://sdmdataaccess.nrcs.usda.gov/) using R. Queries are written using [a dialect of SQL](https://technet.microsoft.com/en-us/library/bb264565(v=sql.90).aspx). On first glance SQL appears similar to the language used to write NASIS queries and reports, however, these are two distinct languages. Soil Data Access is a "window" into the spatial and tabular data associated with the current SSURGO snapshot. Queries can contain spatial and tabular filters. If you are new to SDA or SQL, have a look at [this page](http://sdmdataaccess.nrcs.usda.gov/QueryHelp.aspx).

Spatial queries can be included in SQL statements submitted to SDA as long as the geometry has first been transformed to [WGS84 geographic](http://spatialreference.org/ref/epsg/wgs-84/) (or [psuedo-Web Mercator](http://spatialreference.org/ref/sr-org/7483/)) coordinates and formatted as "[well known text](https://en.wikipedia.org/wiki/Well-known_text)" (WKT). The `sp` and `rgdal` packages provide functionality for converting between coordinate systems via `spTransform()`. Coordinate reference system definitions (a "CRS") are typically provided using [proj4](https://en.wikipedia.org/wiki/PROJ.4) notation. You can search for various CRS definitions in a variety of formats using [spatialreference.org/](http://spatialreference.org/). 

The `soilDB` library for R provides a helper function (`SDA_query()`) for submitting queries to SDA, processing the result, and reformatting the results into a rectangular table (a `data.frame`). Most of the work required to use the `SDA_query()` function will be writing SQL to describe the columns you would like returned and how the data should be filtered and possibly grouped. 

Follow along with the blocks of code below by copying / pasting into a new R "script" document. Each block of command can be run by pasting into the R console, or by "stepping through" lines of code by moving the cursor to the top of a block (in the R script panel) and repeatedly pressing ctrl + enter.

If you are feeling adventurous, have a look at a draft [tutorial on queries that return geometry from SDA](http://ncss-tech.github.io/AQP/soilDB/SDA-tutorial-2.html). Additional tips on advanced spatial queries can be found [here](http://sdmdataaccess.sc.egov.usda.gov/documents/AdvancedSpatialDataAccess.pdf).

### Install Required R Packages
You only need to do this once. If you haven't installed these packages, then copy the code below and paste into the RStudio "console" pane.
```{r install-deps, eval=FALSE}
# run these commands in the R console
# stable version from CRAN + dependencies
install.packages("httr", dep=TRUE)
install.packages("soilDB", dep=TRUE)
install.packages("rgdal", dep = TRUE)
install.packages("raster", dep = TRUE)
install.packages("rgeos", dep = TRUE)
# latest versions from r-forge
install.packages("soilDB", repos = "http://R-Forge.R-project.org", type = "source")
```


### Simple Queries 

Now that you have the required packages, load them into the current R session.
```{r load-libs}
library(soilDB)
library(sp)
library(rgdal)
library(plyr)
library(raster)
library(rgeos)
```


When was the CA653 survey area last exported to SSURGO?
```{r SDA-one-liner-1}
SDA_query("SELECT areasymbol, saverest FROM sacatalog WHERE areasymbol = 'CA653'")
```

Are there any survey areas that haven't been updated since Jan 1, 2010?
```{r SDA-one-liner-2}
SDA_query("SELECT areasymbol, saverest FROM sacatalog WHERE saverest < '01/01/2010'")
```

What is the most recently updates survey in CA?
```{r SDA-one-liner-3}
SDA_query("SELECT areasymbol, saverest FROM sacatalog WHERE areasymbol LIKE 'CA%' ORDER BY saverest DESC")[1, ]
```

A simple query from the component table, for a single map unit: `mukey = '461958'`.
```{r basic-SDA-1}
q <- "SELECT 
mukey, cokey, comppct_r, compname, taxclname
FROM component
WHERE mukey = '461958'"

# run the query
res <- SDA_query(q)

# check
head(res)
```

Get a list of map units that contain "Amador" as minor component.
```{r basic-SDA-2}
q <- "SELECT 
muname, mapunit.mukey, cokey, compname, comppct_r
FROM mapunit INNER JOIN component on mapunit.mukey = component.mukey
WHERE compname LIKE '%amador%'
AND majcompflag = 'No'"

# run the query
res <- SDA_query(q)

# check
head(res)

# optionally save the results to CSV file
# write.csv(res, file='path-to-file.csv', row.names=FALSE)
```

Get basic map unit and component data for a single survey area, Yolo County (ca113).
```{r basic-SDA-3}
q <- "SELECT 
component.mukey, cokey, comppct_r, compname, taxclname, 
taxorder, taxsuborder, taxgrtgroup, taxsubgrp
FROM legend
INNER JOIN mapunit ON mapunit.lkey = legend.lkey
LEFT OUTER JOIN component ON component.mukey = mapunit.mukey
WHERE legend.areasymbol = 'ca113'"

# run the query
res <- SDA_query(q)

# check
head(res)
```

Cross tabulate the occurrence of components named "Auburn" and "Dunstone" with survey areasymbol.
```{r basic-SDA-4}
q <- "SELECT areasymbol, component.mukey, cokey, comppct_r, compname, compkind, taxclname
FROM legend
INNER JOIN mapunit ON mapunit.lkey = legend.lkey
LEFT OUTER JOIN component ON component.mukey = mapunit.mukey
WHERE compname IN ('Auburn', 'Dunstone')
AND areasymbol != 'US'
ORDER BY areasymbol, compname"

res <- SDA_query(q)

xtabs(~ areasymbol + compname, data=res)
```

### Queries Using Simple Spatial Filters

Get the map unit key and name at a single, manually-defined point (-121.77100 37.368402).
```{r sda-spatial-query-basics-1}
q <- "SELECT mukey, muname
FROM mapunit
WHERE mukey IN (
SELECT * from SDA_Get_Mukey_from_intersection_with_WktWgs84('point(-121.77100 37.368402)')
)"

SDA_query(q)
```

Get the map names and mukey values for a 1000m buffer around a manually-defined point (-121.77100 37.368402). A 1000m buffer applied to geographic coordinates will require several spatial transformations. First, the query point needs to be initialized in a geographic coordinate system with WGS84 datum. Next, the point is transformed to a planar coordinate system with units in meters; the [NLCD](http://www.mrlc.gov/index.php) [coordinate reference system](http://spatialreference.org/ref/sr-org/6630/) works well for points within the continental US. After computing a buffer in planar coordinates, the resulting polygon is converted back to geographic coordinates--this is what SDA is expecting.

```{r sda-spatial-query-basics-1.1}
# the query point is in geographic coordinates with WGS84 datum
p <- SpatialPoints(cbind(-121.77100, 37.368402), proj4string = CRS('+proj=longlat +datum=WGS84'))
# transform to planar coordinate system for buffering
p.aea <- spTransform(p, CRS('+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=23 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs '))
# create 1000 meter buffer
p.aea <- gBuffer(p.aea, width = 1000)
# transform back to WGS84 GCS
p.buff <- spTransform(p.aea, CRS('+proj=longlat +datum=WGS84'))
# convert to WKT
p.wkt <- writeWKT(p.buff)

q <- paste0("SELECT mukey, muname
FROM mapunit
WHERE mukey IN (
SELECT * from SDA_Get_Mukey_from_intersection_with_WktWgs84('", p.wkt, "')
)")

res <- SDA_query(q)
head(res)
```


It is possible to download small collections of SSURGO map unit polygons from SDA using a bounding-box in WGS84 geographic coordinates. SDA will return polygons and their map unit keys that **overlap** with the BBOX query.
```{r sda-spatial-query-basics-1.2, results='hide'}
# extract bounding-box from out last point
# coordinates are in WGS84 GCS
b <- as.vector(bbox(p.buff))
# download map unit polygons that overlap with bbox
p.mu.polys <- mapunit_geom_by_ll_bbox(b)
```

Graphical description of the previous steps: query point, 1000m buffer, buffer bounding box (BBOX), intersecting map unit polygons, and overlapping polygons.
```{r sda-spatial-query-basics-1.3, results='hide', fig.width=5, fig.height=5}
# plot
par(mar=c(0,0,0,0))
plot(p.mu.polys)
plot(p.mu.polys[which(p.mu.polys$mukey %in% setdiff(p.mu.polys$mukey, res$mukey)), ], add=TRUE, col='grey')
lines(p.buff, col='red', lwd=2)
plot(extent(bbox(p.buff)), add=TRUE, col='RoyalBlue')
points(p, col='orange', pch=15)
legend('bottomleft', legend=c('query point', '1000m buffer', 'buffer BBOX', 'intersected polygons', 'overlapping polygons'), col=c('orange', 'red', 'royalblue', 'black', 'grey'), lwd=c(NA, 2, 2, 2, 2), pch=c(15, NA, NA, NA, NA))
box()
```


Get some component data for a manually-defined bounding box, defined in WGS84 geographic coordinates.
```{r sda-spatial-query-basics-2}
# define a bounding box: xmin, xmax, ymin, ymax
#
#         +-------------------(ymax, xmax)
#         |                        |
#         |                        |
#     (ymin, xmin) ----------------+

b <- c(-120.9, -120.8, 37.7, 37.8)
# convert bounding box to WKT
p <- writeWKT(as(extent(b), 'SpatialPolygons'))
# compose query, using WKT BBOX as filtering criteria
q <- paste0("SELECT mukey, cokey, compname, comppct_r
            FROM component 
            WHERE mukey IN (SELECT DISTINCT mukey FROM SDA_Get_Mukey_from_intersection_with_WktWgs84('", p, "') )
            ORDER BY mukey, cokey, comppct_r DESC")

res <- SDA_query(q)

# check
head(res)
```


### Queries Using Complex Spatial Filters
The examples above illustrate how to perform SDA queries using a single spatial filter. Typically we need to perform these queries over a collection of points, lines or polygons. The `soilDB` package provides some helper functions for iterating over a collection of features (usually points). Note that spatial queries for more than 1000 features should probably be done using a local copy of the map unit polygons.

The first function `SDA_make_spatial_query()` will convert a single `Spatial*` (Points, Lines, Polygons) object to WKT and submit a spatial query to SDA, returning intersecting map unit keys and names. Let's try it using a `SpatialPoints` object with 1 feature.
```{r sda-spatial-query-helper-functions-examples-1, results='hide'}
# the query point is in geographic coordinates with WGS84 datum
p <- SpatialPoints(cbind(-121.77100, 37.368402), proj4string = CRS('+proj=longlat +datum=WGS84'))
SDA_make_spatial_query(p)
```

The second function `SDA_query_features()` will iterate over the features of a `Spatial*` (Points, Lines, Polygons) object, send a query to SDA, and return a set of the results as a `data.frame` object. This time, our example set of 2 points will also have some fake pedons IDs.
```{r sda-spatial-query-helper-functions-examples-2, results='hide'}
# the query points are in geographic coordinates with WGS84 datum
p <- SpatialPointsDataFrame(cbind(c(-121, -122), c(37, 37)), data=data.frame(pedon_id=1:2), proj4string = CRS('+proj=longlat +datum=WGS84'))
SDA_query_features(p, id='pedon_id')
```


Let's apply the `SDA_query_features()` function to some real data; KSSL pedons correlated to "Auburn". Not all of these pedons have coordinates, so we will have to do some filtering first. See the in-line comments for details on each line of code.
```{r sda-spatial-kssl, results='hide'}
# get KSSL pedons with taxonname = Auburn
# coordinates will be NAD83 GCS
auburn <- fetchKSSL('auburn')
# keep only those pedons with valid coordinates
auburn <- subsetProfiles(auburn, s='!is.na(x) & !is.na(y)')
# init spatial information
coordinates(auburn) <- ~ x + y
# define projection
proj4string(auburn) <- '+proj=longlat +datum=NAD83'

# extract "site data"
auburn.sp <- as(auburn, 'SpatialPointsDataFrame')
# perform SDA query on each "site", result is a data.frame
mu.data <- SDA_query_features(auburn.sp, id='pedlabsampnum')

# join results to original SoilProfileCollection using 'pedlabsampnum'
site(auburn) <- mu.data
```


Check the results and plot the "Auburn" KSSL pedons, grouped by intersecting map unit key and coloring horizons according to clay content.
```{r sda-spatial-kssl-plot, fig.width=12, fig.height=5}
# check results
print(mu.data)

# plot profiles, grouped by mukey
# color horizons with clay content
par(mar=c(0,0,4,0))
groupedProfilePlot(auburn, groups='mukey', group.name.cex=0.65, color='clay', name='hzn_desgn', id.style='side', label='pedon_id', max.depth=100)
# describe IDs
mtext('user pedon ID', side=2, line=-1.5)
mtext('mukey', side=3, line=-1, at = c(0,0), adj = 0)
```


More examples pending...



```{r sda-spatial-query-basics-3, eval=FALSE, echo=FALSE}
## Get the map unit name and mukey that intersect from several points randomly located within MLRA 17. These points could also come from a SHP file of point locations.

# MLRA boundaries, GCS NAD83
mlra <- readOGR('L:/Geodata/Boundaries', layer='mlra_v42', stringsAsFactors = FALSE)

# subset to MLRA 17
mlra <- mlra[which(mlra$MLRARSYM == '17'), ]
# generate 10 randomly located sample points
s <- spsample(mlra, n = 10, type = 'stratified')

# check
plot(mlra)
points(s, col='red')



# check
head(d)

# optionally combine with sample point geometry
# s.new <- SpatialPointsDataFrame(s, data=d)

# optionally save to SHP
# writeOGR(s.new, dsn='.', layer='sample-points', driver='ESRI Shapefile', overwrite_layer = TRUE)
```

```{r sda-spatial-query-via-raster, eval=FALSE, echo=FALSE}
##
## this is getting rather convoluted... 
##


# mean frost-free-days: NAD83 GCS
r <- raster('E:/gis_data/prism/ffd_mean_800m.tif')
# MLRA boundaries, NAD83 GCS 
mlra <- readOGR('L:/Geodata/Boundaries', layer='mlra_v42', stringsAsFactors = FALSE)
# subset to MLRA 15
mlra <- mlra[which(mlra$MLRARSYM == '15'), ]
# crop FFD to mlra 15 extent
r <- crop(r, mlra)
# extract 250 FFD contour
# ffd.250 <- rasterToPolygons(r, fun=function(i) {i > 250}, dissolve=TRUE)

# check: ok
plot(r)
plot(mlra, add=TRUE)
plot(ffd.250, add=TRUE, col='red')

# ... now what?

# rasterize MLRA 15 using same grid as FFD
mlra.r <- rasterize(mlra, r)
# convert all non-NA values to 1
mlra.r[!is.na(mlra.r)] <- 1

# overlay FFD > 200 and MLRA 15 rasters
x <- r + mlra.r
# check: OK
plot(x)

# convert to points
s <- rasterToPoints(x, spatial = TRUE)

# query ?



```

----------------------------
This document is based on `soilDB` version `r utils::packageDescription("soilDB", field="Version")`.
