---
title: "SDA Point Query Examples"
author: "D.E. Beaudette"
date: "`r Sys.Date()`"
output:
  html_document:
    mathjax: null
    jquery: null
    smart: no
    number_sections: no
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: false
---


```{r setup, echo=FALSE, results='hide', warning=FALSE}
# setup
library(knitr, quietly=TRUE)
library(kableExtra, quietly=TRUE)
opts_chunk$set(message=FALSE, warning=FALSE, background='#F7F7F7', fig.align='center', dev='svglite', tidy=FALSE, verbose=FALSE)
options(width=100, stringsAsFactors=FALSE)
```

# Setup
```{r install-deps, eval=FALSE}
# stable packages from CRAN
install.packages('aqp')
install.packages('soilDB')
install.packages('sharpshootR')
install.packages('sf')
install.packages('cowplot')

# latest versions from GitHub
remotes::install_github("ncss-tech/aqp", dependencies=FALSE, upgrade_dependencies=FALSE)
remotes::install_github("ncss-tech/soilDB", dependencies=FALSE, upgrade_dependencies=FALSE)
remotes::install_github("ncss-tech/sharpshootR", dependencies=FALSE, upgrade_dependencies=FALSE)
```

# Demonstration

Query [detailed soil survey data (SSURGO)](https://www.nrcs.usda.gov/wps/portal/nrcs/detail/soils/survey/?cid=nrcs142p2_053627) and associated Official Series Descriptions (OSD) at one or more points. In this example, we are using a single [point located near the NC / GA border](https://casoilresource.lawr.ucdavis.edu/gmap/?loc=35.04229,-83.45657,z15). These examples demonstration how to use web-services to request spatial + tabular data from Soil Data Access (SDA) and SoilWeb.


Load required packages and define one or more points using functionality from the `sf` package. 
```{r}
library(soilDB)
library(aqp)
library(sharpshootR)
library(sf)


# example point, WGS84 coordinates
p <- st_as_sf(data.frame(x = -83.45657,
                         y = 35.04229),
              coords = c('x', 'y'),
              crs = 4326)

# 500m buffer applied to point defined in geographic coordinates
b <- st_buffer(p, dist = units::set_units(500, m))

# check with interactive map
# mapview::mapview(b)
```

Submit a spatial query to SDA, asking for the spatial intersection of our example buffered point and SSURGO delineations.
```{r}
# result is an sf object
mu <- SDA_spatialQuery(b, what = 'mupolygon', byFeature = TRUE, geomIntersection = TRUE)

# check: OK
plot(st_geometry(mu))
plot(p, add = TRUE, pch = 16, col = 2)

# copy / encode as a factor for thematic map
mu$mukey.factor <- factor(mu$mukey)

par(cex.axis = 0.75)
plot(mu['mukey.factor'], key.width = lcm(4.5), main = 'Map Unit Key -- 500m Buffer', reset = FALSE)
plot(p, add = TRUE, pch = 16, col = 1)

# be sure to reset the graphics device
# dev.off()
```


Request associated tabular data from SDA, using the unique set of map unit keys. The result is a [`SoilProfileCollection` object](http://ncss-tech.github.io/AQP/aqp/aqp-intro.html).
```{r}
# format map unit keys for WHERE clause
.is <- format_SQL_in_statement(unique(mu$mukey))

# format WHERE clause
# don't use a semicolon to finish the SQL statement, 
# there is more added by fetchSDA()
.where <- sprintf("mukey IN %s AND majcompflag = 'Yes' AND compkind != 'Miscellaneous area' ", .is)

# query SDA, result is a SoilProfileCollection of map unit / component / horizon data
x <- fetchSDA(WHERE = .where, duplicates = TRUE, childs = FALSE)

# encode soil texture as ordered factor
x$texture <- droplevels(factor(x$texture, levels = SoilTextureLevels(which = 'codes')))
```

Visual examination via soil profile sketches.
```{r fig.width=14, fig.height=5.5}
# quick visual check
par(mar = c(0, 0, 3, 0))

# profile sketches from SSURGO
plotSPC(x, label = 'compname', color = 'texture', name.style = 'center-center', cex.names = 0.75, hz.depths = TRUE, plot.depth.axis = FALSE, col.label = 'Soil Texture Class', fixLabelCollisions = TRUE, hz.depths.offset = 0.05)

# group by national map unit symbol
groupedProfilePlot(x, group.name.offset = c(-15), groups = 'nationalmusym', label = 'compname', color = 'texture', name.style = 'center-center', cex.names = 0.75, hz.depths = TRUE, plot.depth.axis = FALSE, col.label = 'Soil Texture Class', fixLabelCollisions = TRUE, hz.depths.offset = 0.05)
```

Get OSD typical pedon morphology from SoilWeb.
```{r fig.width=12, fig.height=6.5}
# get OSDs and display
o <- fetchOSD(unique(x$compname))

par(mar = c(0, 0, 0, 0))

# OSDs
plotSPC(o, width = 0.3, name.style = 'center-center', cex.names = 0.75, hz.depths = TRUE, plot.depth.axis = FALSE, fixLabelCollisions = TRUE, hz.depths.offset = 0.05, shrink = TRUE)

# OSDs arranged by subgroup taxonomy
SoilTaxonomyDendrogram(o, width = 0.3, name.style = 'center-center', cex.names = 0.75, hz.depths = TRUE, plot.depth.axis = FALSE, fixLabelCollisions = TRUE, hz.depths.offset = 0.05, shrink = TRUE)
```

Just for fun, generate a hexagonal grid and query SSURGO data within grid cells.
```{r fig.width=8, fig.height=8}
# make hexagonal grid based on buffered point
g <- st_make_grid(b, n = 1, square = FALSE, flat_topped = TRUE)
g <- st_as_sf(g)

# query SDA for intersecting map unit delineations
mu <- SDA_spatialQuery(g, what = 'mupolygon', byFeature = TRUE, geomIntersection = TRUE)

# note that intersecting delineations are grouped by 'gid'
table(mu$gid)

# copy / encode as a factor for thematic map
mu$mukey.factor <- factor(mu$mukey)

# pretty figure
par(cex.axis = 0.75)
plot(mu['mukey.factor'], key.width = lcm(4.5), main = 'Map Unit Keys', reset = FALSE)
plot(p, add = TRUE, pch = 16, col = 1)
```

----------------------------
This document is based on `aqp` version `r utils::packageDescription("aqp", field="Version")`, `soilDB` version `r utils::packageDescription("soilDB", field="Version")`, and `sharpshootR` version `r utils::packageDescription("sharpshootR", field="Version")`.
