---
title: "Writing SDA Queries that Return Geometry"
author: "D.E. Beaudette"
date: "`r Sys.Date()`"
output:
  html_document:
    mathjax: null
    jquery: null
    smart: no
---
```{r setup, echo=FALSE, results='hide', warning=FALSE}
# setup
library(knitr, quietly=TRUE)

opts_chunk$set(message=FALSE, warning=FALSE, background='#F7F7F7', fig.align='center', fig.retina=1, dev='png', tidy=FALSE, verbose=FALSE)
options(width=100, stringsAsFactors=FALSE)
```

# Introduction
This is a short tutorial on how to interact with the [Soil Data Access (SDA) web-service](http://sdmdataaccess.nrcs.usda.gov/) using R. Queries are written using [a dialect of SQL](https://technet.microsoft.com/en-us/library/bb264565(v=sql.90).aspx). On first glance SQL appears similar to the language used to write NASIS queries and reports, however, these are two distinct languages. Soil Data Access is a "window" into the spatial and tabular data associated with the current SSURGO snapshot. Queries can contain spatial and tabular filters. If you are new to SDA or SQL, have a look at [this page](http://sdmdataaccess.nrcs.usda.gov/QueryHelp.aspx).

If this is your first time using SDA, please see a related [tutorial](http://ncss-tech.github.io/AQP/soilDB/SDA-tutorial.html) to get started.

Additional tips on advanced spatial queries can be found [here](http://sdmdataaccess.sc.egov.usda.gov/documents/AdvancedSpatialDataAccess.pdf).

[details pending]

Follow along with the blocks of code below by copying / pasting into a new R "script" document. Each block of commands can be run by pasting them into the R console, or by "stepping through" lines of code by moving the cursor to the top of a block (in the R script panel) and repeatedly pressing ctrl + enter.

## Install Required R Packages
You only need to do this once. If you haven't installed these packages, then copy the code below and paste into the RStudio "console" pane.
```{r install-deps, eval=FALSE}
# run these commands in the R console
# stable version from CRAN + dependencies
install.packages("httr", dep=TRUE)
install.packages("soilDB", dep=TRUE)
install.packages("rgdal", dep = TRUE)
install.packages("raster", dep = TRUE)
install.packages("rgeos", dep = TRUE)
```


## Critical Note: SSURGO vs. STATSGO
SSURGO (1:24k soil survey) and STATSGO (1:250k soil survey) records are stored *together* in SDA. Therefore, it is critical that evey query to SDA include some kind of filter for selecting the appropriate records. Filtering strategies include:

   * explicit exclusion of STATSGO records, via `legend.areasymbol != 'US'` in the WHERE clause
   * implicit exclusion of STATSGO records, via SSURGO areasymbol in the WHERE clause
   * spatial queries using SDA helper functions: e.g. `SDA_Get_Mukey_from_intersection_with_WktWgs84()`
   * explicit selection of SSURGO / STATSGO records by record ID: e.g. `mukey`, `cokey`, etc.


# Simple Queries 

These queries implicitly *filter-out* STATSGO via `SDA_Get_MupolygonWktWgs84_from_Mukey`.
```{r}
library(soilDB)
library(rgeos)
library(sp)
library(raster)
library(maps)


# get polygons for a single mukey
q <- "SELECT G.MupolygonWktWgs84 as geom, '462594' as mukey from SDA_Get_MupolygonWktWgs84_from_Mukey('462594') as G"
res <- SDA_query(q)

# result is a data.frame, "MupolygonWktWgs84" contains WKT representation of geometry
str(res)

# convert to SPDF
s <- processSDA_WKT(res)

# check
head(s@data)
plot(s)



# get polygons associated with map units that contain "amador" as a major component
q <- "select G.MupolygonWktWgs84 as geom, mapunit.mukey, muname
FROM mapunit
CROSS APPLY SDA_Get_MupolygonWktWgs84_from_Mukey(mapunit.mukey) as G
WHERE mukey IN (SELECT DISTINCT mukey FROM component WHERE compname like 'amador%' AND majcompflag = 'Yes')"

# result is a data.frame, "MupolygonWktWgs84" contains WKT representation of geometry
res <- SDA_query(q)
str(res)

# convert to SPDF
s <- processSDA_WKT(res)

# check: OK
head(s@data)

# map
par(mar=c(0,0,0,0))
map('county', 'California', xlim=c(-123.25, -118.75), ylim=c(36.5, 39))
plot(s, add=TRUE, col='royalblue', border='royalblue')
```


----------------------------
This document is based on `soilDB` version `r utils::packageDescription("soilDB", field="Version")`.
