---
title: "Component Interpretation Summary via SDA"
author: "D.E. Beaudette"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_caption: yes
    jquery: null
    mathjax: null
    smart: no
---


```{r setup, echo=FALSE, results='hide', warning=FALSE}
library(knitr, quietly = TRUE)
opts_chunk$set(message=FALSE, warning=FALSE, out.width="100%", fig.retina = 2, fig.align='center', dev='png', dev.args=list(pointsize=10, antialias='subpixel'), tidy=FALSE)
options(width=100, stringsAsFactors=FALSE)

knit_hooks$set(htmlcap = function(before, options, envir) {
  if(!before) {
    paste('<p class="caption" style="font-size:85%; font-style: italic; font-weight: bold;">',options$htmlcap,"</p><hr>",sep="")
    }
    })
```


You probably need to install some packages before this will work. Have a look at [this tutorial](http://ncss-tech.github.io/AQP/soilDB/SDA-tutorial.html) to get setup.

Lets get some interpretations that are relevant to MLRA 18 (Sierra Nevada Foothills) soils so that we can compare all components named [**Amador**](https://casoilresource.lawr.ucdavis.edu/sde/?series=amador) and [**Millvilla**](https://casoilresource.lawr.ucdavis.edu/sde/?series=millvilla). Just for fun lets throw in a very different soil from MLRA 17 (San Joaquin Valley); [**Hanford**](https://casoilresource.lawr.ucdavis.edu/sde/?series=hanford).

Note that SDA has a 100,000 row limit, so we have to be a little clever when writing the queries. In this case, letting the database filter out NULL fuzzy ratings and the reasons for ratings.
```{r}
library(soilDB)
library(plyr)
library(reshape2)
library(lattice)

# beware, there are hard limits (100k rows) on what can be returned by SDA
q <- "SELECT component.cokey, compname, mrulename, interplr, interplrc
FROM legend
INNER JOIN mapunit ON mapunit.lkey = legend.lkey
INNER JOIN component ON component.mukey = mapunit.mukey
INNER JOIN cointerp ON component.cokey = cointerp.cokey
WHERE
-- exclude STATSGO
areasymbol != 'US'
AND compname IN ('Millvilla', 'Amador', 'Hanford')
AND seqnum = 0
AND mrulename IN ('ENG - Construction Materials; Topsoil', 
'ENG - Sewage Lagoons', 'ENG - Septic Tank Absorption Fields', 
'ENG - Unpaved Local Roads and Streets', 
'AGR - California Revised Storie Index (CA)')
AND interplr IS NOT NULL;"

# query and check
x <- SDA_query(q)
kable(head(x))
```

OK, so how do the fuzzy numbers (`interplr` values) for these components compare?
```{r, fig.width=8, fig.height=8}
# compute number of rules and soils
n.soils <- length(unique(x$compname))
n.rules <- length(unique(x$mrulename))

# compare population with box-whisker plot
bwplot(compname ~ interplr | mrulename, data=x, layout=c(1,n.rules), strip=strip.custom(bg=grey(0.8)))
```


Those figures were useful, sometimes it is nice to see the median values for each interpretation (rows) and component (columns).
```{r}
s <- ddply(x, c('mrulename', 'compname'), plyr::summarize, 
           low=quantile(interplr, probs=0.05, na.rm=TRUE), 
           rv=quantile(interplr, probs=0.5, na.rm=TRUE), 
           high=quantile(interplr, probs=0.95, na.rm=TRUE))

knitr::kable(dcast(s, mrulename ~ compname, value.var = 'rv'), caption = "Median Fuzzy Ratings")
```


What about the categorical ratings? Note that the `kable` function from the **knitr** package makes nice HTML tables for us.
```{r}
knitr::kable(xtabs(~ interplrc + compname , data=x, subset= mrulename == 'ENG - Construction Materials; Topsoil'), caption="ENG - Construction Materials; Topsoil")

knitr::kable(xtabs(~ interplrc + compname , data=x, subset= mrulename == 'AGR - California Revised Storie Index (CA)'), caption = "AGR - California Revised Storie Index (CA)'")

knitr::kable(xtabs(~ interplrc + compname , data=x, subset= mrulename == 'ENG - Septic Tank Absorption Fields'), caption = "ENG - Septic Tank Absorption Fields")

knitr::kable(xtabs(~ interplrc + compname , data=x, subset= mrulename == 'ENG - Sewage Lagoons'), caption = "ENG - Sewage Lagoons")
```


## Interpretation Based Similarity

Here is a crazy idea, what if we could sort a collection of soil series based on a subset of relevant interpretations? We can, as long as some assumptions are made:

 1. there exists a small set of interpretations that reliably describe some aspect of "similarity"
 
 2. the mean fuzzy rating is a realistic description of central tendency  
 
 3. in aggregate, the collection of components named for a soil series will approximate the central tendency of that series
 
Lets try it with a small set of MLRA 17, 18, and 22A soils. NULL ratings will confound interpretation of the results--there are no "AGR - Pesticide Loss Potential-Leaching" ratings for components named "Dunstone". I have tried to select a small set of relevant interpretations, but clearly there are many possibilities. Soil profile sketches ([derived from Official Series Descriptions](http://ncss-tech.github.io/AQP/soilDB/soil-series-query-functions.html)) are used in the final figure. Note that these are not the same as the component data that were used to generate the interpretation-based similarity.
```{r, fig.width=10, fig.height=6}
library(cluster)
library(ape)
library(sharpshootR)

# set list of soil series (component names)
soil.list <- c('Pardee', 'Yolo', 'Capay', 'Aiken', 'Amador', 'Pentz', 'Sobrante',
'Argonaut', 'Toomes', 'Jocal', 'Holland', 'Auburn', 'Dunstone', 
'Hanford', 'Redding', 'Columbia', 'San Joaquin', 'Fresno')

# set list of relevant interpretations
interp.list <- c('ENG - Construction Materials; Topsoil', 
'ENG - Sewage Lagoons', 'ENG - Septic Tank Absorption Fields', 
'ENG - Unpaved Local Roads and Streets', 
'AGR - California Revised Storie Index (CA)')

# compose query
q <- paste0("SELECT compname, mrulename, AVG(interplr) as interplr_mean
FROM legend
INNER JOIN mapunit ON mapunit.lkey = legend.lkey
INNER JOIN component ON component.mukey = mapunit.mukey
INNER JOIN cointerp ON component.cokey = cointerp.cokey
WHERE 
-- exclude STATSGO
areasymbol != 'US'
AND compname IN ", format_SQL_in_statement(soil.list), "
AND seqnum = 0
AND mrulename IN ", format_SQL_in_statement(interp.list), "
AND interplr IS NOT NULL
GROUP BY compname, mrulename;")

# send query
x <- SDA_query(q)

# reshape long -> wide
x.wide <- dcast(x, compname ~ mrulename, value.var = 'interplr_mean')
knitr::kable(x.wide, digits = 3, caption="Mean Fuzzy Ratings for Select Soil Series")

# set row names to uppercase compname
# this will allow us to integrate with an SPC of OSD data later
row.names(x.wide) <- toupper(x.wide$compname)

# create distance matrix
# leave out compname
d <- daisy(x.wide[, -1])

# cluster via divisive hierachical method
# labels are recovered from row names -> distance matrix dimnames
h <- diana(d)

# fetch typical profiles from official series descriptions
osds <- fetchOSD(soil.list)

# magic happens here
# hang profile sketches from interpretation-based clustering
# may have to tinker with arguments to get figure right
par(mar=c(2,1,2,1))
plotProfileDendrogram(osds, clust = diana(d), scaling.factor = 0.008, max.depth=200, width = 0.2)
title('Component Similarity via Interpretation Fuzzy Values')
mtext('profile sketches from OSD', side = 1)
```

Interesting. Discussion to be continued...


-------------------------------------
This document is based on `soilDB` version `r utils::packageDescription("soilDB", field="Version")`.

