---
output:
  html_document:
    fig_caption: yes
    jquery: null
    mathjax: null
    smart: no
---

```{r setup, echo=FALSE, results='hide', warning=FALSE}
# setup
library(knitr, quietly=TRUE)
# library(printr, quietly=TRUE)
opts_chunk$set(message=FALSE, warning=FALSE, background='#F7F7F7', fig.align='center', fig.retina=1, dev='png', tidy=FALSE, verbose=FALSE)
options(width=100, stringsAsFactors=FALSE)
```

Getting, Plotting, and Saving Soil Series Extent Data from SoilWeb
===============================================================
`r format(Sys.time(), "%Y-%m-%d")`
<br>Dylan Beaudette

### Introduction
This document demonstrates how to use the [soilDB](http://ncss-tech.github.io/AQP/soilDB/soilDB-Intro.html) package to access detailed soil series extent maps via SoilWeb. These maps can be directly displayed in **R**, overlayed on Google Maps, or saved as local files (e.g. shapefiles). Data returned from SoilWeb represent bounding-boxes that enclose SSURGO polygons associated with map units containing the search criteria. Bounding-boxes are snapped to 0.01 degree precision in order to reduce processing time and file size. Note that these files are cached server-side after the first request, and the cache is purged when SoilWeb is periodically synced to the official data.

### Installation
With a recent version of R, it should be possible to get all of the packages that this tutorial depends on via:
```{r install-deps, eval=FALSE}
# run these commands in the R console
install.packages('soilDB', dep=TRUE) # stable version from CRAN + dependencies
install.packages('dismo', dep=TRUE)
install.packages('RColorBrewer', dep=TRUE)
install.packages('maps', dep=TRUE)
```

### Examples 
**Illustrate the extent of SSURGO map units associated with the Amador series.**
```{r libraries, results='hide'}
# load required libraries
library(soilDB)
library(rgdal)
library(dismo)
library(maps)
library(RColorBrewer)
```

```{r example-1, eval=FALSE}
# plot map
seriesExtentAsGmap('amador')
```

**Illustrate the extent of SSURGO map units associated with several soil series from MLRA 17 and 18.**
```{r example-2, results='hide'}
# define some nice colors
cols <- brewer.pal('Set1', n=3)

# get the spatial extent of select series
amador <- seriesExtent('amador')
pardee <- seriesExtent('pardee')
san.joaquin <- seriesExtent('san joaquin', timeout=120)

# plot the results:
# set figure margins to 0
par(mar=c(0,0,0,0))

# plot select counties from California
map(database='county', regions='california,calaveras|tuolumne|amador|san joaquin|stanislaus|merced|mariposa|sacramento')

# add each soil series extent object
plot(amador, col=cols[1], add=TRUE)
plot(pardee, col=cols[2], add=TRUE)
plot(san.joaquin, col=cols[3], add=TRUE)

# add a neatline around the map
box()

# make a simple legend
legend('topright', col=cols, pch=15, legend=c('Amador','Pardee','San Joaquin'))
```


**Investigate the structure of soil series extent data.**
```{r example-3}
# internal structure / class
str(amador, 2)

# coordinate reference system in PROJ4 notation
proj4string(amador)

# extract acreage, details: http://casoilresource.lawr.ucdavis.edu/see/
amador$acres
pardee$acres
san.joaquin$acres
```

**Save soil series extent data in multiple formats.**
```{r example-4, eval=FALSE}
# save using the coordinate reference system associated with this object (GCS WGS84)
writeOGR(amador, dsn='.', layer='amador-extent', driver='ESRI Shapefile')

# save as KML for use in Google Earth
writeOGR(amador, dsn='amador-extent.kml', layer='amador', driver='KML')

# project to UTM zone 10 NAD83 and save
amador.utm <- spTransform(amador, CRS('+proj=utm +zone=10 +datum=NAD83'))
writeOGR(amador.utm, dsn='.', layer='amador-extent-utm', driver='ESRI Shapefile')
```

----------------------------
This document is based on `aqp` version `r utils::packageDescription("aqp", field="Version")` and `soilDB` version `r utils::packageDescription("soilDB", field="Version")`.