---
title: "Soil Series Climate Summary Evaluation"
author: "D.E. Beaudette"
date: "`r Sys.Date()`"
output:
  html_document:
  mathjax: null
jquery: null
smart: no
---
  
  
```{r setup, echo=FALSE, results='hide', warning=FALSE}
# setup
library(knitr, quietly=TRUE)
library(kableExtra, quietly=TRUE)
opts_chunk$set(message=FALSE, warning=FALSE, background='#F7F7F7', fig.align='center', fig.retina=2, dev='png', tidy=FALSE, verbose=FALSE)
options(width=100, stringsAsFactors=FALSE)
```

# Introduction


## Setup
With a recent version of R (>= 2.15), it is possible to get all of the packages that this tutorial depends on via:
```{r install-deps, eval=FALSE}
# stable packages from CRAN
install.packages('aqp', dep=TRUE)
install.packages('soilDB', dep=TRUE)
install.packages('sharpshootR', dep=TRUE)
install.packages('latticeExtra', dep=TRUE)
install.packages('ggplot2', dep=TRUE)
install.packages('raster', dep=TRUE)

# latest versions from GitHub
devtools::install_github("ncss-tech/soilDB", dependencies=FALSE, upgrade_dependencies=FALSE)
devtools::install_github("ncss-tech/sharpshootR", dependencies=FALSE, upgrade_dependencies=FALSE)
```


## Using this Tutorial
This tutorial can be used as a stand-alone demonstration of the functionality provided by `fetchOSD` and `OSDquery`, or as a template for your own project. Copy any paste blocks of code *sequentially* into your own script if you would like to follow-along.

First, you will need to load some packages. Be sure to install these if missing, and install the latest versions of `soilDB` and `sharpshootR` as much of this tutorial depends on recent updates.
```{r eval=TRUE}
library(soilDB)
library(raster)
library(maps)
library(sharpshootR)
library(latticeExtra)
library(ggplot2)
library(RColorBrewer)
```




# Climate Summaries


## Estimates from a single point per map unit polygon

Annual and monthly climate summaries have been estimated from the [SSR2 standard stack of 1981–2010 PRISM data](https://ncss-tech.github.io/soilReports/docs/region2_mu-comparison-data-sources.html). For now, percentiles are estimates from a single sample from within each map unit polygon, weighted by $log(polygon\ area * component\ percentage)$.

```{r fig.width=10, fig.height=6}
soils <- c('argonaut', 'pierre', 'zook', 'cecil')
s <- fetchOSD(soils, extended = TRUE)
```

Here is a preview of the data: 8 annual summaries and monthly summaries for PPT and PET.
```{r echo=FALSE}
kable_styling(kable(s$climate.annual[1:8, ], format='html', digits = 2), font_size = 9, full_width = FALSE)
kable_styling(kable(s$climate.monthly[c(1:3, 13:15), ], format='html', digits = 2), font_size = 9, full_width = FALSE)
```


Visualization of the annual climate summaries using select percentiles.
```{r fig.width=12, fig.height=5}
segplot(factor(series) ~ q05 + q95 | climate_var, centers=q50, data=s$climate.annual, draw.bands=FALSE, segments.fun=panel.arrows, ends='both', angle=90, length=1, unit='mm', scales=list(y=list(alternating=3), x=list(relation='free')), as.table=TRUE, col='RoyalBlue', strip=strip.custom(bg=grey(0.85), par.strip.text=list(cex=0.7)), xlab='5th-25th-50th-75th-95th Percentiles', panel=function(x, y, z, q25=s$climate.annual$q25, q75=s$climate.annual$q75, subscripts, ...) {
  # basic plot
  panel.grid(h=FALSE, v=-1, col='grey', lty=3)
  panel.abline(h=1:length(unique(s$climate.annual$series)), col='grey', lty=3)
  panel.segplot(x, y, z, subscripts=subscripts, ...)
  
  # add interquartile range
  q25 <- q25[subscripts]
  q75 <- q75[subscripts]
  zz <- z[subscripts]
  panel.rect(xleft=q25, xright=q75, ybottom=as.numeric(zz) - 0.025, ytop=as.numeric(zz) + 0.025, border='RoyalBlue', col='RoyalBlue')
})
```


One possible depiction of monthly PPT and PET.
```{r fig.width=12, fig.height=5}
# reasonable colors for a couple of groups
cols <- brewer.pal(9, 'Set1') 
cols <- cols[c(1:5,7,9)]

ggplot(s$climate.monthly, aes(x = month, group=series)) + 
  geom_ribbon(aes(ymin = q25, ymax = q75, fill=series)) + 
  geom_line(aes(month, q25)) + 
  geom_line(aes(month, q75)) + 
  geom_abline(intercept=0, slope = 0, lty=2) +
  xlab('') + ylab('mm') + 
  ggtitle('Monthly IQR') +
  scale_fill_manual(values=alpha(cols, 0.75)) +
  facet_wrap(vars(variable), scales = 'free_y') +
  theme_bw()
```


## Estimates from SoilWeb series extent boundary

```{r}
# the result is a SpatialPolygonsDataFrame object
see.boundary <- seriesExtent('amador')
# collect as many samples as are present in the fetchOSD summary
see.samples <- spsample(see.boundary, n = 394, type='regular')

# quick map of samples
par(mar=c(0,0,0,0))
map(database='county', regions='california')
points(see.samples, col='royalblue', cex=0.25)
mtext('SEE boundary, 394 samples', side = 1, line = -1.25, at = -124.25, adj = 0)
box()
```

Sample raster stack and compute select percentiles
```{r}


```



## Estimates from sampling original map unit polygons
```{r}
# get polygons associated with map units that contain "amador" as a major component
q <- "select G.MupolygonWktWgs84 as geom, mapunit.mukey, muname
FROM mapunit
CROSS APPLY SDA_Get_MupolygonWktWgs84_from_Mukey(mapunit.mukey) as G
WHERE mukey IN (SELECT DISTINCT mukey FROM component WHERE compname like 'amador%' AND majcompflag = 'Yes')"

# result is a data.frame, "MupolygonWktWgs84" contains WKT representation of geometry
res <- SDA_query(q)

# convert to SPDF
mu.poly <- processSDA_WKT(res)

# map
par(mar=c(0,0,0,0))
map(database='county', regions='california')
plot(mu.poly, add=TRUE, border='royalblue', col='royalblue')
mtext('MU polygons', side = 1, line = -1.25, at = -124.25, adj = 0)
box()

```



----------------------------
This document is based on `aqp` version `r utils::packageDescription("aqp", field="Version")`, `soilDB` version `r utils::packageDescription("soilDB", field="Version")`, and `sharpshootR` version `r utils::packageDescription("sharpshootR", field="Version")`.
