---
title: "Soil Series Co-Occurrence Data"
author: "D.E. Beaudette"
date: "`r Sys.Date()`"
output:
  html_document:
    mathjax: null
    jquery: null
    smart: no
---


```{r setup, echo=FALSE, results='hide', warning=FALSE}
# setup
library(knitr, quietly=TRUE)
library(kableExtra, quietly=TRUE)

opts_chunk$set(
  message = FALSE, 
  warning = FALSE, 
  background = '#F7F7F7', 
  fig.align = 'center', 
  dev = 'svglite', 
  tidy = FALSE, 
  verbose = FALSE
)

options(width = 100, stringsAsFactors = FALSE)
```


TODO: 

* write definitions
* demonstrate major/minor siblings
* extend with joint area calculation


Recreate functionality of previous API.
```{r, fig.width=10}
library(soilDB)
library(sharpshootR)
library(latticeExtra)
library(reshape2)
library(cluster)
library(ape)
library(corrplot)

# new function as of 2.2-8
# get sibling data for Amador soil series via SoilWeb
# note that sib$sib may contain duplicates, based on major/minor component status
s <- siblings('amador', component.data = TRUE)

str(s)

# convert into adjacency matrix
m <- component.adj.matrix(
  s$sib.data[s$sib.data$compkind == 'Series', ], 
  mu = 'mukey', 
  co = 'compname', 
  wt = 'comppct_r'
)
```

A simple depiction of the resulting adjacency matrix. Cell values are adjacency weight (strength of association).
```{r fig.width=10}
.cp <- c('white', hcl.colors(25))

corrplot(
  m, 
  col = .cp, 
  is.corr = FALSE, 
  diag = TRUE,
  col.lim = c(0, 1), 
  method = "color",
  type = "upper", 
  tl.pos = "td",
  tl.cex = 0.8,
  tl.col = 'black',
  addgrid.col = 'black'
) 

```

Adjacency as a network.
```{r fig.width=10}
# plot network diagram, with Amador soil highlighted
par(mar = c(0.5,0.5,0.5,0.5))

plotSoilRelationGraph(
  m, 
  s = 'Amador', 
  vertex.scaling.factor = 2, 
  edge.transparency = 0.75, 
  edge.col = grey(0.85), 
  edge.highlight.col = 'black', 
  vertex.label.family = 'sans'
)
```


Sibling type and counts.
```{r echo=FALSE}
kableExtra::kable_styling(knitr::kable(s$sib, format = 'html'), full_width = FALSE, font_size = 10)
```


Split sibling lists and get basic morphology from OSDs.
```{r}
idx <- which(s$sib$majcompflag)
major.siblings <- c(s$sib$series[1], s$sib$sibling[idx])
all.siblings <- unique(c(s$sib$series[1], s$sib$sibling))

# get basic OSD data for queried series and siblings
h.major <- fetchOSD(major.siblings)
h.all <- fetchOSD(all.siblings)
```


Siblings that are major components.
```{r fig.width=7}
SoilTaxonomyDendrogram(h.major, depth.axis = list(line = -4))
```

All siblings.
```{r fig.width=10}
SoilTaxonomyDendrogram(h.all, width = 0.3, depth.axis = list(line = -3.5))
```


Cousins are siblings of siblings. This can generate a lot of output.
```{r}
# takes a while to run, there will be duplicates in counsins here
s <- siblings('amador', component.data = TRUE, cousins = TRUE)

# combine sibling + cousin data, remove duplicates
d <- unique(rbind(s$sib.data, s$cousin.data))

# subset to components that are correlated to a soil series
d <- d[which(d$compkind == 'Series'), ]

# convert into adjacency matrix
m <- component.adj.matrix(d, mu = 'mukey', co = 'compname', wt = 'comppct_r')
```

Adjaceny matrix.
```{r fig.width=15, fig.height=15}
corrplot(
  m, 
  col = .cp, 
  is.corr = FALSE, 
  diag = TRUE,
  col.lim = c(0, 1), 
  method = "color",
  type = "upper", 
  tl.pos = "td",
  tl.cex = 0.8,
  tl.col = 'black',
  addgrid.col = 'black'
) 
```

Adjaceny network.
```{r fig.width=15, fig.height=15}
# plot network diagram, with Amador soil highlighted
par(mar=c(1,1,1,1))

plotSoilRelationGraph(
  m, 
  s = 'Amador',
  vertex.scaling.factor = 2, 
  edge.transparency = 0.75, 
  edge.col = grey(0.85), 
  edge.highlight.col = 'black', 
  vertex.label.family = 'sans'
)
```


Try on your own, this is too much data to "see" all in one figure.
```{r eval=FALSE}
# fetch and convert data into an SPC
h <- fetchOSD(c(s$sib$series[1], unique(s$cousins$sibling)))

# plot dendrogram + profiles
SoilTaxonomyDendrogram(h, width = 0.3)
```

## Geomorphic Summaries

```{r}
s.all.siblings <- fetchOSD(all.siblings, extended = TRUE)
```

Compare hillslope position.
```{r fig.width=10, fig.height=6, echo=FALSE}
res <- vizHillslopePosition(s.all.siblings$hillpos, s = 'AMADOR')
print(res$fig)
```

Compare geomorphic component.
```{r fig.width=10, fig.height=6, echo=FALSE}
res <- vizGeomorphicComponent(s.all.siblings$geomcomp, s = 'AMADOR')
print(res$fig)
```

## MLRA and Parent Material Summaries
TODO

## Investigate Spatial Patterns
TODO

## KSSL Data
TODO


----------------------------
This document is based on `aqp` version `r utils::packageDescription("aqp", field="Version")`, `soilDB` version `r utils::packageDescription("soilDB", field="Version")`, and `sharpshootR` version `r utils::packageDescription("sharpshootR", field="Version")`.

