---
title: "Visualizing Soil Temperature Regime"
author: "D.E. Beaudette"
date: "`r Sys.Date()`"
output:
  html_document:
  mathjax: null
jquery: null
smart: no
---
  
```{r setup, echo=FALSE, results='hide', warning=FALSE}
library(knitr, quietly=TRUE)
opts_chunk$set(message=FALSE, warning=FALSE, background='#F7F7F7', fig.align='center', fig.retina=2, dev='png', tidy=FALSE, verbose=FALSE)
options(width=100, stringsAsFactors=FALSE)
```

## Background
The `estimateSTR()` function from the `soilDB` package applies a [soil temperature regime](https://en.wikipedia.org/wiki/USDA_soil_taxonomy#Soil_Temperature_Regimes_(STR)) (STR) classification using rules from Soil Taxonomy (12th edition). Accurate classificaiton of STR depends on the following properties:

   * MAST: mean annual soil temperature (degrees C, at 50cm)
   * MSST: mean summer soil temperature (degrees C, at 50cm)
   * MWST: mean winter soil temperature (degrees C, at 50cm)
   * O horizon presence / absense
   * permafrost presence / absense

The function will generate resonable, but possibly wrong, results if missing O horizon, saturation, and permafrost values. This is only an issue for the gelic / cryic / frigid STR, e.g. in very cold climates.


STR include:

   * gelic (suggests `permafrost` argument)
   * cryic (suggests `permafrost`, `O.hz`, and `saturated` arguments)
   * frigid (suggests `O.hz`, and `saturated` arguments)
   * mesic
   * thermic
   * hyperthermic
   * isofrigid
   * isomesic
   * isothermic
   * isohyperthermic


## Setup
You will need the latest version of `soilDB`:
```{r eval=FALSE}
# install.packages('devtools', dep=TRUE)
devtools::install_github("ncss-tech/soilDB", dependencies=FALSE, upgrade_dependencies=FALSE)
```


## Basic Usage
Note that this function is vectorized and works as expected, as long as each argument is the same length.


```{r, fig.width=9, fig.height=7}
library(soilDB)
```


```{r, fig.width=9, fig.height=7}
estimateSTR(mast = 12, mean.summer = 15, mean.winter = 8)
estimateSTR(mast = 4, mean.summer = 8, mean.winter = 1, O.hz = TRUE, saturated = FALSE, permafrost = FALSE)
```

## Graphical Summaries
These are quite dense. Some notes:

   * The "O horizon / Saturation" flags come into play at the cryic / frigid break. "X" means "no", "O" means O horizon present, and "S" means "saturation" as specified in the keys.
   * Rows with a "?" describe situations in which STR is estimated by splitting the difference in temperature thresholds when O horizon and / or Saturation values are missing. 


```{r, fig.width=9, fig.height=7}
par(mar=c(4,1,0,1))
STRplot(mast = 0:25, msst = 10, mwst = 1)
STRplot(mast = 0:25, msst = 9, mwst = 1)
STRplot(mast = 0:25, msst = 7, mwst = 1)
STRplot(mast = 0:25, msst = 6, mwst = 1)
STRplot(mast = 0:25, msst = 5, mwst = 1)
```

```{r, fig.width=9, fig.height=7}
par(mar=c(4,1,0,1))
STRplot(mast = 5, msst = 1:18, mwst = 0)
STRplot(mast = 5, msst = 1:18, mwst = 1)
STRplot(mast = 5, msst = 1:18, mwst = 2)
STRplot(mast = 5, msst = 1:18, mwst = 3)
STRplot(mast = 5, msst = 1:18, mwst = 5)
```

```{r, fig.width=9, fig.height=7}
par(mar=c(4,1,0,1))
STRplot(mast = 10, msst = 1:18, mwst = 5)
```



----------------------------
This document is based on `aqp` version `r utils::packageDescription("aqp", field="Version")` and `soilDB` version `r utils::packageDescription("soilDB", field="Version")`.

