---
title: "Map Unit Key Web Coverage Service"
date: "`r Sys.Date()`"
output:
  html_document:
    mathjax: null
    jquery: null
    smart: no
    number_sections: no
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: false
---
  
  
```{r setup, echo=FALSE, results='hide', warning=FALSE}
# setup
library(knitr, quietly=TRUE)
library(kableExtra, quietly=TRUE)
opts_chunk$set(message=FALSE, warning=FALSE, background='#F7F7F7', fig.align='center', fig.retina=2, dev='png', tidy=FALSE, verbose=FALSE)
options(width=100, stringsAsFactors=FALSE)
```


# Introduction
A web coverage service (WCS) is provided for the gSSURGO and gNATGSO [map unit key grids](https://nrcs.app.box.com/v/soils/folder/149859400396). These grids represent a rasterizaion of the gSSURGO and gNATSGO map unit keys for the conterminous United States at a resolution of 30m, referenced to an Albers Equal Area Conic (NAD83) coordinate reference system (EPSG:5070). Grids are LZW compressed and internally tiled for efficient random access. Cell values are map unit keys, encoded as unsigned 32-bit integers. The grid topology and cell values are identical to the rasters contained within the gSSURGO and gNATSGO file geo-databases (FGDB). The GeoTiff format is used to ensure maximum compatibility. Files are re-created as part of the annual SSURGO refresh cycle.

See the [related tutorial on using the CONUS mukey grids directly]([mukey web coverage service](http://ncss-tech.github.io/AQP/soilDB/WCS-demonstration-01.html)).

An experimental WCS is provided for raster soil survey (RSS) products at 10m resolution.

A similar WCS is provided for the [ISSR-800 soil property grids](https://casoilresource.lawr.ucdavis.edu/soil-properties/).

## Grid Selection
See the [`mukey.wcs`](http://ncss-tech.github.io/soilDB/docs/reference/mukey.wcs.html) and [`ISSR800.wcs`](http://ncss-tech.github.io/soilDB/docs/reference/ISSR800.wcs.html) manual pages for details.

Basic usage.
```{r eval = FALSE}
# select gSSURGO grid, 30m resolution
x <- mukey.wcs(aoi = aoi, db = 'gssurgo', ...)

# select gNATSGO grid, 30m resolution
x <- mukey.wcs(aoi = aoi, db = 'gnatsgo', ...)

# select RSS grid, 10m resolution
x <- mukey.wcs(aoi = aoi, db = 'RSS', ...)

# select various ISSR-800 grids, details below
x <- ISSR800.wcs(aoi = aoi, var = 'paws')
```


## gSSURGO
Excerpt from the [gSSURGO documentation](https://www.nrcs.usda.gov/wps/portal/nrcs/detail/soils/home/?cid=nrcs142p2_053628#release).

<div style="padding: 10px; font-size:90%;">
The gSSURGO Database is derived from the official Soil Survey Geographic (SSURGO) Database. SSURGO generally has the most detailed level of soil geographic data developed by the National Cooperative Soil Survey (NCSS) in accordance with NCSS mapping standards. The tabular data represent the soil attributes and are derived from properties and characteristics stored in the National Soil Information System (NASIS). The gSSURGO data were prepared by merging the traditional vector-based SSURGO digital map data and tabular data into statewide extents, adding a statewide gridded map layer derived from the vector layer, and adding a new value-added look up table (Valu1) containing “ready to map” attributes. The gridded map layer is a file geodatabase raster in an ArcGIS file geodatabase. The raster and vector map data have a statewide extent. The raster map data have a 10-meter cell size that approximates the vector polygons in an Albers Equal Area projection. Each cell (and polygon) is linked to a map unit identifier called the map unit key. A unique map unit key is used to link the raster cells and polygons to attribute tables. Due to file size, the raster layer for the conterminous United States is only available in a 30-meter resolution.
</div>

## gNATSGO
Excerpt from the [gNATSGO documentation](https://www.nrcs.usda.gov/wps/portal/nrcs/detail/soils/survey/geo/?cid=nrcseprd1464625).

<div style="padding: 10px; font-size:90%;">
The gNATSGO databases contain a raster of the soil map units and 70 related tables of soil properties and interpretations. They are designed to work with the SPSD gSSURGO ArcTools. Users can create full coverage thematic maps and grids of soil properties and interpretations for large geographic areas, such as the extent of a State or the conterminous United States. Please note that the State-wide geodatabases contain a 10 meter raster and the CONUS database contains a 30 meter raster.

The gNATSGO database is composed primarily of SSURGO data, but STATSGO2 data was used to fill in the gaps. The RSSs are newer product with relatively limited spatial extent.  These RSSs were merged into the gNATSGO after combining the SSURGO and STATSGO2 data. The extent of RSS is expected to increase in the coming years.
</div>


## Raster Soil Survey
Excerpt from the [RSS documentation](https://www.nrcs.usda.gov/wps/portal/nrcs/detail/soils/survey/geo/?cid=nrcseprd1759022).

<div style="padding: 10px; font-size:90%;">
Raster Soil Survey is a reference to the products of soil survey work completed using digital soil mapping methodologies. Digital soil mapping is the production of georeferenced soil databases based on the quantitative relationships between soil measurements made in the field or laboratory and environmental data and may be represented as either discrete classes or continuous soil properties. Both digital and traditional soil mapping use a conceptual soil-landscape model as a means for organizing environmental information into discrete divisions. The primary difference between these two approaches is that digital methods exploit quantitative relationships of the environmental information, while traditional methods utilize a more subjective approach and the approximate relationships of the environmental information to spatially represent where the divisions are represented.
</div>


## Thematic Mapping
Thematic mapping or analysis of soil information requires connecting the grids to our tabular data sources, either using local files or Soil Data Access (SDA) web-service. The `soilDB` package provides many convenient interfaces to SDA. Note that SDA does not yet contain tabular data for the raster soil surveys.


## Caveats / Limitations
  * Requests are limited to images sizes of 5000x5000 pixels, this is approximately 1x1 degrees at 30m.
  * These instructions are based on the development version of [`soilDB`](https://github.com/ncss-tech/soilDB/), which has [recently switched over to `sf` and `terra` packages](https://github.com/ncss-tech/soilDB/blob/master/NEWS.md).


## Manually Creating bounding-boxes for WCS requests
The WCS interface functions `mukey.wcs` and `ISSR800.wcs` can automatically generate a bounding-box (BBOX) from most spatial data formats. However, sometimes it is more convenient to manually specify a BBOX created from a website or single point specified in WGS84 coordinates. [SoilWeb](https://casoilresource.lawr.ucdavis.edu/gmap/) provides two convenient keyboard shortcuts, available after clicking on the map interface:

   * 'b' copy bounding-box, returned as: <span style = 'font-size: 80%;'><tt>-118.6609 36.4820,-118.6609 36.5972,-118.3979 36.5972,-118.3979 36.4820,-118.6609 36.4820</tt></span>
   
   * 'p' copy link to center coordinate, returned as: `https://casoilresource.lawr.ucdavis.edu/gmap/?loc=36.53964,-118.52943,z13`
   
Right-clicking anywhere in the map interface will also generate a link to those coordinates and zoom-level.

A SoilWeb-style BBOX can be directly converted into an `sf` object:
```{r fig.width = 6.5, fig.height=5}
library(sf)
library(soilDB)
library(terra)

# paste your BBOX text here
bb <- '-118.6609 36.4820,-118.6609 36.5972,-118.3979 36.5972,-118.3979 36.4820,-118.6609 36.4820'

# convert text -> WKT -> sf
wkt <- sprintf('POLYGON((%s))', bb)
x <- st_as_sfc(wkt)

# set coordinate reference system as GCS/WGS84
st_crs(x) <- 4326

# query WCS
m <- mukey.wcs(x, db = 'gSSURGO')

# looks OK
plot(m, legend = FALSE, axes = FALSE, main = attr(m, 'layer name'))
box()

# add original BBOX, after transforming to mukey grid CRS
plot(st_transform(x, 5070), add = TRUE)
```


A buffer applied to a [single WGS84 coordinate](https://casoilresource.lawr.ucdavis.edu/gmap/?loc=36.52578,-118.55639,z13) can create a BBOX:
```{r fig.width = 5, fig.height = 5}
library(sf)
library(soilDB)

# example point, WGS84 coordinates
p <- st_as_sf(
  data.frame(
    x = -118.55639,
    y = 36.52578
  ),
  coords = c('x', 'y'),
  crs = 4326
)

# 1000m buffer applied to WGS84 coordinate
# radius defined in meters
b <- st_buffer(p, dist = units::set_units(1000, m))
b <- st_as_sf(b)

# query WCS
# result is in EPSG:5070
m <- mukey.wcs(b, db = 'gSSURGO')

# looks OK
plot(m, legend = FALSE, axes = FALSE, main = attr(m, 'layer name'))
box()

# add buffer, after transforming to mukey grid CRS
plot(st_transform(b, 5070), add = TRUE)

# add original point, after transforming to mukey grid CRS
plot(st_transform(p, 5070), add = TRUE, pch = 16)
```









# Setup
Need the latest version of `soilDB` and `terra`, along with all packages specified below (CRAN versions fine).
```{r eval = FALSE}
install.packages('soilDB')
remotes::install_github("ncss-tech/soilDB", dependencies = FALSE, upgrade = FALSE, build = FALSE)

# terra
install.packages('terra', repos='https://rspatial.r-universe.dev')
```

Load required packages.
```{r}
# latest from GitHub
library(soilDB)

# wrangling polygons and CRS transformations
library(sf)

# raster data / analysis
# need latest: install.packages('terra', repos='https://rspatial.r-universe.dev')
library(terra)

# raster visualization
library(rasterVis)
library(viridisLite)

# soil classification
library(aqp)

# color palettes and manipulation
library(RColorBrewer)
library(colorspace)
```


# Map Unit Key Grids

Use the `mukey.wcs()` function to access chunks of the CONUS map unit key grids, based on some representation of an area of interest (AOI). The AOI can be defined manually, as below, or automatically extracted from `sf`, `sfc`, `bbox`, `SpatRaster`, or `SpatVector` objects. Backwards-compatibility with `Spatial*` and `RasterLayer` objects is also provided. The resulting grid of integers isn't all that useful by itself; join data from Soil Data Access (SDA) or local files (by map unit key) to create thematic maps. Note: the gSSURGO and gNATSGO grids are updated annually, and typically in-sync with the live version of the data hosted by SDA (updated Oct 1st) by early November. Map unit keys can change over time, especially in soil survey areas that were updated during the last fiscal year.
```{r}
# make a bounding box and assign a CRS (4326: GCS, WGS84)
a <- st_bbox(
  c(xmin = -114.16, xmax = -114.08, ymin = 47.65, ymax = 47.68), 
  crs = st_crs(4326)
)

# fetch gSSURGO map unit keys at native resolution (30m)
x <- mukey.wcs(aoi = a, db = 'gssurgo')

# check:
print(x)

# looks good
plot(
  x, 
  main = 'gSSURGO map unit keys',
  sub = 'Albers Equal Area Projection',
  axes = FALSE, 
  legend = FALSE, 
  col = viridis(100)
)
```



## SSURGO Polygons from SDA
It is possible to retrieve small areas of vector geometry (SSURGO polygons) from SDA with the `SDA_spatialQuery()` function. These data are stored and delivered in a geographic coordinate system (WGS84). Overlaying the SSURGO polygons and map unit key grids will require a simple transformation.
```{r}
# get intersecting SSURGO polygons from SDA
# result is a SpatVector object
# GCS WGS84
p <- SDA_spatialQuery(x, what = 'mupolygon', geomIntersection = TRUE)

# transform to AEA coordinate reference system used by gSSURGO / gNATSGO
# this is EPSG:5070
p <- project(p, crs(x))

par(mar = c(1, 0, 2, 0))
plot(p, main = 'SSURGO Polygons (SDA)', axes = FALSE)
mtext('Albers Equal Area Projection', side = 1, line = -0.5)
```


Overlay SSURGO polygons and 30m map unit key grid.
```{r}
par(mar = c(1, 0, 2, 0))
plot(x, main = 'gSSURGO Grid (WCS)\nSSURGO Polygons (SDA)', col = viridis(50), axes = FALSE, legend = FALSE)
plot(p, add = TRUE, border = 'white')

mtext('CONUS Albers Equal Area Projection (EPSG:5070)', side = 1, line = 1)
```


## Grid Resolution Specification
Requesting map unit key grids at a resolution other than 30m is possible, but only suitable for a quick "preview" of the data. For example, it is possible to get a much larger chunk of data by requesting grids at 800m. However, pixels are selected by nearest-neighbor and not generalized to the coarser scale. In short: don't use the resulting data for anything other than a simple preview.

```{r}
# part of CA
# make a bounding box and assign a CRS (4326: GCS, WGS84)
a.CA <- st_bbox(
  c(xmin = -121, xmax = -120, ymin = 37, ymax = 38), 
  crs = st_crs(4326)
)

# fetch gSSURGO map unit keys at ~~ 800m
# nearest-neighbor resampling = this is a "preview"
# result is a SpatRaster object
x.800 <- mukey.wcs(aoi = a.CA, db = 'gssurgo', res = 800)

# OK
plot(
  x.800, 
  main = 'A Preview of gSSURGO Map Unit Keys',
  sub = 'Albers Equal Area Projection (800m)\nnearest-neighbor resampling',
  axes = FALSE, 
  legend = FALSE, 
  col = viridis(100),
  )
```



## Raster Soil Survey Data
```{r fig.width=12, fig.height=6}
# slightly larger than Coweeta RSS BBOX
# specified in EPSG:5070
a <- st_bbox(
  c(xmin = 1129000, xmax = 1135000, ymin = 1403000, ymax = 1411000), 
  crs = st_crs(5070)
)

# convert to sf polygon for later use in figures
a <- st_as_sfc(a)

# gSSURGO grid: 30m resolution
(x <- mukey.wcs(a, db = 'gSSURGO', res = 30))

# gNATSGO grid: 300m resolution
(y <- mukey.wcs(a, db = 'gNATSGO', res = 30))

# RSS grid: 10m resolution
(z <- mukey.wcs(a, db = 'RSS', res = 10))

# graphical comparison
par(mfcol = c(1, 3))

# gSSURGO
plot(x, axes = FALSE, legend = FALSE, main = attr(x, 'layer name'))
# original BBOX
plot(a, add = TRUE)

# gNATSGO
plot(y, axes = FALSE, legend = FALSE, main = attr(y, 'layer name'))
# original BBOX
plot(a, add = TRUE)

# RSS
plot(z, axes = FALSE, legend = FALSE, main = attr(z, 'layer name'), ext = x)
# original BBOX
plot(a, add = TRUE)
```



# Thematic Mapping


## Several horizon-level soil properties
The [`get_SDA_property()`](http://ncss-tech.github.io/soilDB/reference/get_SDA_property.html) function from `soilDB` is convenient interface to aggregated SSURGO/STATSGO tabular data via SDA.

Get mukey grid.
```{r}
# https://casoilresource.lawr.ucdavis.edu/gmap/?loc=36.57666,-96.70175,z14
# make a bounding box and assign a CRS (4326: GCS, WGS84)
a <- st_bbox(
  c(xmin = -96.7696, xmax = -96.6477, ymin = 36.5477, ymax = 36.6139), 
  crs = st_crs(4326)
)

# fetch gSSURGO map unit keys at native resolution (~30m)
mu <- mukey.wcs(aoi = a, db = 'gssurgo')

# check: OK
plot(
  mu, 
  legend = FALSE, 
  axes = FALSE, 
  col = viridis(100),
  main = 'gSSURGO Map Unit Key Grid'
)
```

Derive aggregate soil properties, merge with raster attribute table (RAT).
```{r}
# extract RAT for thematic mapping
rat <- cats(mu)[[1]]

# variables of interest
vars <- c("dbthirdbar_r", "awc_r", "ph1to1h2o_r")

# get / aggregate specific horizon-level properties from SDA
# be sure to see the manual page for this function
p <-  get_SDA_property(property = vars,
                       method = "Dominant Component (Numeric)", 
                       mukeys = as.integer(rat$mukey),
                       top_depth = 0,
                       bottom_depth = 25)


# check: OK
head(p)

# convert areasymbol into a factor easy plotting later
p$areasymbol <- factor(p$areasymbol)

# merge aggregate soil data into RAT
rat <- merge(rat, p, by.x = 'mukey', by.y = 'mukey', sort = FALSE, all.x = TRUE)
levels(mu) <- rat

# result is a grid of map unit keys + RAT
```

Plot `areasymbol`.
```{r fig.width = 6, fig.height = 4}
# list variables in the RAT
names(cats(mu)[[1]])

# select 'areasymbol'
activeCat(mu) <- 'awc_r'
plot(mu)
box()
```

Plot aggregated soil properties.
```{r}
# reset the active category to 1
# this will ensure that all variables in the RAT will be converted by catalyze(mu)
activeCat(mu) <- 1

# convert mukey grid + RAT -> stack of numerical grids
mu.stack <- catalyze(mu)

# keep only properties / remove IDs
mu.stack <- mu.stack[[vars]]


# note implicit simplification via maxpixels
levelplot(
  mu.stack[['dbthirdbar_r']], 
  main = '1/3 Bar Bulk Density (g/cm^3)\n0-25cm\nDominant Component',
  margin = FALSE, 
  scales = list(draw = FALSE), 
  col.regions = viridis,
  maxpixels = 1e5
)

levelplot(
  mu.stack[['awc_r']], 
  main = 'AWC (cm/cm) 0-25cm\nDominant Component',
  margin = FALSE, 
  scales = list(draw = FALSE), 
  col.regions = viridis,
  maxpixels = 1e5
)

levelplot(
  mu.stack[['ph1to1h2o_r']], 
  main = 'pH 1:1 H2O 0-25cm\nDominant Component',
  margin = FALSE, 
  scales = list(draw = FALSE), 
  col.regions = viridis,
  maxpixels = 1e5
)
```


## Sand, Silt, and Clay at a Soil Survey Area Boundary
A not so great SSA join (1979 vs 2004).

Setup BBOX and query mukey WCS.
```{r}
# extract a BBOX like this from SoilWeb by pressing "b"
bb <- '-91.6853 36.4617,-91.6853 36.5281,-91.5475 36.5281,-91.5475 36.4617,-91.6853 36.4617'
wkt <- sprintf('POLYGON((%s))', bb)

# init sf object from WKT
x <- st_as_sfc(wkt)

# assign GCS WGS84 coordinate reference system
st_crs(x) <- 4326

# get gSSURGO grid here
mu <- mukey.wcs(aoi = x, db = 'gssurgo')

# note SSA boundary
plot(mu, legend = FALSE, axes = FALSE, col = viridis(100))
```

Derive aggregate sand, silt, clay (RV) values from the largest component, taking the weighted mean over 25-50cm depth interval.
```{r fig.width = 8, fig.height = 6}
# extract RAT for thematic mapping
rat <- cats(mu)[[1]]

# variables of interest
vars <- c("sandtotal_r","silttotal_r","claytotal_r")

# get thematic data from SDA
# dominant component
# depth-weighted average
# sand, silt, clay (RV)
p <-  get_SDA_property(property = vars,
                       method = "Dominant Component (Numeric)", 
                       mukeys = as.integer(rat$mukey),
                       top_depth = 25,
                       bottom_depth = 50)


# check
head(p)

# merge aggregate soil data into RAT
rat <- merge(rat, p, by.x = 'mukey', by.y = 'mukey', sort = FALSE, all.x = TRUE)
levels(mu) <- rat

# convert mukey grid + RAT -> stack of numerical grids
# retaining only sand, silt, clay via [[vars]]
ssc <- catalyze(mu)[[vars]]

# graphical check
# note implicit simplification via maxpixels
levelplot(
  ssc, 
  main = 'Sand, Silt, Clay (RV) 25-50cm\nDominant Component',
  margin = FALSE, 
  scales = list(draw = FALSE), 
  col.regions = viridis,
  maxpixels = 1e4,
  layout = c(3, 1)
)
```


Convert to soil texture class of <2mm fraction.
```{r fig.width = 8, fig.height = 4}
# create a new grid by copy
# we will place soil texture classes into this grid
texture.class <- ssc[[1]]
names(texture.class) <- 'soil.texture'

# assign soil texture classes for the fine earth fraction
# using sand and clay percentages
values(texture.class) <- ssc_to_texcl(
  sand = values(ssc[['sandtotal_r']]), 
  clay = values(ssc[['claytotal_r']]), 
  droplevels = FALSE
)

plot(texture.class, col = viridis(50), axes = FALSE, main = 'Soil Texture Class <2mm Fraction\nDominant Component, 25-50cm (RV)')
box()
```

## Component level properties

Steel corrosion potential.
```{r, fig.width = 6, fig.height = 4}
# make a bounding box and assign a CRS (4326: GCS, WGS84)
a <- st_bbox(
  c(xmin = -114.16, xmax = -114.08, ymin = 47.65, ymax = 47.68), 
  crs = st_crs(4326)
)

# convert bbox to sf geometry
a <- st_as_sfc(a)

# fetch gSSURGO map unit keys at native resolution (~30m)
x <- mukey.wcs(aoi = a, db = 'gssurgo')

# extract RAT for thematic mapping
rat <- cats(x)[[1]]

tab <- get_SDA_property(property = 'Corrosion of Steel', 
                        method = 'DOMINANT CONDITION',
                        mukeys = as.integer(rat$mukey)
)


# convert to factor and set levels in logical order
tab$corsteel <- factor(tab$corsteel, levels = c('Low', 'Moderate', 'High'))

# merge aggregate soil data into RAT
rat <- merge(rat, tab, by.x = 'mukey', by.y = 'mukey', sort = FALSE, all.x = TRUE)
levels(x) <- rat

# set active category to corsteel
activeCat(x) <- 'corsteel'

# plot
plot(x, axes = FALSE, col = viridis(3))
box()
```


Simplified component parent material group.
```{r, fig.width = 6, fig.height = 4}
# make a bounding box and assign a CRS (4326: GCS, WGS84)
a <- st_bbox(
  c(xmin = -114.16, xmax = -114.08, ymin = 47.65, ymax = 47.68), 
  crs = st_crs(4326)
)

# convert bbox to sf geometry
a <- st_as_sfc(a)

# fetch gSSURGO map unit keys at native resolution (~30m)
x <- mukey.wcs(aoi = a, db = 'gssurgo')

# extract RAT for thematic mapping
rat <- cats(x)[[1]]

# simplified parent material group name
tab <- get_SDA_pmgroupname(mukeys = as.integer(rat$mukey))

# merge aggregate soil data into RAT
rat <- merge(rat, tab, by.x = 'mukey', by.y = 'mukey', sort = FALSE, all.x = TRUE)
levels(x) <- rat

# set active category to corsteel
activeCat(x) <- 'pmgroupname'
plot(x, axes = FALSE, col = viridis(10))
box()
```


Hydric Rating.
```{r, fig.width = 6, fig.height = 5.5}
# make a bounding box and assign a CRS (4326: GCS, WGS84)
a <- st_bbox(
  c(xmin = -114.16, xmax = -114.08, ymin = 47.65, ymax = 47.68), 
  crs = st_crs(4326)
)

# convert bbox to sf geometry
a <- st_as_sfc(a)

# fetch gSSURGO map unit keys at native resolution (~30m)
x <- mukey.wcs(aoi = a, db = 'gssurgo')

# extract RAT for thematic mapping
rat <- cats(x)[[1]]

# simplified parent material group name
tab <- get_SDA_hydric(mukeys = as.integer(rat$mukey))

# merge aggregate soil data into RAT
rat <- merge(rat, tab, by.x = 'mukey', by.y = 'mukey', sort = FALSE, all.x = TRUE)
levels(x) <- rat

# set active category to corsteel
activeCat(x) <- 'HYDRIC_RATING'
plot(x, axes = FALSE, col = viridis(10))
```


## Additional interfaces to SDA with aggregation

```{r eval=FALSE}
# tab <- get_SDA_muaggatt(mukeys = ll$ID)

# tab <- get_SDA_interpretation(mukeys = ll$ID, rulename = '')
```


## TODO SOC Stock Estimates + Expected Variation



# Other Relevant WCS

## ISSR-800

Available data sets. (TODO: there are more more to add, including NCCPI)
```{r, eval = FALSE}
WCS_details(wcs = 'ISSR800')
```

```{r, echo = FALSE}
kableExtra::kable_styling(knitr::kable(WCS_details(wcs = 'ISSR800'), row.names = FALSE), full_width = FALSE)
```

Examples for a broad slice of California
```{r}
# make a bounding box and assign a CRS (4326: GCS, WGS84)
a.CA <- st_bbox(
  c(xmin = -121, xmax = -120, ymin = 37, ymax = 38), 
  crs = st_crs(4326)
)

# convert bbox to sf geometry
a.CA <- st_as_sfc(a.CA)

# floating point grids
paws_total <- ISSR800.wcs(aoi = a.CA, var = 'paws')
paws_025cm <- ISSR800.wcs(aoi = a.CA, var = 'paws_025cm')
paws_050cm <- ISSR800.wcs(aoi = a.CA, var = 'paws_050cm')

pH_05cm <- ISSR800.wcs(aoi = a.CA, var = 'ph_05cm')
pH_3060cm <- ISSR800.wcs(aoi = a.CA, var = 'ph_3060cm')

clay_05cm <- ISSR800.wcs(aoi = a.CA, var = 'clay_05cm')
clay_3060cm <- ISSR800.wcs(aoi = a.CA, var = 'clay_3060cm')

silt_3060cm <- ISSR800.wcs(aoi = a.CA, var = 'silt_3060cm')
sand_3060cm <- ISSR800.wcs(aoi = a.CA, var = 'sand_3060cm')

## NA encoding issues
# 0 is a real value
sar <- ISSR800.wcs(aoi = a.CA, var = 'sar')

# 16bit integer grids
wei <- ISSR800.wcs(aoi = a.CA, var = 'wei')

# 8bit unsigned (BYTE) grids with RAT
drainage_class <- ISSR800.wcs(aoi = a.CA, var = 'drainage_class')
weg <- ISSR800.wcs(aoi = a.CA, var = 'weg')
str <- ISSR800.wcs(aoi = a.CA, var = 'str')
```

Check to see if aggregated sand + silt + clay is roughly = 100%. Almost. This isn't surprising since each component of soil texture is aggregated separately.
```{r, fig.width = 6, fig.height = 6.5}
z <- sand_3060cm + silt_3060cm + clay_3060cm
plot(z - 100)
```

Soil texture class 25-50cm.
```{r, fig.width = 6, fig.height = 6.5}
texture_2550cm <- ISSR800.wcs(aoi = a.CA, var = 'texture_2550cm')

# use colors from RAT
cols <- cats(texture_2550cm)[[1]]$hex

# plot using SoilWeb soil texture class colors
plot(texture_2550cm, col = cols, axes = FALSE)
box()
```


Other soil properties of interest.
```{r, fig.width = 6, fig.height = 6.5}
plot(sar, axes = FALSE,  col = viridis(50), main = 'SAR')
box()

plot(wei, axes = FALSE,  col = viridis(50), main = 'Wind Erodibility Index')
box()

plot(drainage_class, axes = FALSE,  col = viridis(50), main = 'Drainage Class')
box()

plot(weg, axes = FALSE,  col = viridis(50), main = 'Wind Erodibility Group')
box()

plot(str, axes = FALSE,  col = viridis(50), main = 'Soil Temperature Regime')
box()
```

Stacking soil property grids for multi-panel figures that share a common color ramp / legend.
```{r fig.width = 10, fig.height=6}
# combine into multi-layer SpatRaster
# like raster::stack()
s <- c(pH_05cm, pH_3060cm)

# plot multiple layers via SpatRaster plot method
# common color map / legend
plot(s, col = viridis(50), axes = FALSE, mar = c(2, 2, 2, 4), range = c(4.5, 9))
```


I like this approach.
```{r fig.width = 10, fig.height=6}
# or, plot using rasterVis::levelplot

# note laundering of layer names, required when original layer names contain spaces, punctuation, etc.
names(s) <- make.names(names(s))

# manually specify panel names
# in original ordering specified to c()
# formatting ideas: https://stackoverflow.com/questions/34652580/change-raster-panel-titles-using-levelplot
levelplot(s, 
          scales = list(draw = FALSE), 
          col.regions = viridis, 
          names.attr = c('pH 1:1 H20, 0-5cm', 'pH 1:1 H20, 30-60cm')
)
```


Another example, with 3 panels.
```{r fig.width = 10, fig.height=6}
s <- c(sand_3060cm, silt_3060cm, clay_3060cm)
names(s) <- make.names(names(s))
levelplot(s, 
          scales = list(draw = FALSE), 
          col.regions = viridis,
          names.attr = c('Percent Sand, 30-60cm', 'Percent Silt, 30-60cm', 'Percent Clay, 30-60cm')
)
```



### Soil / Land Classification
```{r fig.width=8, fig.height=8}
# make a bounding box and assign a CRS (4326: GCS, WGS84)
a <- st_bbox(
  c(xmin = -114.30, xmax = -114.00, ymin = 47.5, ymax = 47.70), 
  crs = st_crs(4326)
)

# convert bbox to sf geometry
a <- st_as_sfc(a)

# these are grids + RAT
greatgroup <- ISSR800.wcs(aoi = a, var = 'greatgroup')
lcc_irrigated <- ISSR800.wcs(aoi = a, var = 'lcc_irrigated')

# create a large color palette for greatgroup classes
cols <- brewer.pal(12, 'Set3')
cols <- colorspace::darken(cols, amount = 0.25)
cr <- colorRampPalette(cols, space = 'Lab', interpolate = 'spline')

# rasterVis method still fails
# levelplot(greatgroup, scales = list(draw = FALSE), col.regions = cr)
# terra plot method
plot(greatgroup, axes = FALSE, col = cr(50))

# colors for irrigated LCC
cols <- rev(brewer.pal(8, 'Spectral'))
cols <- colorspace::darken(cols, amount = 0.1)
cr <- colorRampPalette(cols, space = 'Lab', interpolate = 'spline')

# rasterVis method still fails
# levelplot(lcc_irrigated, scales = list(draw = FALSE), col.regions = cr)
plot(lcc_irrigated, axes = FALSE, col = cr(50))
```


### Soil Color
```{r fig.width = 6, fig.height = 6.5}
# make a bounding box and assign a CRS (4326: GCS, WGS84)
a <- st_bbox(
  c(xmin = -121, xmax = -120, ymin = 37, ymax = 38), 
  crs = st_crs(4326)
)

# convert bbox to sf geometry
a <- st_as_sfc(a)

# moist soil color at 25cm, low-res version
s <- soilColor.wcs(aoi = a, var = 'sc025cm', res = 270)

# note access to colors via RAT
rat <- cats(s)[[1]]
plot(s, col = rat$col, legend = FALSE, axes = FALSE, main = attr(s, 'layer name'))
box()
```

Soil color distribution.
```{r fig.width = 8, fig.height = 4}
# specific to this AOI
barplot(s, col = rat$col, las = 3, cex.names = 0.66)

# these colors, but frequencies tabulated across all of CONUS
# barplot(rat$freq, col = rat$col, las = 3, names.arg = rat$munsell, cex.names = 0.6)
```


Detailed soil color mapping.
```{r fig.width = 9, fig.height = 6}
# from SoilWeb
bb <- '-90.1576 41.8579,-90.1576 41.9193,-90.0100 41.9193,-90.0100 41.8579,-90.1576 41.8579'

# convert text -> WKT -> sf
wkt <- sprintf('POLYGON((%s))', bb)
a <- st_as_sfc(wkt, crs = 4326)

# moist soil color at 25cm, low-res version
s <- soilColor.wcs(aoi = a, var = 'sc025cm_hr', res = 30)

# note access to colors via RAT
rat <- cats(s)[[1]]
plot(s, col = rat$col, legend = FALSE, axes = FALSE, main = attr(s, 'layer name'))
box()
```


### Classification
```{r}
# part of CA

# make a bounding box and assign a CRS (4326: GCS, WGS84)
a <- st_bbox(
  c(xmin = -121, xmax = -120, ymin = 37, ymax = 38), 
  crs = st_crs(4326)
)

# convert bbox to sf geometry
a <- st_as_sfc(a)

# EC 0-25cm
ec <- ISSR800.wcs(aoi = a, var = 'ec_025cm')

# pH 0-25cm
pH <- ISSR800.wcs(aoi = a, var = 'ph_025cm')

# no ESP grid, use a constant value of 1%
esp <- ec
values(esp) <- 0
names(esp) <- 'ESP'

# copy EC grid, replace values with classification in next step
ss <- ec
names(ss) <- 'FAO Salt Severity'

# resulting raster contains factor levels (integers) and labels
values(ss) <- allocate(EC = values(ec), pH = values(pH), ESP = values(esp), to = 'FAO Salt Severity', droplevels = TRUE)

# note factor details stored in raster attribute table (RAT)
print(ss)

# plot
plot(ss, axes = FALSE, main = 'FAO Salt Severity', col = viridis(10))
box()
```


FAO Salt Severity Classification
```{r eval = FALSE}
## TODO: NODATA not correctly handled, possibly need to mask

# combine individual SpatRaster objects into a multi-band SpatRaster
s <- c(pH, ec, esp)

# classification wrapper
.fun <- function(x) {
  
  .pH <- x[1]
  .ec <- x[2]
  .esp <- x[3]
  
  .res <- allocate(EC = .ec, pH = .pH, ESP = .esp, to = 'FAO Salt Severity', droplevels = FALSE)
  return(.res)
}

# apply classification pixel-wise to stack
ss <- app(s, fun = .fun)

## TODO: can this be done within app()?
# manually specified levels
ll <- c("nonsaline", "slightly saline", "moderately saline", 
"strongly saline", "very strongly saline", "extremely saline", 
"saline-sodic", "slightly sodic", "moderately sodic", "strongly sodic", 
"very strongly sodic")

# convert to grid + RAT
levels(ss) <- data.frame(ID = 1:length(ll), class = ll)

plot(ss, axes = FALSE, main = 'FAO Salt Severity', col = viridis(10))
box()
```

## PRISM Stack

## Others

# Artistic License
https://github.com/riatelab/tanaka

```{r fig.width = 8, fig.height = 8}
library(tanaka)

tanaka(
  pH_3060cm, 
  breaks = c(6, 7, 8), 
  legend.pos = "topright", 
  legend.title = "pH 1:1 H2O (30-60cm)", 
  col = viridis(4)
)

```


----------------------------
This document is based on `soilDB` version `r utils::packageDescription("soilDB", field="Version")` and `terra` version `r utils::packageDescription("terra", field="Version")`.


