---
title: "gNATSGO / gSSURGO Web Coverage Service"
author: "D.E. Beaudette"
date: "`r Sys.Date()`"
output:
  html_document:
    mathjax: null
    jquery: null
    smart: no
    number_sections: no
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: false
---
  
  
```{r setup, echo=FALSE, results='hide', warning=FALSE}
# setup
library(knitr, quietly=TRUE)
library(kableExtra, quietly=TRUE)
opts_chunk$set(message=FALSE, warning=FALSE, background='#F7F7F7', fig.align='center', fig.retina=2, dev='png', tidy=FALSE, verbose=FALSE)
options(width=100, stringsAsFactors=FALSE)
```


# Introduction

Borrowing content from the official [gSSURGO](https://www.nrcs.usda.gov/wps/portal/nrcs/detail/soils/home/?cid=nrcs142p2_053628#release) and [gNATSGO](https://www.nrcs.usda.gov/wps/portal/nrcs/detail/soils/survey/geo/?cid=nrcseprd1464625) pages.

## gSSURGO
The gSSURGO Database is derived from the official Soil Survey Geographic (SSURGO) Database. SSURGO generally has the most detailed level of soil geographic data developed by the National Cooperative Soil Survey (NCSS) in accordance with NCSS mapping standards. The tabular data represent the soil attributes and are derived from properties and characteristics stored in the National Soil Information System (NASIS). The gSSURGO data were prepared by merging the traditional vector-based SSURGO digital map data and tabular data into statewide extents, adding a statewide gridded map layer derived from the vector layer, and adding a new value-added look up table (Valu1) containing “ready to map” attributes. The gridded map layer is a file geodatabase raster in an ArcGIS file geodatabase. The raster and vector map data have a statewide extent. The raster map data have a 10-meter cell size that approximates the vector polygons in an Albers Equal Area projection. Each cell (and polygon) is linked to a map unit identifier called the map unit key. A unique map unit key is used to link the raster cells and polygons to attribute tables. Due to file size, the raster layer for the conterminous United States is only available in a 30-meter resolution.


## gNATSGO
The gNATSGO databases contain a raster of the soil map units and 70 related tables of soil properties and interpretations. They are designed to work with the SPSD gSSURGO ArcTools. Users can create full coverage thematic maps and grids of soil properties and interpretations for large geographic areas, such as the extent of a State or the conterminous United States. Please note that the State-wide geodatabases contain a 10 meter raster and the CONUS database contains a 30 meter raster.

The gNATSGO database is composed primarily of SSURGO data, but STATSGO2 data was used to fill in the gaps. The RSSs are newer product with relatively limited spatial extent.  These RSSs were merged into the gNATSGO after combining the SSURGO and STATSGO2 data. The extent of RSS is expected to increase in the coming years.


## An Experimental Web Coverage Service


```
http://soilmap2-1.lawr.ucdavis.edu/cgi-bin/mapserv?map=/soilmap2/website/wcs/mukey.map&SERVICE=WCS&VERSION=2.0.1&REQUEST=GetCoverage&COVERAGEID=gnatsgo&FORMAT=image/tiff&GEOTIFF:COMPRESSION=Deflate&SUBSETTINGCRS=EPSG:6350&FORMAT=GEOTIFF_FLOAT&SUBSET=x(-1365484,-1358915)&SUBSET=y(2869259,2873658)&RESOLUTION=x(30)&RESOLUTION=y(30)
```


## Setup
Need the latest version of `soilDB`, along with all packages specfied below (CRAN versions fine).
```{r eval = FALSE}
remotes::install_github("ncss-tech/soilDB", dependencies = FALSE, upgrade = FALSE, build = FALSE)
```


## Grid Selection
There are two options.
```{r eval = FALSE}
# select gSSURGO grid
x <- mukey.wcs(aoi = aoi, db = 'gssurgo', ...)

# select gNATSGO grid
x <- mukey.wcs(aoi = aoi, db = 'gnatsgo', ...)
```


## Caveats / Limitations

Requests are limited to images sizes of 5000x5000 pixels, this is approximately 1x1 degrees at 30m.

### To-Fix
   * Ensure that data are not resampled unless a resolution other than 30m is requested.
   * Currently the data come back *close* to 30m, usually 29.998-30.111. I'm not sure if this is something that has to be solved in the creation of the BBOX or if it is a Mapserver configuration issue. Still tinkering.
   * Consider converting source mukey grids to COG-GeoTiff, this will require a newer version of GDAL.
   * Simplify / unify soil data aggregation within SDA: this is much more generic than this WCS idea, but is a perfect use case.



# A Simple Demonstration

```{r}
# latest from GitHub
library(soilDB)

# wrangling polygons and CRS transformations
library(sp)
library(rgdal)

# geometric calculations
library(rgeos)

# raster data visualization
library(raster)
library(rasterVis)
library(viridis)
```


```{r}

# aoi: bounding-box corners (xmin, ymin, xmax, ymax)
# coordinate reference system of AOI (GCS WGS84 )
a <- list(
  aoi = c(-114.16, 47.65, -114.08, 47.68),
  crs = '+init=EPSG:4326'
)

# too big for SDA geometry (>32Mb WKT serialization)
# a <- c(-114.5, 47, -114, 47.5)

# fetch gSSURGO map unit keys at native resolution (~30m)
x <- mukey.wcs(aoi = a, db = 'gnatsgo')


# OK
levelplot(
  x, 
  att = 'ID',
  main = 'gSSURGO map unit keys',
  sub = 'Albers Equal Area Projection',
  margin = FALSE, 
  colorkey = FALSE, 
  col.regions = viridis
  )
```

## BBOX Specification
Bounding box corners (xmin, ymin, xmax, ymax).


## Fetching Polygons from SDA
```{r}
# convert raster extent into vector 
g <- as(extent(x), 'SpatialPolygons')
proj4string(g) <- projection(x)

# get intersecting SSURGO linework as SpatialPolygonsDataFrame from SDA
p <- SDA_spatialQuery(g, what = 'geom', geomIntersection = TRUE)

# transform to AEA CRS
p <- spTransform(p, CRS(projection(x)))

rgeos::gArea(p) * 0.000247105

par(mar = c(1, 0, 2, 0))
plot(p, main = 'SSURGO Polygons (SDA)')
mtext('Albers Equal Area Projection', side = 1, line = -0.5)
```


Overlay on gSSURGO grid.
```{r}
levelplot(
  x, att = 'ID', 
  main = 'gSSURGO Grid (WCS)\nSSURGO Polygons (SDA)',
  margin = FALSE, 
  colorkey = FALSE,
  col.regions = viridis,
  panel = function(...) {
    panel.levelplot(...)
    sp.lines(p, col = 'white')
  }
)

```

## Grid Resolution Specification
Native vs. alternative resolutions (preview not aggregation).

```{r}
# part of CA
a.CA <- list(
  aoi = c(-121, 37, -120, 38),
  crs = '+init=EPSG:4326'
)

# fetch gSSURGO map unit keys at ~~ 800m
# nearest-neighbor resampling = this is a "preview"
x.800 <- mukey.wcs(aoi = a.CA, db = 'gssurgo', res = 800)


# OK
levelplot(
  x.800, 
  att = 'ID',
  main = 'A Preview of gSSURGO Map Unit Keys',
  sub = 'Albers Equal Area Projection (800m)\nnearest-neighbor resampling',
  margin = FALSE, 
  colorkey = FALSE, 
  col.regions = viridis
  )
```

# Tabular Data from SDA 

```{r}
# get unique mukeys from grid
ll <- levels(x)[[1]]

# convert into an SQL "IN" statement
IS <- format_SQL_in_statement(ll$ID)

# query SDA by mukey
sql <- sprintf("SELECT mukey, aws025wta, aws050wta, aws0100wta, aws0150wta FROM muaggatt WHERE mukey IN %s", IS)
tab <- SDA_query(sql)

```


## RAT Management
```{r}

# merge AWS by mukey into raster attribute table (RAT)
rat <- merge(ll, tab, by.x = 'ID', by.y = 'mukey', sort = FALSE, all.x = TRUE)

# re-pack RAT
levels(x) <- rat

# check: note RAT
x

# raster + RAT --> grid of values
# when more than a single attribute, result is a raster stack
# note this is a raster stack with 4 layers
(aws <- deratify(x))

# graphical check
levelplot(
  aws, 
  main = 'gNATSGO WCS + SDA',
  margin = FALSE, 
  scales = list(draw = FALSE), 
  col.regions = viridis
  )
```

## Map Unit Aggregatation Strategies
A "soil data aggregation engine".

   * SOD SQL templates
   * R functions
   * code from ArcGIS toolboxes
   * ???


# Related WCS

## ISSR-800

Available datasets.
```{r, eval = FALSE}
WCS_details(wcs = 'ISSR800')
```

```{r, echo = FALSE}
kableExtra::kable_styling(knitr::kable(WCS_details(wcs = 'ISSR800'), row.names = FALSE), full_width = FALSE)
```

```{r}
# part of CA
a <- list(
  aoi = c(-121, 37, -120, 38),
  crs = '+init=EPSG:4326'
)

# floating point grids
paws_total <- ISSR800.wcs(aoi = a, var = 'paws')
paws_025cm <- ISSR800.wcs(aoi = a, var = 'paws_025cm')
paws_050cm <- ISSR800.wcs(aoi = a, var = 'paws_050cm')

pH_05cm <- ISSR800.wcs(aoi = a, var = 'ph_05cm')
pH_3060cm <- ISSR800.wcs(aoi = a, var = 'ph_3060cm')

clay_05cm <- ISSR800.wcs(aoi = a, var = 'clay_05cm')
clay_3060cm <- ISSR800.wcs(aoi = a, var = 'clay_3060cm')

silt_3060cm <- ISSR800.wcs(aoi = a, var = 'silt_3060cm')
sand_3060cm <- ISSR800.wcs(aoi = a, var = 'sand_3060cm')

## NA encoding issues
# 0 is a real value
sar <- ISSR800.wcs(aoi = a, var = 'sar')

# 16bit integer grids
wei <- ISSR800.wcs(aoi = a, var = 'wei')

# 8bit unsigned (BYTE) grids with RAT
drainage_class <- ISSR800.wcs(aoi = a, var = 'drainage_class')
weg <- ISSR800.wcs(aoi = a, var = 'weg')
str <- ISSR800.wcs(aoi = a, var = 'str')
```

```{r}
# compositional data may lose some fidelity due to marginal aggregation
z <- sand_3060cm + silt_3060cm + clay_3060cm
levelplot(z - 100)
```

```{r}
levelplot(sar, margin = FALSE)

levelplot(wei, margin = FALSE)

levelplot(drainage_class, margin = FALSE)
levelplot(weg, margin = FALSE)
levelplot(str, margin = FALSE)
```

```{r fig.width = 10, fig.height=6}
levelplot(stack(pH_05cm, pH_3060cm), margin = FALSE)

levelplot(stack(clay_05cm, clay_3060cm), margin = FALSE)

levelplot(stack(sand_3060cm, silt_3060cm, clay_3060cm), margin = FALSE)
```



```{r}
a <- list(
  aoi = c(-114.20, 47.65, -114.10, 47.70),
  crs = '+init=EPSG:4326'
)

series_name <- ISSR800.wcs(aoi = a, var = 'series_name')
levelplot(series_name, margin = FALSE)


greatgroup <- ISSR800.wcs(aoi = a, var = 'greatgroup')
levelplot(greatgroup, margin = FALSE)


texture_2550cm <- ISSR800.wcs(aoi = a, var = 'texture_2550cm')
levelplot(texture_2550cm, margin = FALSE)

lcc_irrigated <- ISSR800.wcs(aoi = a, var = 'lcc_irrigated')
levelplot(lcc_irrigated, margin = FALSE)
```

## PRISM Stack

## Others


----------------------------
This document is based on `soilDB` version `r utils::packageDescription("soilDB", field="Version")`.


